---
title: "Can PESTO recover APSIM's own parameters?"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Can PESTO recover APSIM's own parameters?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~3 s under rmarkdown::render(); macOS arm64 (Darwin 26.5),
     R 4.5.2, one core. APSIM is not run at build time: both calibrations
     are frozen results shipped with the package and read from disk here,
     so the time is figure rendering alone. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(PESTO)
PAL <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#000000")
```

## Why

I have APSIM configured for my site and a calibration method I have only
seen work on toy problems. Before I trust it with parameters I cannot
measure, I want two things answered: *if I hide a set of APSIM parameters I
already know and hand the method only noisy yields, does it get them back,
and does it report clearly which of them the data could not pin down? And on
real measured biomass, where there is no truth to check against, does it
agree with an established calibrator?* This vignette answers both against the
APSIM Wheat example, in two parts.

Part 1 is a synthetic-truth recovery experiment, sometimes called a twin
experiment: fix known-true APSIM parameters, simulate per-season wheat
yields, add measurement noise, then try to recover the truth from the noisy
yields. Because the truth is known by construction, correctness is
verifiable without any field data, which is the cleanest way to test a
calibration method before confronting real observations. Part 2 drops the
known truth and calibrates physiological parameters to real observed wheat
biomass, cross-checking PESTO's ensemble posterior against the `apsimx`
package's own point optimiser.

## What

The forward model in both parts is APSIM itself, driven in process through
[`apsim_callback()`](../html/apsim_callback.html) and wrapped in a
[`pesto_forward_model()`](../html/pesto_forward_model.html), so a matrix of
parameter realisations maps to a matrix of simulated outputs.
[`pesto_ies_callback()`](../html/pesto_ies_callback.html) conditions the
prior ensemble on the calibration observations, and
[`as_manifest()`](../html/as_manifest.html) wraps the result as the
hash-verified record a downstream tool consumes. The coupling mechanics are
the subject of *Wiring your own simulator into PESTO*; this vignette is the
calibration counterpart, and reads the results those verbs produced.

```{r load}
rds <- system.file("extdata", "apsim_case_study",
                   "apsim_wheat_calibration_result.rds", package = "PESTO")
if (!nzchar(rds)) {
  rds <- file.path("..", "inst", "extdata", "apsim_case_study",
                   "apsim_wheat_calibration_result.rds")
}
R <- readRDS(rds)
spec <- R$spec
pnm <- R$pnm
```

The results below are frozen real outputs of the calibration, shipped with
the package and read from disk here; APSIM is not run at build time, because
it is an external simulator and is absent from package check farms. They
were produced by `r R$apsim_version` through `apsimx` `r R$apsimx_version`,
PESTO `r R$pesto_version`, seed `r R$seed`, on `r R$generated_on`. The
runnable driver is
`system.file("case_studies/apsim_wheat_calibration.R", package = "PESTO")`.

## Do

### The calibration target

Three parameters are estimated, chosen from a sensitivity screen of the real
model to span the identifiability spectrum: two strong and mechanistically
distinct levers, soil water and nitrogen, and one deliberately weak lever.

```{r target-table}
tab_target <- data.frame(
  Parameter = pnm,
  Role  = vapply(spec, `[[`, "", "role"),
  Unit  = vapply(spec, `[[`, "", "unit"),
  Truth = vapply(spec, `[[`, 0,  "truth"),
  `Prior range` = sprintf("%g to %g",
                          vapply(spec, `[[`, 0, "lo"),
                          vapply(spec, `[[`, 0, "hi")),
  check.names = FALSE, row.names = NULL
)
knitr::kable(
  tab_target,
  caption = "Calibration target: the known truth and the uniform prior."
)
```

### The synthetic observations

The truth run produces one wheat yield per season. Gaussian measurement
noise of `r R$obs_sd` kg/ha is added, and the seasons are split into a
calibration set used in the fit and a held-out validation set used only to
test out-of-sample prediction.

```{r obs-figure, fig.height = 4.2, fig.cap = "Per-season wheat yield across the multi-decade run. The grey line is the noise-free truth; the points are the synthetic observations, split into the seasons used for calibration and the seasons held out for validation."}
tab_obs <- data.table::data.table(
  year = R$season_year,
  truth = R$y_truth,
  obs = R$y_obs_all,
  set = ifelse(seq_along(R$y_truth) %in% R$cal_idx,
               "calibration", "validation")
)
ggplot2::ggplot(tab_obs, ggplot2::aes(year, obs, colour = set)) +
  ggplot2::geom_line(ggplot2::aes(y = truth), colour = "grey60",
                     linewidth = 0.3) +
  ggplot2::geom_point(size = 1.3) +
  ggplot2::scale_colour_manual(
    values = c(calibration = PAL[1], validation = PAL[2]), name = NULL
  ) +
  ggplot2::labs(
    x = "season", y = "wheat yield (kg/ha)",
    title = "Synthetic observations: truth plus measurement noise"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

### The calibration call

This is the whole calibration, shown for reference; the frozen fit is loaded
above.

```{r calib-code, eval = FALSE}
fm <- apsim_callback(
  template  = "Wheat.apsimx",
  param_map = list(
    CN2Bare = ".Simulations.Simulation.Field.Soil.SoilWater.CN2Bare",
    FertN   = ".Simulations.Simulation.Field.Fertilise at sowing.Amount",
    SowingDepth =
      ".Simulations.Simulation.Field.Sow using a variable rule.SowingDepth"
  ),
  output_extractor = function(sim) as.numeric(sim$Yield)
)
# Run APSIM realisations in parallel; hold out validation seasons via obs_sd.
fm <- pesto_forward_model(fm, parallel = "multicore")
fit <- pesto_ies_callback(fm, prior_ensemble = prior, obs = obs_all,
                          obs_sd = sd_vec, noptmax = 6)
```

### Convergence

```{r phi-figure, fig.height = 4.2, fig.cap = "Objective function by iteration for the APSIM calibration. The ribbon spans the ensemble's minimum to maximum misfit and the line its mean: the misfit drops sharply over the first iterations and then plateaus, which is the ensemble reaching the noise floor of the synthetic observations."}
tab_phi_wide <- data.table::dcast(
  R$fit$phi, iteration ~ realisation, value.var = "phi"
)
vec_complete <- !vapply(tab_phi_wide, anyNA, logical(1L))
plot_phi(tab_phi_wide[, which(vec_complete), with = FALSE],
         title = "APSIM calibration: misfit by iteration")
```

### Parameter recovery

```{r recovery-table}
tab_recovery <- data.table::rbindlist(lapply(pnm, function(p) {
  q <- stats::quantile(R$post[, p], c(0.05, 0.5, 0.95), na.rm = TRUE)
  data.table::data.table(
    Parameter = p,
    Truth = spec[[p]]$truth,
    `Posterior mean` = round(mean(R$post[, p], na.rm = TRUE), 1),
    `5%` = round(q[1], 1),
    `95%` = round(q[3], 1),
    `Truth in 90% band` = ifelse(
      spec[[p]]$truth >= q[1] & spec[[p]]$truth <= q[3], "yes", "no"
    )
  )
})) # ends the loop over calibrated parameters
knitr::kable(
  tab_recovery,
  caption = "Recovery of the known truth from noisy per-season yields."
)
```

```{r recovery-figure, fig.height = 4.2, fig.cap = "Prior and posterior densities for each parameter, with the true value marked by a dashed line. All three posteriors contract substantially and all three brackets the truth; the weak lever, sowing depth, contracts as much as the others but settles furthest from its true value."}
tab_dens <- data.table::rbindlist(lapply(pnm, function(p) rbind(
  data.table::data.table(Parameter = p, value = R$prior[, p], dist = "prior"),
  data.table::data.table(Parameter = p, value = R$post[, p],
                         dist = "posterior")
))) # ends the loop over calibrated parameters
tab_truth <- data.table::data.table(
  Parameter = pnm, truth = vapply(spec, `[[`, 0, "truth")
)
ggplot2::ggplot(tab_dens, ggplot2::aes(value, fill = dist)) +
  ggplot2::geom_density(alpha = 0.5, colour = NA) +
  ggplot2::geom_vline(data = tab_truth, ggplot2::aes(xintercept = truth),
                      linetype = "dashed", linewidth = 0.5) +
  ggplot2::facet_wrap(~ Parameter, scales = "free") +
  ggplot2::scale_fill_manual(
    values = c(prior = "grey70", posterior = PAL[1]), name = NULL
  ) +
  ggplot2::labs(
    x = "parameter value (units as in the target table)", y = "density",
    title = "Prior to posterior; the dashed line is the truth"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

### Identifiability: what the data can constrain

A parameter is well identified when conditioning on the data collapses its
spread far below the prior. The ratio of posterior to prior standard
deviation makes that explicit, with the caveat that this frozen result comes
from a release whose smoother under-disperses, so the ratios are comparable
with each other but are all smaller than the current smoother would
produce.

```{r ident-figure, fig.height = 3.6, fig.cap = "Ratio of posterior to prior standard deviation by parameter. A low value means conditioning on the yields contracted the parameter's spread; a value near the dotted line at one would mean the data said almost nothing about it. Read the ratios against each other rather than absolutely: the release that produced this result under-disperses, so every posterior here is narrower than it should be."}
tab_ident <- data.table::data.table(
  Parameter = pnm,
  ratio = vapply(pnm, function(p) {
    stats::sd(R$post[, p], na.rm = TRUE) / stats::sd(R$prior[, p])
  }, 0)
)
ggplot2::ggplot(
  tab_ident, ggplot2::aes(stats::reorder(Parameter, ratio), ratio)
) +
  ggplot2::geom_col(fill = PAL[1], width = 0.6) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dotted") +
  ggplot2::coord_flip() +
  ggplot2::labs(
    x = NULL,
    y = "posterior SD / prior SD (dimensionless; lower is better identified)",
    title = "Identifiability from the observed yields"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

### Out-of-sample validation, and the under-dispersion caveat

Using the posterior ensemble to predict the held-out seasons tests genuine
out-of-sample skill. The finding is that the raw ensemble's predictive band
is too narrow.

```{r validation-figure, fig.height = 4.2, fig.cap = "Held-out seasonal yields against the posterior 90 per cent predictive band, without and with covariance inflation. The panel headings carry the realised coverage of each band; the raw band misses several points that the inflated band contains."}
oe <- R$obs_ensemble
band <- function(mat) {
  data.table::data.table(
    year = R$season_year[R$val_idx],
    truth = R$y_truth[R$val_idx],
    lo = apply(mat[, R$val_idx, drop = FALSE], 2L, stats::quantile, 0.05,
               na.rm = TRUE),
    hi = apply(mat[, R$val_idx, drop = FALSE], 2L, stats::quantile, 0.95,
               na.rm = TRUE)
  )
}
cov_of <- function(b) mean(b$truth >= b$lo & b$truth <= b$hi)
b_base <- band(oe)
cov_base <- cov_of(b_base)
b_base[, panel := sprintf("raw ensemble (%.0f%% covered)", 100 * cov_base)]
lst_panels <- list(b_base)
cov_inf <- NA_real_
if (!is.null(R$inflation)) {
  b_inf <- band(R$inflation$obs_ensemble)
  cov_inf <- cov_of(b_inf)
  b_inf[, panel := sprintf("with RTPS inflation (%.0f%% covered)",
                           100 * cov_inf)]
  lst_panels <- list(b_base, b_inf)
}
tab_band <- data.table::rbindlist(lst_panels)

ggplot2::ggplot(tab_band, ggplot2::aes(year)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), fill = PAL[1],
                       alpha = 0.25) +
  ggplot2::geom_point(ggplot2::aes(y = truth), colour = PAL[2], size = 1.3) +
  ggplot2::facet_wrap(~ panel) +
  ggplot2::labs(
    x = "season", y = "held-out yield (kg/ha)",
    title = "Held-out seasons: 90 per cent predictive band against truth"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

Read that panel as a pre-fix figure. The frozen result loaded above was
produced by PESTO `r R$pesto_version`, whose smoother assimilated the same
unperturbed observation vector into every realisation, which is the dominant
cause of the narrow band shown here and one that inflation was never going
to reach. The development version perturbs the observations per realisation
under an ES-MDA schedule and recovers the analytic posterior covariance on a
problem with a closed-form answer, as *Posterior spread and observation
perturbation* in
[`?pesto_ies_callback`](../html/pesto_ies_callback.html) sets out and the
ensemble-size sweep in *My posterior looks too confident: is the ensemble
collapsing?* demonstrates. Regenerating this case study needs a machine with
APSIM installed, so the panel above still shows the earlier behaviour and
the inflation comparison beside it is calibrated against that behaviour
rather than against the current smoother. What survives the fix unchanged is
the practice: report coverage on your own held-out data, tune inflation to
what you measure, and treat coverage as measured rather than assumed.

### The record the run emits

A calibrated run emits a portable, hash-verified manifest, so a downstream
tool can validate and reuse the posterior without re-running APSIM.

```{r manifest}
man <- as_manifest(R$fit)
man@method
man@schema_version
man@noptmax
```

## Part 2: calibration to real observed data

Part 1 proved recovery against a known truth. Part 2 confronts real field
observations: the `obsWheat` dataset shipped with the `apsimx` package
(Miguez 2025), ten dates of measured wheat above-ground biomass over a
2016-2017 season with `Ames.met` weather. There is no known truth here, so
the test is different: does the calibrated model fit the observations,
predict held-out ones, and agree with an independent calibrator?

The two genuinely uncertain physiological parameters of the bundled
`Wheat-opt-ex.apsimx` model are calibrated: radiation-use efficiency and the
cultivar base phyllochron, which governs phenology. They are the same
parameters, model and data that the `apsimx` package's own `optim_apsimx()`
example optimises, which is what makes the comparison an independent
cross-check of PESTO's ensemble posterior against a point optimum obtained
by a different method.

```{r p2-load}
prd <- system.file("extdata", "apsim_case_study",
                   "apsim_realdata_result.rds", package = "PESTO")
if (!nzchar(prd)) {
  prd <- file.path("..", "inst", "extdata", "apsim_case_study",
                   "apsim_realdata_result.rds")
}
P <- readRDS(prd)
P_spec <- P$spec
P_pnm <- P$pnm
```

```{r p2-obs-figure, fig.height = 4.2, fig.cap = "Observed wheat above-ground biomass at ten dates through the season, split into the dates used for calibration and the dates held out. The measurements are real: the obsWheat dataset shipped with the apsimx package."}
tab_p2obs <- data.table::data.table(
  date = P$obs_dates, biomass = P$y_obs,
  set = ifelse(seq_along(P$y_obs) %in% P$cal_idx, "calibration", "held-out")
)
ggplot2::ggplot(tab_p2obs, ggplot2::aes(date, biomass, colour = set)) +
  ggplot2::geom_line(colour = "grey70", linewidth = 0.3) +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_colour_manual(
    values = c(calibration = PAL[1], `held-out` = PAL[2]), name = NULL
  ) +
  ggplot2::labs(
    x = "date", y = "observed biomass (g/m2)",
    title = "obsWheat: real measured above-ground biomass"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

```{r p2-estimates}
tab_est <- data.table::rbindlist(lapply(P_pnm, function(p) {
  q <- stats::quantile(P$post[, p], c(0.05, 0.5, 0.95), na.rm = TRUE)
  data.table::data.table(
    Parameter = p,
    Role = P_spec[[p]]$role,
    Unit = P_spec[[p]]$unit,
    `PESTO mean` = round(mean(P$post[, p], na.rm = TRUE), 2),
    `5%` = round(q[1], 2),
    `95%` = round(q[3], 2),
    `apsimx optimum` = P_spec[[p]]$apsimx_opt,
    Default = P_spec[[p]]$default,
    `Optimum inside band` = ifelse(
      P_spec[[p]]$apsimx_opt >= q[1] & P_spec[[p]]$apsimx_opt <= q[3],
      "yes", "no"
    )
  )
})) # ends the loop over the two physiological parameters
knitr::kable(
  tab_est,
  caption = paste0(
    "PESTO's posterior against the apsimx optim_apsimx() optimum, an ",
    "independent calibrator, and against the model default."
  )
)
```

```{r p2-fit-figure, fig.height = 4.2, fig.cap = "Posterior predictive 90 per cent band for biomass at each observation date, with the observations overlaid. Filled points are the dates used in the fit; open points are the held-out dates the band was not conditioned on."}
qb <- function(i) {
  stats::quantile(P$obs_ensemble[, i], c(0.05, 0.95), na.rm = TRUE)
}
tab_fitb <- data.table::data.table(
  date = P$obs_dates, obs = P$y_obs,
  lo = vapply(seq_along(P$obs_dates), function(i) qb(i)[1], 0),
  hi = vapply(seq_along(P$obs_dates), function(i) qb(i)[2], 0),
  set = ifelse(seq_along(P$y_obs) %in% P$cal_idx, "calibration", "held-out")
)
ggplot2::ggplot(tab_fitb, ggplot2::aes(date)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), fill = PAL[1],
                       alpha = 0.25) +
  ggplot2::geom_point(ggplot2::aes(y = obs, shape = set), colour = PAL[2],
                      size = 2) +
  ggplot2::scale_shape_manual(
    values = c(calibration = 16L, `held-out` = 1L), name = NULL
  ) +
  ggplot2::labs(
    x = "date", y = "biomass (g/m2)",
    title = "Posterior predictive band against observed biomass"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

```{r read-facts, include = FALSE}
vl <- vapply(P$val_idx, function(i) {
  stats::quantile(P$obs_ensemble[, i], 0.05, na.rm = TRUE)
}, 0)
vh <- vapply(P$val_idx, function(i) {
  stats::quantile(P$obs_ensemble[, i], 0.95, na.rm = TRUE)
}, 0)
p2_cov <- mean(P$y_obs[P$val_idx] >= vl & P$y_obs[P$val_idx] <= vh)
n_inside <- sum(tab_est$`Optimum inside band` == "yes")
rel_err <- 100 * abs(tab_recovery$`Posterior mean` - tab_recovery$Truth) /
  tab_recovery$Truth
names(rel_err) <- tab_recovery$Parameter
err_strong <- max(rel_err[c("CN2Bare", "FertN")])
err_weak <- rel_err[["SowingDepth"]]
q05_weak <- tab_recovery$`5%`[tab_recovery$Parameter == "SowingDepth"]
truth_weak <- tab_recovery$Truth[tab_recovery$Parameter == "SowingDepth"]
n_recovered <- sum(tab_recovery$`Truth in 90% band` == "yes")
ident_best <- tab_ident$Parameter[which.min(tab_ident$ratio)]
ident_worst <- tab_ident$Parameter[which.max(tab_ident$ratio)]
```

## Read

Part 1 answers the first half of the question, and the answer is yes with
one qualification. All `r n_recovered` of the `r length(pnm)` parameters
have the known truth inside their 90 per cent credible band. Accuracy
separates them: the two strong levers land within
`r round(err_strong)` per cent of their true values, while the weak lever,
sowing depth, settles `r round(err_weak)` per cent above its truth and its
band's lower edge sits at `r round(q05_weak, 1)` mm against a truth of
`r truth_weak` mm, so it is covered but only just.

The identifiability figure does not order the three the way the design
intended, and that is worth stating rather than smoothing over. Conditioning
contracts every parameter's spread, to between
`r round(100 * min(tab_ident$ratio))` and
`r round(100 * max(tab_ident$ratio))` per cent of the prior spread, and by
that measure the parameter chosen as the weak lever contracts most
(`r ident_best`, at `r round(100 * min(tab_ident$ratio))` per cent) and the
soil-runoff parameter least (`r ident_worst`, at
`r round(100 * max(tab_ident$ratio))` per cent). The ratio measures
contraction relative to each parameter's own prior width, which is not the
same as the information the yields carry about it, and this frozen result
comes from the release whose posterior spread is known to be too narrow
across the board. What the pair of diagnostics supports jointly is the
narrower claim: every parameter's interval covers its truth, and the
parameter the sensitivity screen marked as the weak lever is the one
recovered least accurately.

The held-out panel is where the raw ensemble falls short. Its 90 per cent
predictive band covers `r round(100 * cov_base)` per cent of the held-out
seasons against a nominal 90, and with RTPS inflation applied that rises to
`r round(100 * cov_inf)` per cent. As stated above the panel, the dominant
cause of the shortfall was the unperturbed-observation defect in PESTO
`r R$pesto_version`, since fixed, and the panel is retained as a pre-fix
figure rather than re-run, because regenerating it needs a machine with
APSIM installed. The practice it teaches is unchanged by the fix: measure
coverage on held-out data and treat it as measured.

Part 2 answers the second half. On real `obsWheat` biomass, with no truth to
compare against, the `apsimx` point optimum lies inside PESTO's 90 per cent
credible band for `r n_inside` of the `r length(P_pnm)` parameters, and both
calibrators move in the same direction away from the model default: a higher
radiation-use efficiency, PESTO's posterior mean
`r round(mean(P$post[, "RUE"], na.rm = TRUE), 2)` against a default of
`r P_spec$RUE$default`, and a shorter base phyllochron,
`r round(mean(P$post[, "Phyllochron"], na.rm = TRUE), 1)` against a default
of `r P_spec$Phyllochron$default`. Two calibration engines built on
different principles, an ensemble smoother and a derivative-free point
optimiser, agree, and PESTO additionally reports the uncertainty a point
optimum does not carry. The held-out dates, excluded from the fit, fall
within the posterior predictive band at a rate of
`r sprintf("%.0f%%", 100 * p2_cov)` over `r length(P$val_idx)` points; with
so few held-out points that figure is indicative rather than a precise
coverage estimate.

## Limits

Part 1's truth is APSIM's own output, so it establishes that the smoother
inverts APSIM correctly and nothing about whether APSIM describes a real
paddock; a structural error in the simulator would be invisible to a twin
experiment by construction. Its noise model is Gaussian, independent across
seasons and known, whereas real yield error is neither independent nor
known. Part 2 has real observations but no truth, so its cross-check
establishes agreement between two calibrators rather than correctness of
either, and two methods can agree on the same wrong answer if they share the
same forward model, which here they do. The held-out set in Part 2 carries
`r length(P$val_idx)` points, far too few to estimate coverage precisely,
and the Part 1 coverage figures come from a superseded release of the
smoother and are retained as a record of the earlier behaviour rather than
as a current measurement. Both parts calibrate two or three parameters of a
model that has hundreds, chosen by a sensitivity screen, so nothing here
speaks to the behaviour of the method in a highly parameterised APSIM
calibration. Neither part is re-run at build time, so the numbers above are
read from shipped artefacts rather than recomputed; the runnable drivers are
named in the *Reproduce* section for anyone with APSIM installed.

## What to read next

*Wiring your own simulator into PESTO* is the coupling side of this study:
the callback driver, the typed forward-model contract, the parallel
evaluation path, and the guard on the observation standard deviation that
decides whether a posterior is credible at all. *My posterior looks too
confident: is the ensemble collapsing?* takes up the under-dispersion seen
in the held-out panel and separates the two causes. *Handing an ensemble to
somebody else* documents the manifest emitted at the end of Part 1.
*Is PESTO the same algorithm as the tools it descends from?* places the smoother beside the
tools it descends from.

## References

- Holzworth, D., et al. (2014). APSIM: evolution towards a new generation of
  agricultural systems simulation. *Environmental Modelling & Software*, 62,
  327-350.
- Miguez, F. (2025). *apsimx: Inspect, Read, Edit and Run APSIM Next
  Generation and APSIM Classic*. R package version 2.8.235.
  <https://doi.org/10.32614/CRAN.package.apsimx> (source of the `obsWheat`
  data and the `Wheat-opt-ex.apsimx` model used in Part 2).

## Reproduce

Both calibrations are reproducible on a machine with APSIM installed. The
drivers hardcode no machine paths: point them at an APSIM installation
through the `PESTO_APSIM_EXE` and `PESTO_APSIM_EXAMPLES` environment
variables, adding `PESTO_DOTNET_ROOT` if the build needs an explicit
runtime, then run them.

```{r repro, eval = FALSE}
# Part 1 (synthetic-truth recovery):
source(system.file("case_studies/apsim_wheat_calibration.R",
                   package = "PESTO"))
# Part 2 (real obsWheat data):
source(system.file("case_studies/apsim_wheat_realdata.R", package = "PESTO"))
```

Provenance for the frozen results read above: `r R$apsim_version`, `apsimx`
`r R$apsimx_version`, PESTO `r R$pesto_version`, seed `r R$seed`, generated
`r R$generated_on`. Part 2 uses the `obsWheat` data and the
`Wheat-opt-ex.apsimx` model shipped with `apsimx`. Package versions for this
rendering follow.

```{r session-info}
sessionInfo()
```
