---
title: "Calibrating APSIM Wheat with PESTO: a recovery case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating APSIM Wheat with PESTO: a recovery case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.2, dpi = 96,
  message = FALSE, warning = FALSE
)
library(PESTO)
library(data.table)
library(ggplot2)
PAL <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#000000")
```

## Why this study

PESTO's flagship partner is **APSIM**, the agricultural-systems simulator
(see the *Calibrating APSIM with PESTO* vignette for the coupling
mechanics). This case study is the calibration counterpart: a complete,
reproducible demonstration that PESTO's iterative ensemble smoother (IES)
recovers known APSIM parameters and quantifies their uncertainty honestly.

It is deliberately a **synthetic-truth recovery experiment** (an OSSE, or
"twin experiment"): we fix known-true APSIM parameters, simulate per-season
wheat yields, add measurement noise, and then try to recover the truth from
the noisy yields. Because the truth is known *by construction*, correctness
is verifiable without any field data -- the honest way to show a calibration
method works before confronting real observations. The forward model is the
**bundled APSIM Wheat example** (a multi-decade run at a single site), driven
in-process through `apsim_callback()`.

```{r load}
rds <- system.file("extdata", "apsim_case_study",
                   "apsim_wheat_calibration_result.rds", package = "PESTO")
if (!nzchar(rds)) {
  rds <- file.path("..", "inst", "extdata", "apsim_case_study",
                   "apsim_wheat_calibration_result.rds")
}
R <- readRDS(rds)
spec <- R$spec; pnm <- R$pnm
```

The results below are **frozen real outputs** of the calibration, shipped
with the package and loaded here; APSIM is not run at build time (it is an
external simulator, absent on CRAN check farms). They were produced by
`r R$apsim_version` via `apsimx` `r R$apsimx_version`, PESTO
`r R$pesto_version`, seed `r R$seed`. The runnable driver is
`system.file("case_studies/apsim_wheat_calibration.R", package = "PESTO")`.

## The calibration target

Three parameters are estimated, chosen from a sensitivity screen of the real
model to span the identifiability spectrum: two strong, mechanistically
distinct levers (soil water and nitrogen) and one deliberately weak lever.

```{r target-table}
tt <- data.frame(
  Parameter = pnm,
  Role  = vapply(spec, `[[`, "", "role"),
  Unit  = vapply(spec, `[[`, "", "unit"),
  Truth = vapply(spec, `[[`, 0,  "truth"),
  `Prior range` = sprintf("%g -- %g",
                          vapply(spec, `[[`, 0, "lo"),
                          vapply(spec, `[[`, 0, "hi")),
  check.names = FALSE, row.names = NULL
)
knitr::kable(tt, caption = "Calibration target: known truth and uniform prior.")
```

## The synthetic observations

The truth run produces one wheat yield per season; we add Gaussian
measurement noise (SD `r R$obs_sd` kg/ha) and split the seasons into a
calibration set (used in the fit) and a held-out validation set (used only
to test out-of-sample prediction).

```{r obs-plot, fig.alt = "Per-season wheat yield across the multi-decade run: the noise-free truth and the noisy synthetic observations, with calibration and held-out validation seasons distinguished."}
od <- data.table(
  year  = R$season_year,
  truth = R$y_truth,
  obs   = R$y_obs_all,
  set   = ifelse(seq_along(R$y_truth) %in% R$cal_idx, "calibration", "validation")
)
ggplot(od, aes(year, obs, colour = set)) +
  geom_line(aes(y = truth), colour = "grey60", linewidth = 0.3) +
  geom_point(size = 1.3) +
  scale_colour_manual(values = c(calibration = PAL[1], validation = PAL[2])) +
  labs(x = NULL, y = "Wheat yield (kg/ha)", colour = NULL,
       title = "Synthetic observations: truth (grey) + measurement noise") +
  theme_minimal(base_size = 11)
```

## The calibration

The forward model is APSIM itself, wrapped so a parameter matrix maps to a
matrix of simulated yields; the IES then conditions the ensemble on the
calibration-season observations. This is the entire calibration call (shown
for reference; the frozen fit is loaded above):

```{r calib-code, eval = FALSE}
fm <- apsim_callback(
  template  = "Wheat.apsimx",
  param_map = list(
    CN2Bare     = ".Simulations.Simulation.Field.Soil.SoilWater.CN2Bare",
    FertN       = ".Simulations.Simulation.Field.Fertilise at sowing.Amount",
    SowingDepth = ".Simulations.Simulation.Field.Sow using a variable rule.SowingDepth"
  ),
  output_extractor = function(sim) as.numeric(sim$Yield)
)
# Run APSIM realisations in parallel; hold out validation seasons via obs_sd.
fm  <- pesto_forward_model(fm, parallel = "multicore")
fit <- pesto_ies_callback(fm, prior_ensemble = prior, obs = obs_all,
                          obs_sd = sd_vec, noptmax = 6)
```

### Convergence

```{r phi, fig.height = 3.6, fig.alt = "Objective-function (phi) trajectory over IES iterations: the ensemble's misfit drops sharply in the first iterations and then plateaus."}
plot_phi(R$fit) +
  ggtitle("IES convergence: misfit (phi) by iteration")
```

## Parameter recovery

```{r recovery-table}
rec <- rbindlist(lapply(pnm, function(p) {
  q <- stats::quantile(R$post[, p], c(.05, .5, .95), na.rm = TRUE)
  data.table(Parameter = p, Truth = spec[[p]]$truth,
             `Posterior mean` = round(mean(R$post[, p], na.rm = TRUE), 1),
             `5%` = round(q[1], 1), `95%` = round(q[3], 1),
             `Truth in 90% CI` = ifelse(spec[[p]]$truth >= q[1] &
                                        spec[[p]]$truth <= q[3], "yes", "no"))
}))
knitr::kable(rec, caption = "Recovery of the known truth from noisy yields.")
```

```{r recovery-plot, fig.height = 3.4, fig.alt = "Prior versus posterior densities for each parameter with the true value marked: the two strong parameters collapse tightly onto the truth, while the weak parameter's posterior stays broad."}
dens <- rbindlist(lapply(pnm, function(p) rbind(
  data.table(Parameter = p, value = R$prior[, p],     dist = "prior"),
  data.table(Parameter = p, value = R$post[, p],      dist = "posterior")
)))
truth_dt <- data.table(Parameter = pnm,
                       truth = vapply(spec, `[[`, 0, "truth"))
ggplot(dens, aes(value, fill = dist)) +
  geom_density(alpha = 0.5, colour = NA) +
  geom_vline(data = truth_dt, aes(xintercept = truth),
             linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ Parameter, scales = "free") +
  scale_fill_manual(values = c(prior = "grey70", posterior = PAL[1])) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Prior to posterior; dashed line = truth") +
  theme_minimal(base_size = 10)
```

### Identifiability -- honest about what the data can constrain

A parameter is well-identified when conditioning on the data collapses its
spread far below the prior. The ratio of posterior to prior standard
deviation makes this explicit:

```{r ident, fig.height = 3.2, fig.alt = "Posterior-to-prior standard-deviation ratio by parameter: low for the two strong parameters (well constrained), near one for the weak parameter (barely constrained)."}
idr <- data.table(
  Parameter = pnm,
  ratio = vapply(pnm, function(p)
    stats::sd(R$post[, p], na.rm = TRUE) / stats::sd(R$prior[, p]), 0)
)
ggplot(idr, aes(stats::reorder(Parameter, ratio), ratio)) +
  geom_col(fill = PAL[1], width = 0.6) +
  geom_hline(yintercept = 1, linetype = "dotted") +
  coord_flip() +
  labs(x = NULL, y = "posterior SD / prior SD (lower = better identified)",
       title = "Identifiability from the observed yields") +
  theme_minimal(base_size = 11)
```

The yields strongly constrain soil runoff (`CN2Bare`) and fertiliser
nitrogen (`FertN`); sowing depth is barely identifiable from yield alone, and
PESTO's diagnostics say so rather than returning a falsely confident estimate.
That is the value of ensemble UQ: it reports what the data can and cannot
teach you.

## Out-of-sample validation -- and the under-dispersion caveat

Using the posterior ensemble to predict the **held-out** seasons tests
genuine out-of-sample skill. The honest finding is that the *raw* ensemble's
predictive band is too narrow:

```{r validation, fig.height = 3.8, fig.alt = "Held-out seasonal yields with the posterior 90 percent predictive band, without and with covariance inflation: the raw band is too narrow and misses several points, the inflated band covers more of them."}
oe <- R$obs_ensemble
band <- function(mat) {
  data.table(year = R$season_year[R$val_idx],
             truth = R$y_truth[R$val_idx],
             lo = apply(mat[, R$val_idx, drop = FALSE], 2L, stats::quantile, .05, na.rm = TRUE),
             hi = apply(mat[, R$val_idx, drop = FALSE], 2L, stats::quantile, .95, na.rm = TRUE))
}
cov_of <- function(b) mean(b$truth >= b$lo & b$truth <= b$hi)
b_base <- band(oe); b_base[, panel := sprintf("raw ensemble (%.0f%% covered)", 100 * cov_of(b_base))]
panels <- list(b_base)
if (!is.null(R$inflation)) {
  b_inf <- band(R$inflation$obs_ensemble)
  b_inf[, panel := sprintf("with RTPS inflation (%.0f%% covered)", 100 * cov_of(b_inf))]
  panels <- list(b_base, b_inf)
}
bd <- rbindlist(panels)
ggplot(bd, aes(year)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL[1], alpha = 0.25) +
  geom_point(aes(y = truth), colour = PAL[2], size = 1.3) +
  facet_wrap(~ panel) +
  labs(x = NULL, y = "Held-out yield (kg/ha)",
       title = "Held-out seasons: 90% predictive band vs truth") +
  theme_minimal(base_size = 10)
```

The raw 90% band covers well under 90% of the held-out truth -- the
finite-ensemble **under-dispersion** an iterative smoother suffers. PESTO
ships the remedy: covariance inflation (and, for spatial problems,
localisation), documented in the *Inflation and localisation* vignette.
Re-running with RTPS inflation widens the posterior and lifts coverage toward
nominal. Inflation strength is problem-dependent, so report coverage and tune
it rather than trusting a default blindly.

## The ensemble manifest

A calibrated run emits a portable, hash-verified `pesto_ensemble_manifest`,
so a downstream tool can validate and reuse the posterior without re-running
APSIM:

```{r manifest}
man <- as_manifest(R$fit)
class(man)
man@method
man@schema_version
```

## Reproducibility

Every number here is reproducible. The driver
(`apsim_wheat_calibration.R`, shipped under `case_studies/`) hardcodes no
machine paths: point it at an APSIM install with the `PESTO_APSIM_EXE` /
`PESTO_APSIM_EXAMPLES` environment variables (and `PESTO_DOTNET_ROOT` if the
build needs an explicit runtime), then:

```{r repro, eval = FALSE}
script <- system.file("case_studies/apsim_wheat_calibration.R", package = "PESTO")
source(script)   # writes apsim_wheat_calibration_result.rds
```

Provenance: `r R$apsim_version`, `apsimx` `r R$apsimx_version`, PESTO
`r R$pesto_version`, seed `r R$seed`, generated `r R$generated_on`.

## Honest reading

- **Strong parameters recovered.** Soil runoff and fertiliser nitrogen are
  recovered with the truth inside the 90% credible interval.
- **Weak parameter flagged, not faked.** Sowing depth is poorly identifiable
  from yield alone; the posterior stays broad and the identifiability
  diagnostic says so.
- **Raw intervals under-cover.** Out-of-sample coverage is below nominal until
  inflation is applied -- a real property of finite-ensemble smoothers,
  surfaced rather than hidden.
- **It is a twin experiment.** Recovery against a known truth demonstrates the
  method; calibrating against real field observations (with an independent
  validation dataset) is the natural next step.
```
