---
title: "In-Process IES via R Callback -- apsimx Adapter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{In-Process IES via R Callback -- apsimx Adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260516)
```

## Why a callback driver?

`pesto_ies()` shells out to the `pestpp-ies` binary, which writes and
reads `.pst` / ensemble / observation files between every realisation.
For an in-process R forward model (an `apsimx` wrapper, a Python
bridge, or a fast synthetic test problem), that file round-trip is
pure overhead. `pesto_ies_callback()` keeps the same C++ ensemble
kernel (`ensemble_solution()`, Chen & Oliver 2013) but drives the
outer loop in R, calling the forward model in-process.

This document shows:

1. The driver on a synthetic linear-Gaussian problem (recovers truth).
2. The shape of the `apsim_callback()` adapter for `apsimx` (run
   block is disabled -- needs APSIM installed).

## A synthetic recovery example

```{r}
library(PESTO)

npar  <- 4L
nobs  <- 8L
nreal <- 120L
sigma <- 0.05

G          <- matrix(rnorm(nobs * npar), nobs, npar)
theta_true <- c(1.0, -0.5, 2.0, 0.25)
y          <- as.numeric(G %*% theta_true) + rnorm(nobs, sd = sigma)
names(y)   <- paste0("o", seq_len(nobs))

# Forward model: any function that takes an (nreal x npar) matrix and
# returns an (nreal x nobs) matrix.
forward <- function(theta) theta %*% t(G)

prior <- matrix(rnorm(nreal * npar), nreal, npar,
                dimnames = list(NULL, paste0("p", seq_len(npar))))
```

Run six IES iterations with a single lambda:

```{r}
fit <- pesto_ies_callback(
  forward_model  = forward,
  prior_ensemble = prior,
  obs            = y,
  obs_sd         = sigma,
  noptmax        = 6L,
  lambda         = 1.0,
  verbose        = FALSE
)
```

Diagnostics:

```{r}
prior_rmse <- sqrt(mean((colMeans(prior) - theta_true)^2))
post_mean  <- colMeans(as.matrix(fit$par_ensemble[, -1L]))
post_rmse  <- sqrt(mean((post_mean - theta_true)^2))

cat(sprintf("prior RMSE: %.4f\n", prior_rmse))
cat(sprintf("posterior RMSE: %.4f\n", post_rmse))

mean_phi <- vapply(fit$iterations, `[[`, numeric(1L), "mean_phi")
print(mean_phi)
```

Posterior RMSE should be a small fraction of prior RMSE, and mean phi
should decrease across iterations.

The full per-realisation phi trajectory:

```{r, fig.alt = "Per-realisation phi decreasing across IES iterations"}
plot_phi(fit)
```

## Driving APSIM through `apsim_callback()`

The adapter wraps `apsimx` so each realisation gets its own working
copy of an APSIM template, parameter edits per `param_map`, a run, and
an extraction step. The signature returned matches what
`pesto_ies_callback()` expects.

```{r,eval=FALSE}
library(apsimx)   # >= 2.7.0 from CRAN
# Requires a working APSIM Next Gen installation.

forward <- apsim_callback(
  template  = "wheat_wagga.apsimx",
  param_map = list(
    RUE = "Wheat.Leaf.Photosynthesis.RUE.FixedValue",
    CN2 = "Soil.SoilWater.CN2Bare"
  ),
  output_extractor = function(sim) {
    # sim is the data.frame of report-table variables.
    as.numeric(sim$Wheat.Grain.Total.Wt)
  }
)

prior <- cbind(
  RUE = runif(40, 1.0, 2.0),
  CN2 = runif(40, 60,  90)
)

fit <- pesto_ies_callback(
  forward_model  = forward,
  prior_ensemble = prior,
  obs            = c(y_2018 = 4500, y_2019 = 5200, y_2020 = 4900),
  obs_sd         = 250,
  noptmax        = 4L
)

summary(fit$par_ensemble)
```

### Failure handling

Per-realisation APSIM crashes (corrupt config, solver divergence,
missing report variable) are caught by the adapter and emerge as `NA`
rows. `pesto_ies_callback()`'s `on_failure = "na"` (default) carries
those realisations forward unchanged so the ensemble survives partial
failures; `on_failure = "stop"` aborts as soon as one row is missing.

### Concurrency

Phase-1 D4 runs realisations serially. Parallel evaluation (`future`,
`mirai`, or `apsimx`'s own job-server modes) is a planned follow-up
once `apsimx`'s thread-safety under ensemble load has been
characterised. For pure-R synthetic models the serial loop is already
fast enough that the overhead is dominated by `ensemble_solution()`
itself.

## When to prefer `pesto_ies()` instead

Use the classic `.pst` path when:

- Your forward model lives behind a non-R executable that PEST++ can
  drive directly via its template/instruction-file mechanism.
- You need full bit-for-bit compatibility with `pestpp-ies` behaviour
  (e.g. for cross-validating against an existing PEST++ workflow).
- You depend on `pestpp-ies`-specific features not yet exposed by the
  R driver (full lambda line-search, regularisation modes,
  Tikhonov priors).

Otherwise the callback driver is faster to iterate, easier to debug,
and avoids per-realisation file I/O entirely.

## Reproducibility

```{r}
sessionInfo()
```
