---
title: "Wiring your own simulator into PESTO"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Wiring your own simulator into PESTO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render(); macOS arm64 (Darwin 26.5),
     R 4.5.2, one core. No external simulator runs: the APSIM adapter is
     shown but not executed, and every timed block is a synthetic or ODE
     forward model. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(PESTO)
```

## Why

My simulator is APSIM, one run takes about a minute, and I want to calibrate
it from R against a handful of seasons of measured yield. The question is:
*how do I wire a simulator like that into PESTO without paying a file
round-trip on every realisation, and what do I have to get right for the
posterior that comes back to be worth anything?* This vignette answers both.
The wiring is a typed forward-model contract that any R function can satisfy,
demonstrated end to end on problems that run in seconds; the thing to get
right is the observation standard deviation, and the middle of this vignette
measures exactly how wrong an answer becomes when it is set the way most
people first set it.

## What

[`pesto_ies_callback()`](../html/pesto_ies_callback.html) is the in-process
driver. [`pesto_ies()`](../html/pesto_ies.html) shells out to the
`pestpp-ies` binary, which writes and reads control, ensemble and
observation files between every realisation; for an R forward model, whether
that is an `apsimx` wrapper for APSIM, a bridge to another language, or a
fast synthetic test problem, the file round-trip is pure overhead. The
callback driver keeps the same C++ ensemble kernel
([`ensemble_solution()`](../html/ensemble_solution.html), after Chen and
Oliver 2013) and drives the outer loop in R, calling the forward model in
process. For APSIM this is the difference between re-reading an `.apsimx`
file for every realisation and holding the model in memory across the whole
ensemble.

[`pesto_forward_model()`](../html/pesto_forward_model.html) is the typed
contract a forward model travels in. A bare `function(theta) -> obs` is the
quickest way in and carries no metadata: the driver cannot know the output
dimension in advance, cannot run realisations in parallel, and cannot
express a failure budget. The contract object declares all three, and
[`pesto_evaluate()`](../html/pesto_evaluate.html) runs it.
[`apsim_callback()`](../html/apsim_callback.html) builds one for APSIM
through the `apsimx` package: a working copy of a template per realisation,
parameter edits by a `param_map`, a run, and an extraction step.

[`pesto_multifidelity_model()`](../html/pesto_multifidelity_model.html)
stacks an ordered set of fidelity levels, cheapest first, so early
iterations can explore on a coarse model and late ones sharpen against the
expensive one; [`mf_control_variate()`](../html/mf_control_variate.html) is
the variance-minimising affine correction that lifts cheap outputs toward
expensive ones. [`ode_forward_model()`](../html/ode_forward_model.html) and
its two specialisations,
[`crop_growth_forward_model()`](../html/crop_growth_forward_model.html) and
[`seir_forward_model()`](../html/seir_forward_model.html), are ready-made
contracts for the large class of simulators that are a small system of
ordinary differential equations rather than an executable.
[`pesto_ies_filter()`](../html/pesto_ies_filter.html) assimilates
time-ordered observation windows one after another rather than in one batch,
which is what an in-season update needs.

Two diagnostics run throughout.
[`ensemble_spread_ess()`](../html/ensemble_spread_ess.html) is recorded on
every iteration, and [`pesto_abstention()`](../html/pesto_abstention.html) is
the typed decline a reliability gate returns in place of a result.

## Do

### A synthetic recovery example

The problem below is synthetic: a linear forward model at a known truth with
Gaussian noise, which is the cheapest way to check that a driver recovers
what it should before a real simulator is attached.

```{r problem}
set.seed(20260516L)

npar <- 4L
nobs <- 8L
nreal <- 120L
sigma <- 0.05

G <- matrix(rnorm(nobs * npar), nobs, npar)
theta_true <- c(1.0, -0.5, 2.0, 0.25)
y <- as.numeric(G %*% theta_true) + rnorm(nobs, sd = sigma)
names(y) <- paste0("o", seq_len(nobs))

# Forward model: any function taking an (nreal x npar) matrix and returning
# an (nreal x nobs) matrix.
forward <- function(theta) theta %*% t(G)

prior <- matrix(rnorm(nreal * npar), nreal, npar,
                dimnames = list(NULL, paste0("p", seq_len(npar))))
```

Six iterations. `lambda` is left at its default: under the default ES-MDA
scheme the Marquardt damping *is* the inflation factor
(`alpha = lambda + 1`) and is set by the schedule, so it is chosen for you
rather than fixed by hand (see
[`?pesto_ies_callback`](../html/pesto_ies_callback.html)).

```{r fit}
fit <- pesto_ies_callback(
  forward_model  = forward,
  prior_ensemble = prior,
  obs            = y,
  obs_sd         = sigma,
  noptmax        = 6L,
  seed           = 42L,
  verbose        = FALSE
)

prior_rmse <- sqrt(mean((colMeans(prior) - theta_true)^2))
post_mean <- colMeans(as.matrix(fit$par_ensemble[, -1L]))
post_rmse <- sqrt(mean((post_mean - theta_true)^2))
round(c(prior_rmse = prior_rmse, posterior_rmse = post_rmse), 4)
vec_mean_phi <- vapply(fit$iterations, `[[`, numeric(1L), "mean_phi")
round(vec_mean_phi, 3)
```

```{r phi-figure, fig.height = 4.2, fig.cap = "Objective function by iteration on the synthetic recovery problem. The ribbon spans the ensemble's minimum to maximum misfit and each faint line is one realisation; the mean falls by nearly three orders of magnitude between the first and second iterations and then declines slowly."}
tab_phi_wide <- data.table::dcast(
  fit$phi, iteration ~ realisation, value.var = "phi"
)
plot_phi(tab_phi_wide, show_reals = TRUE,
         title = "Misfit by iteration, synthetic recovery problem")
```

### The forward-model contract

`pesto_forward_model()` wraps the same callable in a typed contract that
carries the metadata a bare function cannot. It is the single object both
the native callback driver here and the classic `.pst` path are built to
honour, so a forward model travels between modes unchanged.

```{r contract}
fm <- pesto_forward_model(
  fn          = forward,
  n_obs       = nobs,
  param_names = paste0("p", seq_len(npar)),
  on_failure  = "na"
)

# A contract object is evaluated directly with pesto_evaluate(); the
# returned matrix carries the per-realisation failure count.
sim <- pesto_evaluate(fm, prior)
c(realisations = nrow(sim), outputs = ncol(sim),
  failures = attr(sim, "n_failures"))

# Passing the contract to the driver is equivalent to passing the bare
# function: the bare form is auto-wrapped internally.
fit_typed <- pesto_ies_callback(fm, prior, y, sigma, noptmax = 6L,
                                seed = 42L, verbose = FALSE)
identical(
  as.matrix(fit_typed$par_ensemble[, -1L]),
  as.matrix(fit$par_ensemble[, -1L])
)
```

### Parallel, fault-tolerant ensembles

For an expensive forward model, an APSIM ensemble especially, the
realisations within one iteration are embarrassingly parallel. Build the
contract with `parallel = "multicore"` and the evaluation engine dispatches
rows across forked workers through `parallel::mclapply()`. For reproducible
draws, set an `"L'Ecuyer-CMRG"` generator first, so each realisation
receives an independent stream.

```{r parallel, eval = FALSE}
RNGkind("L'Ecuyer-CMRG")
set.seed(42L)

fm_par <- pesto_forward_model(
  fn       = forward,
  n_obs    = nobs,
  parallel = "multicore",
  n_cores  = 4L
)
fit_par <- pesto_ies_callback(fm_par, prior, y, sigma, noptmax = 6L)
```

The chunk above is shown rather than run, because forking inside a vignette
build is not portable. A custom `map_fn`, any function shaped like `lapply`,
is the escape hatch for a cross-platform or cluster backend. Failures are
governed by `on_failure`, where `"na"` tolerates and `"stop"` aborts, and by
`max_fail_frac`, which aborts once the fraction of failed realisations in a
single evaluation exceeds the budget: a guardrail against an ensemble
collapsing silently.

### The number that decides whether to believe the answer

The single most consequential argument is `obs_sd`, the observation standard
deviation, which sets the weights $w = 1 / \sigma_{\mathrm{obs}}$. It is also
the easiest to get wrong in a way that fails silently. The failure mode is
*over-determination*: conditioning the ensemble on a likelihood far tighter
than the data deserve. The smoother then drives every realisation onto the
same point, the posterior spread collapses, and the run reports an answer
whose credible interval is far too narrow to be credible.

The usual route into the trap is statistical rather than a typo. Suppose
each target observation is itself the mean of $m$ field replicates. The
*replicate* spread, the genuine measurement-plus-process noise a single
modelled value must match, is $\sigma$. The *standard error of that mean* is
$\sigma / \sqrt{m}$, smaller by a factor of $\sqrt{m}$ and shrinking further
as more plots are averaged. Passing the standard error as `obs_sd` tells the
smoother the data pin the model roughly $\sqrt{m}$ times more sharply than
they do. The forward model cannot be that right, so the ensemble has nowhere
to go but onto a single collapsed point.

```{r over-determination}
m <- 40L
obs_se <- sigma / sqrt(m)   # the standard error of a 40-replicate mean

fit_ok <- pesto_ies_callback(
  forward, prior, y,
  obs_sd  = sigma,          # field-realistic replicate spread
  noptmax = 6L, seed = 42L, verbose = FALSE
)

fit_collapsed <- pesto_ies_callback(
  forward, prior, y,
  obs_sd  = obs_se,         # the over-precise likelihood, the trap
  noptmax = 6L, seed = 42L, verbose = FALSE
)
```

The collapse does not announce itself in the posterior *mean*. On this
linear-Gaussian problem both runs recover the truth to three figures. It
announces itself in the posterior *spread*, and in whether the credible
interval still covers the truth.

```{r over-determination-diagnose}
spread <- function(fit) {
  mean(apply(as.matrix(fit$par_ensemble[, -1L]), 2L, sd))
}

coverage_90 <- function(fit, truth) {
  pe <- as.matrix(fit$par_ensemble[, -1L])
  lo <- apply(pe, 2L, quantile, 0.05)
  hi <- apply(pe, 2L, quantile, 0.95)
  mean(truth >= lo & truth <= hi)   # fraction of params inside the band
}

tab_two <- data.frame(
  Case = c("field-realistic", "over-precise (SE of a 40-replicate mean)"),
  `obs_sd` = c(sigma, obs_se),
  `Mean posterior SD` = c(spread(fit_ok), spread(fit_collapsed)),
  `Coverage of the 90% band (%)` = 100 * c(
    coverage_90(fit_ok, theta_true), coverage_90(fit_collapsed, theta_true)
  ),
  check.names = FALSE
)
```

```{r over-determination-table, echo = FALSE}
knitr::kable(
  tab_two, digits = 4,
  caption = paste0(
    "The same problem and the same data under a field-realistic and an ",
    "over-precise observation standard deviation."
  )
)
```

### Where the boundary is

Two cases show that the failure exists. A sweep shows where it starts. The
block below repeats the run at replicate counts from 1 to 160, each time
passing the standard error of an $m$-replicate mean as `obs_sd`, and records
two things: the mean posterior standard deviation, and how far the posterior
mean sits from the truth measured in its own posterior standard deviations.
A calibrated posterior puts the truth within about one of its own standard
deviations; a collapsed one puts it many.

```{r boundary-sweep}
vec_m <- c(1L, 2L, 5L, 10L, 20L, 40L, 80L, 160L)
tab_boundary <- data.table::rbindlist(lapply(vec_m, function(mm) {
  fit_m <- pesto_ies_callback(
    forward, prior, y, obs_sd = sigma / sqrt(mm),
    noptmax = 6L, seed = 42L, verbose = FALSE
  )
  pe <- as.matrix(fit_m$par_ensemble[, -1L])
  par_sd <- apply(pe, 2L, sd)
  data.table::data.table(
    m = mm,
    obs_sd = sigma / sqrt(mm),
    post_sd = mean(par_sd),
    z = mean(abs(colMeans(pe) - theta_true) / par_sd),
    coverage = coverage_90(fit_m, theta_true)
  )
})) # ends the sweep over replicate counts
```

```{r boundary-figure, fig.height = 4.5, fig.cap = "The over-determination boundary. The horizontal axis is the number of replicates whose standard error was passed as obs_sd; the vertical axis is how far the posterior mean sits from the known truth, measured in posterior standard deviations. Point colour is the fraction of parameters whose 90 per cent credible band still covers the truth. The dashed line at one standard deviation is where a posterior stops being able to describe its own error, and everything to the right of the crossing is a run that reports false confidence."}
ggplot2::ggplot(
  tab_boundary, ggplot2::aes(x = m, y = z, colour = coverage)
) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dashed",
                      colour = "grey40") +
  ggplot2::geom_line(colour = "grey60", linewidth = 0.8) +
  ggplot2::geom_point(size = 3.4) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_viridis_c(
    name = "90% coverage", limits = c(0, 1), end = 0.9
  ) +
  ggplot2::labs(
    x = "replicates averaged into each observation (log scale)",
    y = "distance from truth, in posterior SDs (log scale)",
    title = "How over-precise an obs_sd can be before the answer misleads"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

```{r boundary-table, echo = FALSE}
knitr::kable(
  tab_boundary, digits = 4,
  col.names = c("Replicates m", "obs_sd passed", "Mean posterior SD",
                "Distance from truth (posterior SDs)",
                "90% coverage (fraction)"),
  caption = paste0(
    "The same inversion at eight settings of obs_sd, each the standard ",
    "error of a mean over m replicates of a spread of ", sigma, "."
  )
)
```

PESTO does not gate this automatically: `pesto_abstention()`'s reason codes
are not a closed set, and a caller who can measure the boundary can mint the
gate. The function below is that gate, written against the sweep's own
statistic, and it returns a typed decline that a downstream tool recognises
by class rather than by parsing a message.

```{r minted-abstention}
gate_over_determination <- function(fit, truth, z_max = 1.5) {
  pe <- as.matrix(fit$par_ensemble[, -1L])
  par_sd <- apply(pe, 2L, sd)
  z <- mean(abs(colMeans(pe) - truth) / par_sd)
  if (z > z_max) {
    return(pesto_abstention(
      "over_determined",
      detail = sprintf(
        paste0("posterior mean sits %.1f posterior SDs from the reference, ",
               "above the tolerance of %.1f; obs_sd is too small"),
        z, z_max
      ),
      diagnostics = list(z = round(z, 2), mean_post_sd = mean(par_sd))
    ))
  }
  fit
}

is_pesto_abstention(gate_over_determination(fit_ok, theta_true))
gate_over_determination(fit_collapsed, theta_true)
```

### What the spread-ESS diagnostic can and cannot see

`ensemble_spread_ess()` is logged on every iteration as `spread_ess` and as
the ratio `spread_ess_ratio`, and it is worth being precise about its reach.
It is the spectral participation ratio of the parameter anomaly covariance,
the effective number of variance-carrying directions, so it detects the
*shape* of a collapse: variance draining into a few directions. It is a
ratio of eigenvalue sums, hence invariant to a global rescaling of the
anomalies, so a spread that is uniformly too small in every direction is
invisible to it. The two runs above differ in exactly that way.

```{r ess-blind}
ess_ratio <- function(fit) {
  vapply(fit$iterations, `[[`, numeric(1L), "spread_ess_ratio")
}
tab_ess <- rbind(
  `field-realistic` = round(ess_ratio(fit_ok), 3),
  `over-precise`    = round(ess_ratio(fit_collapsed), 3)
)
tab_ess
```

Read `spread_ess_ratio` as the directional-collapse diagnostic it is, and
read the coverage column of the boundary table as the calibration check.
They answer different questions and only the second catches a mis-specified
`obs_sd`.

The fix is to set `obs_sd` to the uncertainty the modelled value must
actually reproduce, the replicate-level measurement-plus-process spread,
never the standard error of an average:

- Use the replicate standard deviation $\sigma$, not $\sigma / \sqrt{m}$,
  when the target is a mean over $m$ measurements.
- Fold structural model discrepancy into `obs_sd`. The forward model is an
  approximation, and the likelihood should not pretend it matches reality
  more tightly than the model can.
- When in genuine doubt, err wide. An over-wide `obs_sd` leaves spread on
  the table and under-uses the data; an over-narrow one destroys the
  ensemble and reports false confidence.

When over-determination is unavoidable, with many observations and few
realisations, the covariance inflation and localisation countermeasures in
*My posterior looks too confident: is the ensemble collapsing?* replenish
the collapsing spread rather than only diagnosing it. A correctly specified
`obs_sd` is the first line of defence and those countermeasures the second.

### Driving APSIM through `apsim_callback()`

The adapter wraps `apsimx` so each realisation gets its own working copy of
an APSIM template, parameter edits per `param_map`, a run, and an extraction
step. What it returns matches what `pesto_ies_callback()` expects.

```{r apsim-probe, echo = FALSE, results = "asis"}
has_apsimx <- requireNamespace("apsimx", quietly = TRUE)
has_apsim_exe <- nzchar(Sys.getenv("PESTO_APSIM_EXE"))
if (has_apsimx && has_apsim_exe) {
  cat("The `apsimx` package and an APSIM installation are both visible on",
      "this machine, so the block below would run; it is shown rather than",
      "executed so that this vignette builds identically everywhere.\n")
} else if (has_apsimx) {
  cat("The `apsimx` package is installed here but no APSIM executable is",
      "advertised through `PESTO_APSIM_EXE`, so the block below is shown",
      "rather than run.\n")
} else {
  cat("Neither the `apsimx` package nor an APSIM installation is present on",
      "this machine, so the block below is shown rather than run. Every",
      "other result in this vignette is computed without them.\n")
}
```

```{r apsim-adapter, eval = FALSE}
library(apsimx)   # >= 2.7.0 from CRAN; needs a working APSIM Next Gen

forward_apsim <- apsim_callback(
  template  = "wheat_wagga.apsimx",
  param_map = list(
    RUE = "Wheat.Leaf.Photosynthesis.RUE.FixedValue",
    CN2 = "Soil.SoilWater.CN2Bare"
  ),
  output_extractor = function(sim) {
    # sim is the data.frame of report-table variables.
    as.numeric(sim$Wheat.Grain.Total.Wt)
  }
)

prior_apsim <- cbind(
  RUE = runif(40L, 1.0, 2.0),
  CN2 = runif(40L, 60, 90)
)

fit_apsim <- pesto_ies_callback(
  forward_model  = forward_apsim,
  prior_ensemble = prior_apsim,
  obs            = c(y_2018 = 4500, y_2019 = 5200, y_2020 = 4900),
  obs_sd         = 250,
  noptmax        = 4L
)
```

Per-realisation APSIM crashes, whether a corrupt configuration, a solver
divergence or a missing report variable, are caught by the adapter and
emerge as `NA` rows. The driver's default `on_failure = "na"` carries those
realisations forward unchanged so the ensemble survives partial failure;
`on_failure = "stop"` aborts as soon as one row is missing.

The closure writes each realisation to its own uniquely named working file,
so it is safe to evaluate in parallel: wrap it in a
`pesto_forward_model(parallel = "multicore")` exactly as above and the
ensemble runs across forked workers, each invoking APSIM on its own input.
Start with a modest `n_cores`, because the thread safety of `apsimx` itself
under heavy ensemble load has not been independently characterised. For
pure-R synthetic models the serial loop is already fast enough that the
overhead is dominated by `ensemble_solution()`.

### Multi-fidelity calibration

Process-based crop models expose a cost-accuracy dial: a daily time step
with a lite soil profile is cheap, a sub-daily step with the full profile is
expensive. `pesto_multifidelity_model()` makes that dial first class. It
bundles an ordered stack of fidelity levels, cheapest first, and the driver
picks a level per iteration through a `fidelity_schedule`, so early
iterations explore cheaply and late ones sharpen against the expensive
truth. The final ensemble is always refreshed at the highest fidelity.

```{r multifidelity}
# Cheap level: the linear model with a small systematic bias.
# Expensive level: the unbiased truth.
cheap_fn <- function(theta) theta %*% t(G) + 0.30
expensive_fn <- forward

mf <- pesto_multifidelity_model(
  levels = list(
    pesto_forward_model(fn = cheap_fn,     n_obs = nobs, fidelity = 0L),
    pesto_forward_model(fn = expensive_fn, n_obs = nobs, fidelity = 1L)
  ),
  costs = c(1, 25)   # the expensive level is about 25 times dearer per run
)

# Ramp: two cheap iterations, then four expensive ones.
fit_mf <- pesto_ies_callback(
  forward_model     = mf,
  prior_ensemble    = prior,
  obs               = y,
  obs_sd            = sigma,
  noptmax           = 6L,
  fidelity_schedule = c(0L, 0L, 1L, 1L, 1L, 1L),
  seed              = 42L,
  verbose           = FALSE
)

mf_rmse <- sqrt(mean(
  (colMeans(as.matrix(fit_mf$par_ensemble[, -1L])) - theta_true)^2
))
round(c(single_fidelity_rmse = post_rmse, multi_fidelity_rmse = mf_rmse), 4)
```

When the cheap level is run over the whole ensemble and the expensive level
over only a subset, `mf_control_variate()` lifts the cheap outputs toward
the expensive ones with the variance-minimising affine correction, which is
the plug-in primitive for a surrogate cascade:

```{r control-variate}
sub <- seq_len(20L)
low_all <- pesto_evaluate(mf, prior, level = 0L)
low_sub <- low_all[sub, , drop = FALSE]
high_sub <- pesto_evaluate(mf, prior[sub, , drop = FALSE], level = 1L)

corrected <- mf_control_variate(low_all, high_sub, low_sub)
high_all <- pesto_evaluate(mf, prior, level = 1L)
gap_before <- mean(abs(low_all - high_all))
gap_after <- mean(abs(corrected - high_all))
signif(c(mean_gap_before = gap_before, mean_gap_after = gap_after), 3)
```

### Simulators that are systems of equations

Not every simulator sits behind an executable. A very large class, crop
growth, epidemic dynamics, nutrient or solute transport and
pharmacokinetics among them, is a small system of ordinary differential
equations integrated forward in time with the observation vector read off
the state trajectory. PESTO ships these as templates returning the same
typed contract as everything above, so they plug straight into the driver,
the multi-fidelity stack and the manifest emitter. Integration is a
self-contained fixed-step fourth-order Runge-Kutta scheme by default, with a
`solver = "desolve"` path delegating to the `deSolve` package for stiff
systems.

The generic builder is `ode_forward_model()`: supply the derivative
function, the initial state, the time grid, and which parameter columns the
model consumes. `crop_growth_forward_model()` is the logistic dry-matter
curve $\mathrm{d}B / \mathrm{d}t = r B (1 - B / b_{\max})$, the canonical
sigmoid description of seasonal biomass, calibrating $r$, $b_{\max}$ and the
starting biomass against an observed biomass series.

```{r ode-crop}
times <- seq(0, 120, by = 15)
crop <- crop_growth_forward_model(times = times)

# Simulate a biomass series at a known parameter, then add field noise.
theta_crop <- c(r = 0.06, b_max = 1400, b0 = 20)
biomass <- as.numeric(pesto_evaluate(
  crop, matrix(theta_crop, nrow = 1L,
               dimnames = list(NULL, names(theta_crop)))
))
y_crop <- biomass + rnorm(length(biomass), sd = 20)
names(y_crop) <- paste0("t", seq_along(y_crop))

prior_crop <- cbind(
  r     = runif(120L, 0.02, 0.12),
  b_max = runif(120L, 900, 2000),
  b0    = runif(120L, 5, 60)
)
fit_crop <- pesto_ies_callback(crop, prior_crop, y_crop, obs_sd = 20,
                              noptmax = 8L, seed = 42L, verbose = FALSE)
post_crop <- colMeans(
  as.matrix(fit_crop$par_ensemble[, -1L])
)[names(theta_crop)]
```

```{r ode-crop-table, echo = FALSE}
knitr::kable(
  data.frame(
    Parameter = names(theta_crop),
    Unit = c("per day", "g/m2", "g/m2"),
    Truth = as.numeric(theta_crop),
    `Posterior mean` = as.numeric(post_crop),
    check.names = FALSE
  ),
  digits = 3,
  caption = paste0(
    "Recovery of the logistic growth parameters from a noisy biomass series."
  )
)
```

`seir_forward_model()` is the closed-population
susceptible-exposed-infectious-recovered epidemic model, calibrating the
transmission, latency and recovery rates against an observed infectious
prevalence curve. The reproduction number $R_0 = \beta / \gamma$ is the
summary an outbreak curve identifies most sharply.

```{r ode-seir}
days <- seq(0, 60, by = 5)
seir <- seir_forward_model(times = days, n_pop = 1000, i0 = 1)

theta_seir <- c(beta = 0.6, sigma = 0.2, gamma = 0.1)
prevalence <- as.numeric(pesto_evaluate(
  seir, matrix(theta_seir, nrow = 1L,
               dimnames = list(NULL, names(theta_seir)))
))
y_seir <- prevalence + rnorm(length(prevalence), sd = 3)
names(y_seir) <- paste0("d", seq_along(y_seir))

prior_seir <- cbind(
  beta  = runif(200L, 0.3, 0.9),
  sigma = runif(200L, 0.1, 0.4),
  gamma = runif(200L, 0.05, 0.2)
)
fit_seir <- pesto_ies_callback(seir, prior_seir, y_seir, obs_sd = 3,
                               noptmax = 10L, seed = 42L, verbose = FALSE)
post_seir <- colMeans(
  as.matrix(fit_seir$par_ensemble[, -1L])
)[names(theta_seir)]
r0_truth <- theta_seir[["beta"]] / theta_seir[["gamma"]]
r0_post <- post_seir[["beta"]] / post_seir[["gamma"]]
round(c(R0_truth = r0_truth, R0_posterior = r0_post), 2)
```

Because both templates return a `pesto_forward_model()`, the coarse and fine
fidelity trick from the previous section applies verbatim: build one
template at a coarse `n_steps` and one at a fine `n_steps`, stack them in a
`pesto_multifidelity_model()`, and ramp the integration resolution across
iterations.

### Assimilating a season as it happens

`pesto_ies_callback()` is a *smoother*: it assimilates every observation in
one batch. For an in-season setting, where a season's observations arrive
over time and the parameter posterior is wanted as the season progresses
rather than only at harvest, `pesto_ies_filter()` is the *filter*. It
assimilates time-ordered observation windows one after another against the
same static calibration parameters, and each window's posterior becomes the
next window's prior, so information accrues and the posterior tightens
window by window.

```{r filter}
# Split the eight observations into three temporal windows, standing in for
# early, mid and late season.
windows <- list(1:3, 4:6, 7:8)

fit_seq <- pesto_ies_filter(
  forward_model  = forward,
  prior_ensemble = prior,
  obs            = y,
  obs_sd         = sigma,
  windows        = windows,
  seed           = 42L,
  verbose        = FALSE
)

sd_trace <- vapply(fit_seq$windows, function(w) mean(w$par_sd), numeric(1L))
tab_filter <- data.frame(
  Window = seq_along(sd_trace),
  `Observations assimilated` = vapply(
    fit_seq$windows, function(w) length(w$obs_indices), integer(1L)
  ),
  `Mean parameter SD` = sd_trace,
  check.names = FALSE
)
```

```{r filter-table, echo = FALSE}
knitr::kable(
  tab_filter, digits = 4,
  caption = paste0(
    "The filter's posterior spread after each window. The value of the ",
    "filter is the trajectory, not only the endpoint."
  )
)
```

A multi-fidelity stack plugs in through `fidelity_schedule`, one level per
window, so a run can be cheap early and sharpen late, and the result wraps
into the same manifest contract through `as_manifest()`, tagged
`method = "ies_filter"`.

### When to prefer the `.pst` path instead

Use `pesto_ies()` and the classic control-file path when the forward model
lives behind a non-R executable that PEST++ can drive directly through its
template and instruction files; when bit-for-bit compatibility with
`pestpp-ies` behaviour is needed, for instance to cross-validate against an
existing PEST++ workflow; or when a `pestpp-ies` feature not yet exposed by
the R driver is required, such as the full lambda line search,
regularisation modes or Tikhonov priors. Otherwise the callback driver is
faster to iterate, easier to debug, and avoids per-realisation file
exchange entirely.

```{r read-facts, include = FALSE}
phi_first_txt <- format(signif(vec_mean_phi[1L], 4L), big.mark = ",",
                        scientific = FALSE)
gap_before_txt <- format(signif(gap_before, 3L), scientific = FALSE)
gap_after_txt <- sprintf("%.0e", gap_after)
```

## Read

The driver recovers what it should. Posterior root-mean-square error to the
truth is `r round(post_rmse, 4)` against a prior figure of
`r round(prior_rmse, 3)`, a reduction of a factor of
`r round(prior_rmse / post_rmse)`, and the ensemble mean misfit falls from
`r phi_first_txt` to
`r signif(vec_mean_phi[length(vec_mean_phi)], 3)` across
`r length(vec_mean_phi)` iterations. Handing the driver a typed
contract instead of a bare function returns a posterior ensemble identical
to the bare-function run, which is the property that lets a forward model
move between calling conventions unchanged.

The `obs_sd` sweep is the centre of this vignette. Passing the replicate
spread gives a mean posterior standard deviation of
`r signif(tab_boundary$post_sd[1], 3)` and a 90 per cent band that covers
`r round(100 * tab_boundary$coverage[1])` per cent of the parameters.
Passing the standard error of a 40-replicate mean, about
`r round(sqrt(40))` times smaller, shrinks the posterior standard deviation
to `r signif(spread(fit_collapsed), 3)`, roughly
`r round(100 * spread(fit_collapsed) / spread(fit_ok))` per cent of the
well-specified run, and drops coverage to
`r round(100 * coverage_90(fit_collapsed, theta_true))` per cent. The figure
shows that this is not a cliff but a slope with a crossing point: the
distance from the truth in posterior standard deviations rises from
`r round(tab_boundary$z[1], 2)` at one replicate to
`r round(tab_boundary$z[nrow(tab_boundary)], 1)` at
`r tab_boundary$m[nrow(tab_boundary)]`, crossing one standard deviation
between `r tab_boundary$m[max(which(tab_boundary$z < 1))]` and
`r tab_boundary$m[min(which(tab_boundary$z > 1))]` replicates. Beyond that
crossing the posterior can no longer describe its own error, and coverage
falls from `r round(100 * max(tab_boundary$coverage))` per cent to
`r round(100 * min(tab_boundary$coverage))` per cent across the sweep.

The spread-ESS trace explains why a diagnostic is not automatically a
safeguard. The two runs' trajectories sit within
`r round(max(abs(ess_ratio(fit_ok) - ess_ratio(fit_collapsed))), 3)` of each
other at every iteration even though one ensemble's spread is a small
fraction of the other's, because the statistic is invariant to a global
rescaling of the anomalies. That is the case for reading it as a
directional-collapse diagnostic and reading coverage as the calibration
check. Because PESTO does not gate on the sweep's statistic itself, the
`gate_over_determination()` function above mints the decline a caller wants:
applied to the well-specified run it returns the run, and applied to the
collapsed one it returns a `pesto_abstention` with reason
`over_determined`, which a downstream tool recognises by class rather than
by reading a message.

The remaining machinery does what it claims, with one price worth naming.
The multi-fidelity stack, running two cheap iterations before four expensive
ones, reaches a posterior root-mean-square error of `r round(mf_rmse, 4)`
against the single-fidelity run's `r round(post_rmse, 4)`: still far better
than the prior, but a factor of `r round(mf_rmse / post_rmse, 1)` worse than
spending every iteration on the expensive level. Two of six iterations spent
on a biased model is not free, and whether that trade is worth taking
depends on the cost ratio between the levels, which here was declared as 25
to 1. The control-variate correction, by contrast, is close to exact on this
problem: the mean absolute gap between the cheap and expensive levels falls
from `r gap_before_txt` to `r gap_after_txt`, which is machine precision,
because the cheap level's bias is a constant offset and an affine correction
removes a constant offset exactly. The logistic growth template recovers a
ceiling of
`r round(post_crop[["b_max"]])` g/m2 against a truth of
`r theta_crop[["b_max"]]` and a growth rate of
`r round(post_crop[["r"]], 3)` against `r theta_crop[["r"]]`. The epidemic
template recovers a reproduction number of `r round(r0_post, 2)` against a
truth of `r round(r0_truth, 1)`. The filter tightens the posterior window by
window, from a mean parameter standard deviation of
`r signif(sd_trace[1], 3)` after the first window to
`r signif(sd_trace[length(sd_trace)], 3)` after the third, which is the
trajectory an in-season user wants rather than the endpoint alone.

## Limits

Every problem run here is synthetic, and the central one is linear and
Gaussian, so the recovery it demonstrates is the easiest case there is;
nothing in this vignette establishes that a real APSIM calibration converges
or that its posterior is calibrated, which is what *Can PESTO
recover APSIM's own parameters?* takes up on a simulator-backed run and on
real measured biomass.
The APSIM adapter is shown and not executed, so its code path is documented
rather than exercised here, and the parallel evaluation path is likewise
shown rather than run. The over-determination boundary is measured against a
*known* truth, which a real calibration does not have: on a real problem the
same diagnosis has to be made from held-out predictive coverage rather than
from distance to a truth, and the minted gate above would have to be
rewritten around that substitute. The boundary's crossing point is a
property of this problem's parameter count, ensemble size and design matrix,
not a universal threshold, so the number to carry away is the method rather
than the value. Multi-fidelity is demonstrated with a cheap level that
differs from the expensive one by a constant offset, which is the friendliest
possible discrepancy; a cheap level whose bias varies with the parameters
would behave differently and is not tested here. The filter is run on
observations split into windows after the fact rather than arriving over a
real season, so it shows the mechanism and not the operational case.

## What to read next

*Can PESTO recover APSIM's own parameters?* is this vignette's counterpart on a
real simulator: a synthetic-truth recovery experiment against the bundled
APSIM Wheat example, then a calibration to real measured biomass
cross-checked against an independent optimiser. *My posterior looks too
confident: is the ensemble collapsing?* is the second line of defence
against a narrow posterior, and separates finite-ensemble collapse from the
mis-specification diagnosed here. *When can a surrogate stand in for the
simulator?* is the other way to spend fewer expensive model calls.
*Handing an ensemble to somebody else* documents the record a finished run
emits. *Is PESTO the same algorithm as the tools it descends from?* measures what the
in-process driver buys against the file-coupled binaries.

## References

- Chen, Y., & Oliver, D. S. (2013). Levenberg-Marquardt forms of the
  iterative ensemble smoother for efficient history matching and uncertainty
  quantification. *Computational Geosciences*, 17(4), 689-703.
- Emerick, A. A., & Reynolds, A. C. (2013). Ensemble smoother with multiple
  data assimilation. *Computers & Geosciences*, 55, 3-15.
- Glasserman, P. (2003). *Monte Carlo Methods in Financial Engineering*.
  Springer, New York.
- Holzworth, D., et al. (2014). APSIM: evolution towards a new generation of
  agricultural systems simulation. *Environmental Modelling & Software*, 62,
  327-350.
- Kennedy, M. C., & O'Hagan, A. (2000). Predicting the output from a complex
  computer code when fast approximations are available. *Biometrika*, 87(1),
  1-13.

## Reproduce

Seed `20260516` sets the design matrix, the truth and the observation noise;
seed `42` is passed to every `pesto_ies_callback()` and
`pesto_ies_filter()` call, where it fixes the per-realisation observation
perturbation, so all runs are compared on identical perturbed data. No
external simulator or binary is used. Package versions follow.

```{r session-info}
sessionInfo()
```
