---
title: "Handing an ensemble to somebody else"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Handing an ensemble to somebody else}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64 (Darwin 26.5),
     R 4.5.2, one core. The inversion is under a second; the rest is
     serialisation, hashing and one figure. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(PESTO)
```

## Why

I have calibrated my crop model and the posterior ensemble now has to go to
somebody else: an agronomist who will price a decision on it, a colleague
who will re-run it next season, a reviewer who will ask whether the numbers
in my figure are the numbers the code produced. The question is: *what
exactly do I hand over, so that the person receiving it can read the
ensemble without knowing anything about PESTO's internals, and can check
that what they are holding is the run I actually made?* This vignette
answers it with the object PESTO emits for the purpose, and shows what
happens when a single number in the handed-over ensemble is altered.

## What

A finished run wraps into a `pesto_ensemble_manifest`, an S7 class carrying
the parameter ensemble, the simulated outputs, the assimilation context
(weights and target observations), the seed, the method, the lambda
schedule, the failure rate, provenance strings, and a SHA-256 hash over the
data. It is a versioned *data contract*, not a convenience wrapper: PESTO
owns it as the orchestra's C2 contract, and consuming members read it
through the four verbs below rather than through PESTO's list internals.

[`as_manifest()`](../html/as_manifest.html) builds one from a run.
[`write_manifest()`](../html/write_manifest.html) serialises it as a YAML
entry plus sidecar files, in one of three formats whose integrity contract
differs. [`read_manifest()`](../html/read_manifest.html) reads it back.
[`verify_manifest()`](../html/verify_manifest.html) recomputes the hash from
what is on disk and reports whether it matches the hash the YAML records.
A consumer's obligation is to call `verify_manifest()` before trusting
anything else, and the full producer and consumer obligations are written
out as a standalone format specification in `inst/manifest_contract.md`.

`as_manifest()` also has a method for
[`pesto_abstention()`](../html/pesto_abstention.html), PESTO's typed decline,
so a run that a reliability gate refused still produces a record rather than
nothing at all.

## Do

### A run worth handing over

The run below is the synthetic crop-growth inversion from *Your first
inversion: from a season of measurements to a posterior*: three growth
parameters, twelve noisy biomass measurements, one hundred and twenty
realisations, eight smoother iterations. It is reproduced here so the
handover has something real to carry.

```{r run}
set.seed(20260902L)
vec_times <- seq(0, 120, by = 10)
vec_days <- vec_times[-1L]
crop <- crop_growth_forward_model(times = vec_times)

theta_true <- c(r = 0.065, b_max = 1450, b0 = 18)
mat_true <- matrix(theta_true, nrow = 1L,
                   dimnames = list(NULL, names(theta_true)))
vec_biomass <- as.numeric(pesto_evaluate(crop, mat_true))

obs_sd <- 45
y_obs <- vec_biomass + rnorm(length(vec_biomass), sd = obs_sd)
names(y_obs) <- paste0("t", seq_along(y_obs))

s1_real <- 120L
prior <- cbind(
  r     = runif(s1_real, 0.02, 0.12),
  b_max = runif(s1_real, 900, 2000),
  b0    = runif(s1_real, 5, 60)
)

fit <- pesto_ies_callback(
  forward_model  = crop,
  prior_ensemble = prior,
  obs            = y_obs,
  obs_sd         = obs_sd,
  noptmax        = 8L,
  seed           = 20260902L,
  verbose        = FALSE
)
```

### Wrapping it

```{r manifest}
man <- as_manifest(fit, seed = 20260902L, apsim_version = NA_character_)
man
```

Every slot is reachable through the standard S7 accessor, which is what a
consumer uses instead of reaching into a list:

```{r slots}
man@run_id
man@schema_version
man@data_hash
man@failure_rate
head(man@params, 3L)
```

```{r slot-table, echo = FALSE}
tab_slots <- data.frame(
  Slot = c("schema_version", "method", "noptmax", "seed", "failure_rate",
           "params", "outputs", "obs_target", "weights", "data_hash"),
  `What a consumer reads it for` = c(
    "which version of the contract this record obeys",
    "which algorithm produced the ensemble",
    "how many assimilation steps were taken",
    "the seed that reproduces the run",
    "the fraction of forward evaluations that returned nothing",
    "the posterior parameter ensemble",
    "the simulated outputs matching those parameters",
    "the observations the ensemble was conditioned on",
    "the observation weights, one over the observation SD",
    "the integrity check over all of the above"
  ),
  Value = c(
    man@schema_version, man@method, as.character(man@noptmax),
    as.character(man@seed), sprintf("%.4f", man@failure_rate),
    sprintf("%d x %d", nrow(man@params), ncol(man@params) - 1L),
    sprintf("%d x %d", nrow(man@outputs), ncol(man@outputs) - 1L),
    sprintf("%d values", length(man@obs_target)),
    sprintf("%d values", length(man@weights)),
    paste0(substr(man@data_hash, 1L, 22L), "...")
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(
  tab_slots,
  caption = paste0(
    "The manifest slots a downstream consumer reads, and what this ",
    "run put in them."
  )
)
```

### Writing, reading, verifying

`write_manifest()` emits the YAML entry plus three RDS sidecars, one for the
parameter ensemble, one for the outputs, and one for the assimilation
context. RDS rather than CSV, because IEEE 754 doubles round-trip
bit-exactly through it and the SHA-256 check would otherwise trip on
formatter precision loss alone.

```{r write}
dir_run <- tempfile("pesto_manifest_")
dir.create(dir_run)
vec_paths <- write_manifest(man, file.path(dir_run, "wagga_2026_run01.yaml"))
basename(vec_paths)
```

```{r read-verify}
man_read <- read_manifest(file.path(dir_run, "wagga_2026_run01.yaml"))
verify_manifest(man_read)$ok
```

The YAML entry is the human-readable front door; the first lines carry the
identity and provenance a reader wants before opening anything else:

```{r yaml-peek}
cat(paste(
  readLines(file.path(dir_run, "wagga_2026_run01.yaml"))[seq_len(14L)],
  collapse = "\n"
))
```

### Three formats, three integrity contracts

Some handovers need a file a collaborator can open in a spreadsheet.
`format = "both"` writes the CSV alongside the RDS: the hash stays bound to
the RDS, so integrity is preserved, and the YAML records the CSV paths under
a separate `inspection_csv:` block that a consumer can ignore.

```{r write-both}
dir_both <- tempfile("pesto_manifest_csv_")
dir.create(dir_both)
vec_paths_both <- write_manifest(
  man, file.path(dir_both, "wagga_2026_run01.yaml"), format = "both"
)
basename(vec_paths_both)
```

`format = "csv_unverified"` is the one-way export for an analyst who will
never load it back into R. It writes CSV sidecars only, records
`integrity: not_verifiable` in the YAML, and `verify_manifest()` then
returns `NA` rather than a spurious `FALSE`, because CSV formatter
precision loss of about one unit in the last place is enough on its own to
move the hash. The mode is named the way it is so the weaker contract is
visible at every call site.

```{r write-unverified}
dir_export <- tempfile("pesto_manifest_unverified_")
dir.create(dir_export)
write_manifest(man, file.path(dir_export, "snapshot.yaml"),
               format = "csv_unverified")
man_export <- read_manifest(file.path(dir_export, "snapshot.yaml"))
verify_manifest(man_export)$ok
verify_manifest(man_export)$message
```

```{r format-table, echo = FALSE}
tab_formats <- data.frame(
  Format = c("rds (default)", "both", "csv_unverified"),
  Sidecars = c("RDS only", "RDS plus inspection CSV", "CSV only"),
  `YAML integrity field` = c("verifiable", "verifiable", "not_verifiable"),
  `verify_manifest()$ok` = c(
    as.character(verify_manifest(man_read)$ok),
    "TRUE",
    as.character(verify_manifest(man_export)$ok)
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(
  tab_formats,
  caption = paste0(
    "The three sidecar formats and the integrity contract each offers."
  )
)
```

### What tampering looks like

The point of the hash is the case nobody notices: a downstream tool re-saves
the outputs sidecar after an accidental edit or a partial overwrite. Below,
one cell of the outputs ensemble is moved by 0.001 g/m2, far below the
measurement noise the run was conditioned on, and the file is written back.

```{r tamper}
path_outputs <- file.path(dir_run, "wagga_2026_run01_outputs.rds")
tab_outputs <- readRDS(path_outputs)
val_before <- tab_outputs[1L, 2L]
tab_outputs[1L, 2L] <- tab_outputs[1L, 2L] + 1e-3
saveRDS(tab_outputs, path_outputs, version = 3L)

man_tampered <- read_manifest(file.path(dir_run, "wagga_2026_run01.yaml"))
res_verify <- verify_manifest(man_tampered)
res_verify$ok
```

```{r tamper-numbers}
tab_orig <- as.matrix(man@outputs[, -1L])
tab_seen <- as.matrix(man_tampered@outputs[, -1L])
mean_shift <- abs(mean(tab_seen[, 1L], na.rm = TRUE) -
                    mean(tab_orig[, 1L], na.rm = TRUE))
signif(mean_shift, 3L)
```

```{r tamper-figure, fig.height = 4, fig.cap = "Cell-by-cell absolute difference between the ensemble written to disk and the ensemble read back after one value was altered. Every cell but one is identical to the last bit; the single altered cell sits at 0.001 g/m2, far below anything an ensemble summary would move. The dashed line marks the measurement noise the run was conditioned on."}
tab_diff <- data.table::data.table(
  cell = seq_len(length(tab_orig)),
  delta = as.numeric(abs(tab_seen - tab_orig))
)
tab_diff <- tab_diff[!is.na(delta)]
tab_diff[, altered := delta > 0]

ggplot2::ggplot(
  tab_diff, ggplot2::aes(x = cell, y = pmax(delta, 1e-18), colour = altered)
) +
  ggplot2::geom_point(size = 1.1, alpha = 0.8) +
  ggplot2::geom_hline(yintercept = obs_sd, linetype = "dashed",
                      colour = "grey40") +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_viridis_d(
    name = "cell altered", end = 0.8, labels = c("no", "yes")
  ) +
  ggplot2::labs(
    x = "cell of the outputs ensemble (realisation by observation)",
    y = "absolute difference on re-read (g/m2, log scale)",
    title = "What the hash sees and the eye does not"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

`verify_manifest()` returns the stored and the recomputed hash side by side,
so a consumer can fail fast and say exactly what diverged rather than
reporting a vague mismatch:

```{r tamper-hashes}
substr(res_verify$stored, 1L, 22L)
substr(res_verify$recomputed, 1L, 22L)
```

### A run that declined to happen

Not every call returns an ensemble. PESTO's reliability gates return a typed
`pesto_abstention()` instead of an error or a half-populated result, and the
manifest has a method for it: an empty-payload record whose `summary` slot
carries the decline. A consumer therefore sees a typed refusal in the same
place it would have seen a posterior, rather than a missing file.

```{r abstention}
mat_par_off <- matrix(rnorm(30L * 12L), 30L, 12L)
mat_obs_off <- matrix(rnorm(30L * 8L), 30L, 8L)
res_off <- pesto_surrogate_ies(
  par_ensemble = mat_par_off,
  obs_ensemble = mat_obs_off,
  obs_target   = rnorm(8L),
  weights      = rep(1, 8L),
  parcov_inv   = rep(1, 12L)
)
is_pesto_abstention(res_off)
res_off
```

```{r abstention-manifest}
man_declined <- as_manifest(res_off)
dim(man_declined@params)
man_declined@summary$abstained
man_declined@summary$reason

dir_declined <- tempfile("pesto_manifest_declined_")
dir.create(dir_declined)
write_manifest(man_declined, file.path(dir_declined, "declined.yaml"))
man_declined_read <- read_manifest(file.path(dir_declined, "declined.yaml"))
man_declined_read@summary$reason
verify_manifest(man_declined_read)$ok
```

### What a conformant consumer does

A consumer reads the YAML with `read_manifest()`, calls `verify_manifest()`
before touching anything, refuses on `integrity: not_verifiable`, on a
`summary` that reports an abstention, or on a `failure_rate` above whatever
it is willing to tolerate, branches on `schema_version` for forward
compatibility, and then works with the typed slots. It threads
`man@data_hash` and `man@run_id` into its own provenance so the lineage
closes. The short function below is that whole obligation:

```{r consumer}
consume <- function(path, max_failure = 0.05) {
  m <- read_manifest(path)
  v <- verify_manifest(m)
  if (isTRUE(is.na(v$ok))) {
    return("declined: integrity not verifiable")
  }
  if (!isTRUE(v$ok)) {
    return("declined: hash mismatch")
  }
  if (isTRUE(m@summary$abstained)) {
    return(paste0("declined: producer abstained (", m@summary$reason, ")"))
  }
  if (m@failure_rate > max_failure) {
    return("declined: failure rate above tolerance")
  }
  paste0("accepted: ", nrow(m@params), " realisations, run ", m@run_id)
}

consume(file.path(dir_both, "wagga_2026_run01.yaml"))
consume(file.path(dir_export, "snapshot.yaml"))
consume(file.path(dir_declined, "declined.yaml"))
consume(file.path(dir_run, "wagga_2026_run01.yaml"))
```

```{r read-facts, include = FALSE}
ok_read <- verify_manifest(man_read)$ok
ok_export <- verify_manifest(man_export)$ok
ok_tamper <- res_verify$ok
mean_shift_txt <- format(signif(mean_shift, 2L), scientific = FALSE)
orders_below <- round(log10(obs_sd / 1e-3), 1)
```

```{r cleanup, include = FALSE}
unlink(c(dir_run, dir_both, dir_export, dir_declined), recursive = TRUE)
```

## Read

The handover works, and it works in both directions. Writing the run
produced `r length(vec_paths)` files: one YAML entry and three sidecars.
Reading them back and recomputing the hash returned
`r ok_read`, so the ensemble a colleague opens is
bit-for-bit the ensemble that came out of the smoother. The record carries
`r nrow(man@params)` realisations of `r ncol(man@params) - 1L` parameters
against `r length(man@obs_target)` target observations, and a failure rate
of `r sprintf("%.3f", man@failure_rate)`, which is the fraction of forward
evaluations that returned nothing during the run and is exactly the kind of
thing a consumer should be allowed to refuse on.

The tamper case is the one that matters. Moving one value by 0.001 g/m2
shifts the ensemble mean of that observation column by
`r mean_shift_txt` g/m2, which is nothing at all beside the `r obs_sd` g/m2
of measurement noise, and the figure shows why no summary statistic would
ever raise it: every other cell is identical to the last bit, and the
altered one sits `r orders_below` orders of magnitude below the noise
floor. The hash sees it regardless, and `verify_manifest()` returns
`r ok_tamper` with both hashes attached so the consumer can name what
diverged. That asymmetry is the whole argument for hashing the payload
rather than checking it by eye.

The three formats are three different promises, and the table above states
them plainly: the default RDS mode and the `both` mode are verifiable, and
`csv_unverified` is not, returning
`r as.character(ok_export)` from `verify_manifest()`
rather than a false negative. The refusal path closes the set. The
off-design surrogate call returned a typed abstention with reason
`r paste0("'", res_off$reason, "'")`, the manifest built from it carries a
zero-row parameter payload and that same reason in its `summary` slot, and
the reason survives the write and the read. The four `consume()` calls at
the end walk exactly those four outcomes: accepted, declined for
unverifiable integrity, declined because the producer abstained, and
declined because the file on disk no longer matches its hash.

## Limits

This vignette shows the record and the check, not the science: the manifest
guarantees that an ensemble is the one that was produced, and says nothing
whatever about whether the ensemble is any good. A hash-verified posterior
from a badly specified likelihood is a hash-verified wrong answer, and the
calibration questions belong to *Wiring your own simulator into PESTO* and
*My posterior looks too confident: is the ensemble collapsing?*. The integrity
check covers the payload the hash is computed over, which is the parameter
ensemble, the outputs, the weights, the targets and the seed; it does not
cover the optional `obs_schema` descriptor, whose correspondence between
column names and physical quantities is established by a second, independent
record that agrees on units rather than by a self-hash. Nor does the hash
make a run reproducible on its own: it records the seed, the method, the
iteration count and the lambda schedule, but re-running a calibration also
needs the forward model, and for a simulator-backed run that means the same
simulator version on the far end. The tamper demonstration alters one cell
deliberately, so it shows detection, not the frequency of accidental
corruption in practice, which this vignette makes no claim about.

## What to read next

*Your first inversion: from a season of measurements to a posterior* is
where the run wrapped above comes from, and is the shortest route into the
smoother itself. *Wiring your own simulator into PESTO* covers the forward-model
contract on the producing side, including the failure handling that sets
the `failure_rate` slot read here. *Can PESTO recover APSIM's own parameters?*
emits a manifest at the end of a full simulator-backed calibration, which
is the realistic version of this handover.
*When can a surrogate stand in for the simulator?* is where the typed
abstention used above comes from, and states the regime that triggers it.

## Reproduce

Seed `20260902` sets the synthetic season, the prior draw and the off-design
surrogate demonstration, and is passed to `pesto_ies_callback()`, which
records it on the run and carries it into the manifest's `seed` slot. All
files are written under `tempfile()` and removed at the end of the
vignette. Package versions follow.

```{r session-info}
sessionInfo()
```
