---
title: "Getting Started with PESTO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PESTO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**PESTO** (**P**arameter **E**stimation, **S**urrogates, and **T**ooling for
**O**ptimisation) is an R package for high-performance parameter estimation,
uncertainty quantification, and inverse modelling. It brings the algorithms of
**PEST** (*Parameter ESTimation*; Doherty 2015) and its C++ successor
**PEST++** (White et al. 2020) natively into the R ecosystem, and extends them
with a typed forward-model contract, an in-process simulator callback,
multi-fidelity acceleration, and surrogate methods.

### Key Features

- **Iterative Ensemble Smoother (IES)**: Ensemble-based parameter
  estimation that scales to 100,000+ parameters without computing
  explicit Jacobians
- **Gauss-Levenberg-Marquardt (GLM)**: Deterministic gradient-based
  estimation with SVD regularisation
- **Global Sensitivity Analysis**: Morris and Sobol methods
- **Parametric Sweep**: Embarrassingly parallel model evaluations
- **Publication-quality diagnostics**: ggplot2-based visualisations

## Installation

```{r install, eval = FALSE}
# From GitHub (development version):
remotes::install_github("max578/PESTO")

# PEST++ binaries must be on your PATH or bundled with the package
```

## Quick Start: Creating a Model Scenario

Rather than requiring a pre-existing PEST control file, PESTO lets you
define problems programmatically:

```{r create-scenario}
library(PESTO)
library(data.table)

# Define parameters
parameters <- data.table(
  parnme    = c("hk1", "hk2", "hk3", "ss1", "rch1"),
  partrans  = c("log", "log", "log", "log", "fixed"),
  parchglim = "factor",
  parval1   = c(10, 5, 1, 1e-4, 1e-3),
  parlbnd   = c(0.1, 0.01, 0.001, 1e-7, 1e-5),
  parubnd   = c(1000, 500, 100, 0.01, 0.1),
  pargp     = c("hk", "hk", "hk", "ss", "rch")
)

# Define observations
observations <- data.table(
  obsnme = paste0("h", 1:10),
  obsval = c(25.3, 24.1, 22.8, 21.5, 20.2,
             19.0, 17.8, 16.5, 15.3, 14.0),
  weight = rep(1.0, 10),
  obgnme = "heads"
)

# Create the scenario
pst <- create_pest_scenario(
  parameters   = parameters,
  observations = observations,
  model_command = "python run_groundwater.py"
)

print(pst)
```

## Core Computational Kernel

PESTO provides the `ensemble_solution()` function -- a high-performance
C++ implementation of the IES update equation. This is the computational
heart of the package:

```{r ensemble-kernel}
set.seed(42)
npar  <- 50
nobs  <- 100
nreal <- 30

# Generate synthetic ensemble differences
par_diff  <- matrix(rnorm(npar * nreal, sd = 0.1), npar, nreal)
obs_diff  <- matrix(rnorm(nobs * nreal, sd = 1.0), nobs, nreal)
obs_resid <- matrix(rnorm(nobs * nreal, sd = 0.5), nobs, nreal)
par_resid <- matrix(rnorm(npar * nreal, sd = 0.1), npar, nreal)
weights   <- rep(1.0, nobs)
parcov_inv <- rep(1.0, npar)
Am <- matrix(rnorm(npar * (nreal - 1)), npar, nreal - 1)

# Run the ensemble update
upgrade <- ensemble_solution(
  par_diff    = par_diff,
  obs_diff    = obs_diff,
  obs_resid   = obs_resid,
  par_resid   = par_resid,
  weights     = weights,
  parcov_inv  = parcov_inv,
  Am          = Am,
  cur_lam     = 1.0
)

cat("Upgrade matrix dimensions:", dim(upgrade), "\n")
cat("Mean absolute upgrade:", mean(abs(upgrade)), "\n")
```

## Performance Benchmarking

The C++ kernel is significantly faster than equivalent R code:

```{r benchmark, eval = TRUE}
# Benchmark the kernel
if (requireNamespace("microbenchmark", quietly = TRUE)) {
  bench <- microbenchmark::microbenchmark(
    PESTO_cpp = ensemble_solution(
      par_diff, obs_diff, obs_resid, par_resid,
      weights, parcov_inv, Am, cur_lam = 1.0
    ),
    times = 100
  )
  print(bench)
}
```

## Computing Phi (Objective Function)

```{r phi}
phi_values <- compute_phi(obs_resid, weights)
cat("Mean phi:", mean(phi_values), "\n")
cat("Min phi:", min(phi_values), "\n")
cat("Max phi:", max(phi_values), "\n")
```

## Visualisation

PESTO provides publication-quality ggplot2-based diagnostics:

```{r plot-phi, fig.cap = "Objective function convergence across iterations."}
# Simulate convergence data
phi_history <- data.table(
  iteration = rep(0:5, each = 30),
  phi = c(
    rlnorm(30, 5, 0.5),
    rlnorm(30, 4, 0.4),
    rlnorm(30, 3.5, 0.3),
    rlnorm(30, 3, 0.25),
    rlnorm(30, 2.8, 0.2),
    rlnorm(30, 2.7, 0.15)
  ),
  real = rep(paste0("r", 1:30), 6)
)

# Reshape wide for plot_phi
phi_wide <- dcast(phi_history, iteration ~ real, value.var = "phi")

plot_phi(phi_wide, show_reals = TRUE, title = "IES Convergence Example")
```

```{r plot-ensemble, fig.cap = "Prior vs posterior parameter distributions."}
# Simulate prior and posterior ensembles
prior <- data.table(
  hk1 = rlnorm(50, log(10), 1),
  hk2 = rlnorm(50, log(5), 1),
  ss1 = rlnorm(50, log(1e-4), 0.5)
)

posterior <- data.table(
  hk1 = rlnorm(50, log(12), 0.3),
  hk2 = rlnorm(50, log(4.5), 0.3),
  ss1 = rlnorm(50, log(8e-5), 0.2)
)

plot_ensemble(posterior, prior_ensemble = prior,
              title = "Prior vs Posterior Parameter Distributions")
```

## Adaptive SVD Backends

PESTO automatically selects the fastest SVD algorithm for your problem.
For low-rank decompositions (typical in IES where ensemble size << observations),
the randomised SVD provides dramatic speedups:

```{r adaptive-svd}
set.seed(42)
A <- matrix(rnorm(1000 * 500), 1000, 500)

# Automatic selection
res_auto <- adaptive_svd(A, k = 20L, method = "auto")
cat("Method:", res_auto$method_used, "\n")
cat("Time:", round(res_auto$time_ms, 2), "ms\n")
cat("Singular values (top 5):", round(res_auto$d[1:5], 3), "\n")

# Compare: randomised SVD (fast, rank-k) vs full
res_rsvd <- rsvd(A, k = 20)
cat("\nrSVD dimensions: U =", dim(res_rsvd$u), ", V =", dim(res_rsvd$v), "\n")
```

## GPU-Accelerated Ensemble Solution

The `ensemble_solution_gpu()` function wraps the IES kernel with
adaptive SVD backend selection and returns performance diagnostics:

```{r gpu-solution}
set.seed(42)
npar <- 100; nobs <- 500; nreal <- 50

pd <- matrix(rnorm(npar * nreal), npar, nreal)
od <- matrix(rnorm(nobs * nreal), nobs, nreal)
or_ <- matrix(rnorm(nobs * nreal), nobs, nreal)
pr <- matrix(rnorm(npar * nreal), npar, nreal)
w <- rep(1.0, nobs); pc <- rep(1.0, npar)
Am <- matrix(rnorm(npar * (nreal - 1)), npar, nreal - 1)

result <- ensemble_solution_gpu(
  pd, od, or_, pr, w, pc, Am,
  cur_lam = 1.0, svd_method = "auto"
)

cat("SVD method:", result$svd_method, "\n")
cat("SVD time:", round(result$svd_time_ms, 2), "ms\n")
cat("Total time:", round(result$total_time_ms, 2), "ms\n")
cat("Singular values used:", result$singular_values_used, "\n")
```

## Surrogate-Accelerated IES

PESTO includes a Gaussian Process surrogate that can replace expensive
model evaluations during IES iterations. See `vignette("surrogate-ies")`
for full details.

```{r surrogate-quick}
set.seed(42)
par_ens <- matrix(rnorm(50 * 10), 50, 10)
obs_ens <- par_ens %*% matrix(rnorm(10 * 20), 10, 20) +
  matrix(rnorm(50 * 20, sd = 0.1), 50, 20)
obs_target <- rnorm(20)

result <- surrogate_ensemble_update(
  par_ens, obs_ens, obs_target,
  weights = rep(1, 20), parcov_inv = rep(1, 10),
  uncertainty_threshold = 0.1
)

cat("Model evaluations needed:", result$n_model_runs, "\n")
cat("Surrogate evaluations:", result$n_surrogate_runs, "\n")
cat("Savings:", sprintf("%.0f%%", result$savings_pct), "\n")
```

## Adaptive Ensemble Sizing

Monitor ensemble health and get sizing recommendations:

```{r adaptive-sizing}
phi_values <- rlnorm(50, 3, 0.5)
sizing <- adaptive_ensemble_size(phi_values, current_size = 50L)
cat("Recommended size:", sizing$recommended_size, "\n")
cat("CV(phi):", round(sizing$cv_phi, 3), "\n")
cat("ESS:", round(sizing$ess, 1), "/", sizing$current_size, "\n")
cat("Reason:", sizing$reasoning, "\n")
```

## Next Steps

- See `vignette("surrogate-ies")` for the full surrogate-accelerated workflow
- See `?pesto_ies` for the high-level IES interface (wraps pestpp-ies)
- See `?pesto_glm` for deterministic Gauss-Levenberg-Marquardt estimation

## References

- Doherty, J. (2015). *PEST: Model-Independent Parameter Estimation --
  User Manual* (6th ed.). Watermark Numerical Computing, Brisbane.
- White, J.T., Hunt, R.J., Fienen, M.N., & Doherty, J.E. (2020).
  Approaches to Highly Parameterized Inversion: PEST++ Version 5.
  USGS Techniques and Methods 7-C26.
- Chen, Y., & Oliver, D.S. (2013). Levenberg-Marquardt forms of
  the iterative ensemble smoother for efficient history matching
  and uncertainty quantification. *Computational Geosciences*, 17(4).
- Evensen, G. (2018). Analysis of iterative ensemble smoothers for
  solving inverse problems. *Computational Geosciences*, 22(3).
