---
title: "Your first inversion: from a season of measurements to a posterior"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Your first inversion: from a season of measurements to a posterior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~4 s under rmarkdown::render(); macOS arm64 (Darwin 26.5),
     R 4.5.2, one core. The inversion itself is under a second; most of the
     time is the kernel timing loop and the two large SVD demonstrations. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(PESTO)
```

## Why

I have a season of measured wheat biomass, a growth model with three
parameters I cannot measure in the field, and a supervisor who wants error
bars rather than a single fitted curve. The question I actually need
answered is: *given these measurements and this model, what is the range of
parameter values consistent with what I saw, and which of those parameters
do the data pin down at all?* This vignette answers it end to end on a
synthetic season, in about a second of compute, with no external binary and
no simulator installation.

## What

Four verbs carry the whole workflow.
[`crop_growth_forward_model()`](../html/crop_growth_forward_model.html)
builds the forward model: the logistic dry-matter curve
$\mathrm{d}B/\mathrm{d}t = r B (1 - B / b_{\max})$ integrated on a time grid
and returned as a typed
[`pesto_forward_model()`](../html/pesto_forward_model.html), which is the
contract every PESTO driver reads. Any R function mapping a parameter matrix
to an output matrix can take its place, which is what *model-independent*
means here.
[`pesto_evaluate()`](../html/pesto_evaluate.html) runs that contract over a
matrix of parameter realisations and returns the matching matrix of model
outputs, carrying a per-evaluation failure count.
[`pesto_ies_callback()`](../html/pesto_ies_callback.html) is the inversion:
an iterative ensemble smoother after Chen and Oliver (2013) that starts from
a prior ensemble, calls the forward model in process, and returns a
posterior ensemble rather than a point estimate.
[`as_manifest()`](../html/as_manifest.html) wraps the finished run as a
hash-verified `pesto_ensemble_manifest`, the object PESTO owns as the
orchestra's C2 manifest contract.

Beneath those sit the pieces you can reach for directly:
[`ensemble_solution()`](../html/ensemble_solution.html), the C++ kernel that
computes one smoother update;
[`compute_phi()`](../html/compute_phi.html), the weighted objective
function; [`adaptive_svd()`](../html/adaptive_svd.html) and
[`rsvd()`](../html/rsvd.html), the decomposition backends the kernel selects
between; [`adaptive_ensemble_size()`](../html/adaptive_ensemble_size.html),
which reads ensemble health and recommends a size; and
[`create_pest_scenario()`](../html/create_pest_scenario.html), which
describes a problem in PEST's own control-file vocabulary so it travels to
the wider PEST toolchain.

PESTO (Parameter ESTimation Optimised) brings the algorithms of PEST
(Doherty 2015) and its C++ successor PEST++ (White et al. 2020) natively
into R, and extends them with the typed forward-model contract used above,
an in-process simulator callback, multi-fidelity acceleration, and surrogate
methods. Its flagship simulator partner is APSIM, coupled through
[`apsim_callback()`](../html/apsim_callback.html) and the `apsimx` package.

## Do

### Installation, and what needs an external binary

PESTO installs from its GitHub repository, `max578/PESTO`. It ships no PEST++
binaries: every algorithm below runs natively in R, and an external binary is
needed only for the optional cross-check and the `.pst`-file path through
[`pesto_ies()`](../html/pesto_ies.html). When one is wanted,
`.find_pestpp_exe()` resolves it in four steps: the per-tool environment
variable (`PESTPP_IES_EXE_PATH` and its siblings), then `PESTPP_BIN_DIR`
holding the PEST++ suite, then the system `PATH`, then the `exe` argument
passed to the calling function.
[`pestpp_available()`](../html/pestpp_available.html) probes that same order
without erroring, and is the documented way for an example or a test to skip
gracefully. Nothing in this vignette depends on the answer:

```{r binary-probe}
pestpp_available("pestpp-ies")
```

```{r binary-note, echo = FALSE, results = "asis"}
if (isTRUE(pestpp_available("pestpp-ies"))) {
  cat("A `pestpp-ies` binary is installed on this machine, so the optional",
      "`.pst` path would run; this vignette does not use it.\n")
} else {
  cat("No `pestpp-ies` binary is installed on this machine, so the optional",
      "`.pst` path is unavailable here; every result below is computed by",
      "PESTO's own R and C++ code.\n")
}
```

### A synthetic season

The season below is synthetic: a logistic biomass curve at a known growth
rate, ceiling and starting biomass, read off every ten days from day 10 to
day 120 and blurred with Gaussian field noise of 45 g/m2. Working from a
known truth is what lets the last section ask what the posterior recovered. Real
measurements enter the same way, as a named numeric vector of targets and a
standard deviation.

```{r season}
set.seed(20260902L)
vec_times <- seq(0, 120, by = 10)
crop <- crop_growth_forward_model(times = vec_times)
# The template integrates from times[1] and reports the trajectory at every
# later time, so the observations sit on days 10 to 120.
vec_days <- vec_times[-1L]

theta_true <- c(r = 0.065, b_max = 1450, b0 = 18)
mat_true <- matrix(theta_true, nrow = 1L,
                   dimnames = list(NULL, names(theta_true)))
vec_biomass <- as.numeric(pesto_evaluate(crop, mat_true))

obs_sd <- 45
y_obs <- vec_biomass + rnorm(length(vec_biomass), sd = obs_sd)
names(y_obs) <- paste0("t", seq_along(y_obs))
round(y_obs)
```

### A prior, and the inversion

The prior is deliberately vague: uniform over a growth rate of 0.02 to 0.12
per day, a ceiling of 900 to 2000 g/m2, and a starting biomass of 5 to
60 g/m2. One hundred and twenty realisations are drawn from it, and eight
smoother iterations condition them on the twelve measurements.

```{r inversion}
s1_real <- 120L
prior <- cbind(
  r     = runif(s1_real, 0.02, 0.12),
  b_max = runif(s1_real, 900, 2000),
  b0    = runif(s1_real, 5, 60)
)

fit <- pesto_ies_callback(
  forward_model  = crop,
  prior_ensemble = prior,
  obs            = y_obs,
  obs_sd         = obs_sd,
  noptmax        = 8L,
  seed           = 20260902L,
  verbose        = FALSE
)

tab_post <- as.matrix(fit$par_ensemble[, -1L])
round(colMeans(tab_post), 3)
```

```{r inversion-facts, echo = FALSE}
vec_prior_sd <- apply(prior, 2L, sd)
vec_post_sd <- apply(tab_post, 2L, sd)
vec_ratio <- vec_post_sd / vec_prior_sd
s1_it <- length(fit$iterations)
vec_mean_phi_pre <- vapply(fit$iterations, `[[`, numeric(1L), "mean_phi")
phi_drop_pct <- 100 * (vec_mean_phi_pre[1L] - vec_mean_phi_pre[4L]) /
  (vec_mean_phi_pre[1L] - vec_mean_phi_pre[s1_it])
post_pred_sd <- mean(apply(as.matrix(fit$obs_ensemble[, -1L]), 2L, sd,
                           na.rm = TRUE))
tab_recovery <- data.frame(
  Parameter = colnames(tab_post),
  Unit = c("per day", "g/m2", "g/m2"),
  Truth = theta_true[colnames(tab_post)],
  `Posterior mean` = colMeans(tab_post),
  `Posterior SD` = vec_post_sd,
  `Posterior SD / prior SD` = vec_ratio,
  check.names = FALSE, row.names = NULL
)
knitr::kable(
  tab_recovery, digits = 3,
  caption = paste0(
    "Recovery of the three growth parameters from ", length(y_obs),
    " noisy biomass measurements, over ", s1_real, " realisations."
  )
)
```

Not every realisation survives. The smoother is unbounded, so a member can
be pushed to a parameter value the model cannot integrate; PESTO records
those as failures rather than dropping them silently, and the default
`on_failure = "na"` carries them forward so the rest of the ensemble is not
thrown away.

```{r failures}
fit$failure_rate
range(tab_post[, "b0"])
```

### Did the misfit fall?

The objective function $\phi$ is the weighted sum of squared residuals; the
smoother should drive its ensemble mean down and then flatten.
[`plot_phi()`](../html/plot_phi.html) draws that trajectory. It reads a wide
table of one column per realisation, which the run's long-format `$phi` slot
casts to directly.

```{r phi-figure, fig.cap = "Objective function by smoother iteration: the ribbon spans the ensemble's minimum to maximum misfit, the line its mean, and each faint line is one realisation. Almost all of the reduction happens in the first four iterations; the tail is the measurement noise the model cannot fit away."}
tab_phi_wide <- data.table::dcast(
  fit$phi, iteration ~ realisation, value.var = "phi"
)
# The realisations that failed carry no misfit; they are dropped here so the
# trajectory plot shows only the members that completed every iteration.
vec_complete <- !vapply(tab_phi_wide, anyNA, logical(1L))
plot_phi(tab_phi_wide[, which(vec_complete), with = FALSE],
         show_reals = TRUE,
         title = "Misfit across the ensemble, by iteration")
```

```{r phi-numbers}
vec_mean_phi <- vapply(fit$iterations, `[[`, numeric(1L), "mean_phi")
round(vec_mean_phi, 1)
```

### What the data did to the prior

```{r ensemble-figure, fig.cap = "Prior and posterior parameter ensembles as violins, on each parameter's own scale. The ceiling and the growth rate contract sharply; the starting biomass barely moves, because a curve sampled from day zero onward carries little information about where it began."}
plot_ensemble(
  fit$par_ensemble[, -1L],
  prior_ensemble = as.data.frame(prior),
  title = "Prior against posterior, by parameter"
)
```

The violins compare parameters one at a time. The figure that answers the
original question is the one in the space where the measurements live: the
biomass trajectories the prior ensemble predicted, the trajectories the
posterior ensemble predicts, and the measurements themselves on the same
axes.

```{r fan-figure, fig.height = 5, fig.cap = "Prior and posterior ensemble fans in observation space. Each faint line is one realisation's biomass trajectory; the points are the synthetic measurements the smoother conditioned on. The prior fan spans the whole plausible envelope; the posterior fan closes onto the measurements and keeps a spread consistent with the 45 g/m2 field noise."}
tab_prior_obs <- pesto_evaluate(crop, prior)
tab_fan <- data.table::rbindlist(list(
  data.table::data.table(
    day = rep(vec_days, each = s1_real),
    biomass = as.numeric(tab_prior_obs),
    real = rep(seq_len(s1_real), times = length(vec_days)),
    ensemble = "prior"
  ),
  data.table::data.table(
    day = rep(vec_days, each = s1_real),
    biomass = as.numeric(as.matrix(fit$obs_ensemble[, -1L])),
    real = rep(seq_len(s1_real), times = length(vec_days)),
    ensemble = "posterior"
  )
))
tab_points <- data.table::data.table(day = vec_days, biomass = y_obs)

ggplot2::ggplot(
  tab_fan,
  ggplot2::aes(x = day, y = biomass, group = interaction(real, ensemble),
               colour = ensemble)
) +
  ggplot2::geom_line(alpha = 0.12, linewidth = 0.3, na.rm = TRUE) +
  ggplot2::geom_point(
    data = tab_points, ggplot2::aes(x = day, y = biomass),
    inherit.aes = FALSE, colour = "black", size = 1.8
  ) +
  ggplot2::scale_colour_viridis_d(name = "ensemble", end = 0.8) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(override.aes = list(alpha = 1,
                                                       linewidth = 1))
  ) +
  ggplot2::labs(
    x = "day of season", y = "above-ground biomass (g/m2)",
    title = "Prior and posterior ensembles against the measurements"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

### The run as a portable record

A finished run wraps into a versioned, hash-verified manifest, which is how
a posterior leaves PESTO for another tool without that tool reaching into
PESTO's internals.

```{r manifest}
man <- as_manifest(fit, seed = 20260902L)
man@method
man@schema_version
verify_manifest(man)$ok
```

### Under the bonnet

The four verbs above are enough to invert a model. The pieces below are what
they are built from, and each is exported so it can be used on its own.

**The update kernel.** One smoother iteration is a single call to
`ensemble_solution()`, a C++ implementation of the Chen and Oliver update
that takes ensemble anomalies and residuals and returns the parameter
upgrade. The block below recomputes the very first update by hand, from the
prior ensemble and the model outputs already evaluated for the fan figure.

```{r kernel}
tab_par_diff <- t(prior) - colMeans(prior)
tab_obs_diff <- t(tab_prior_obs) - colMeans(tab_prior_obs)
tab_obs_resid <- t(tab_prior_obs) - matrix(
  rep(y_obs, s1_real), length(y_obs), s1_real
)
vec_weights <- rep(1 / obs_sd, length(y_obs))
vec_parcov_inv <- 1 / apply(prior, 2L, var)
mat_am <- matrix(rnorm(ncol(prior) * (s1_real - 1L)),
                 ncol(prior), s1_real - 1L)

tab_upgrade <- ensemble_solution(
  par_diff   = tab_par_diff,
  obs_diff   = tab_obs_diff,
  obs_resid  = tab_obs_resid,
  par_resid  = tab_par_diff,
  weights    = vec_weights,
  parcov_inv = vec_parcov_inv,
  Am         = mat_am,
  cur_lam    = 1.0
)
dim(tab_upgrade)
```

**The objective.** `compute_phi()` reduces the residual matrix to one
weighted misfit per realisation, which is what the ribbon in the first
figure summarises:

```{r phi-kernel}
vec_phi_prior <- compute_phi(tab_obs_resid, vec_weights)
round(c(mean = mean(vec_phi_prior), min = min(vec_phi_prior),
        max = max(vec_phi_prior)), 1)
```

The kernel is in C++ because it runs once per iteration on matrices that
grow with the ensemble. The chunk below reports its average cost on this
problem, in milliseconds over one hundred repetitions.

```{r benchmark}
s1_rep <- 100L
elapsed_ms <- 1000 * system.time(
  for (i1 in seq_len(s1_rep)) {
    ensemble_solution(
      tab_par_diff, tab_obs_diff, tab_obs_resid, tab_par_diff,
      vec_weights, vec_parcov_inv, mat_am, cur_lam = 1.0
    )
  } # ends i1, over the timing repetitions
)[["elapsed"]] / s1_rep
round(elapsed_ms, 3)
```

**The decomposition backend.** The update needs a singular value
decomposition, and the fastest algorithm depends on the shape of the
problem. `adaptive_svd()` picks one; `rsvd()` is the randomised
rank-$k$ route it prefers when the ensemble is much smaller than the
observation count, which is the usual smoother regime. The synthetic matrix
below is deliberately larger than this vignette's problem, because the
choice only starts to matter at that size.

```{r svd}
set.seed(20260902L)
mat_a <- matrix(rnorm(1000L * 500L), 1000L, 500L)
res_svd <- adaptive_svd(mat_a, k = 20L, method = "auto")
res_svd$method_used
round(res_svd$time_ms, 1)
round(res_svd$d[seq_len(5L)], 2)
```

`ensemble_solution_adaptive()` is the same kernel with that selection made
for you, returning the timing breakdown alongside the upgrade:

```{r adaptive-solution}
res_adaptive <- ensemble_solution_adaptive(
  tab_par_diff, tab_obs_diff, tab_obs_resid, tab_par_diff,
  vec_weights, vec_parcov_inv, mat_am,
  cur_lam = 1.0, svd_method = "auto"
)
res_adaptive$svd_method
res_adaptive$singular_values_used
```

**Is the ensemble big enough?** `adaptive_ensemble_size()` reads the spread
of final-iteration misfits and recommends a size. A high coefficient of
variation across realisations means a few members carry the fit and the rest
are along for the ride:

```{r sizing}
vec_phi_final <- fit$phi$phi[fit$phi$iteration == max(fit$phi$iteration)]
sizing <- adaptive_ensemble_size(
  vec_phi_final[!is.na(vec_phi_final)], current_size = s1_real
)
sizing$recommended_size
round(sizing$cv_phi, 3)
sizing$reasoning
```

**A first look at the surrogate.** When the forward model is expensive, a
Gaussian process trained on the ensemble can stand in for some of its
evaluations. `surrogate_ensemble_update()` classifies which realisations the
surrogate could cover and returns the update; the *Surrogate-accelerated
iterative ensemble smoother* vignette is the full treatment, including what
the reported saving does and does not mean.

```{r surrogate-quick}
res_surrogate <- surrogate_ensemble_update(
  par_ensemble = prior,
  obs_ensemble = tab_prior_obs,
  obs_target = y_obs,
  weights = vec_weights,
  parcov_inv = vec_parcov_inv,
  uncertainty_threshold = 0.1
)
c(model = res_surrogate$n_model_runs,
  surrogate = res_surrogate$n_surrogate_runs)
```

**Describing a problem in PEST's vocabulary.** Nothing above needed a PEST
control file. When one is wanted, because the forward model lives behind a
non-R executable or because the problem has to travel to another PEST tool,
`create_pest_scenario()` builds it programmatically from a parameter table
and an observation table:

```{r scenario}
tab_parameters <- data.table::data.table(
  parnme    = c("r", "b_max", "b0"),
  partrans  = c("log", "log", "log"),
  parchglim = "factor",
  parval1   = c(0.065, 1450, 18),
  parlbnd   = c(0.02, 900, 5),
  parubnd   = c(0.12, 2000, 60),
  pargp     = c("growth", "growth", "initial")
)
tab_observations <- data.table::data.table(
  obsnme = names(y_obs),
  obsval = as.numeric(y_obs),
  weight = vec_weights,
  obgnme = "biomass"
)
pst <- create_pest_scenario(
  parameters    = tab_parameters,
  observations  = tab_observations,
  model_command = "Rscript run_crop.R"
)
pst$control_data$npar
pst$control_data$nobs
```

## Read

The inversion recovers the season it was given. The growth rate comes back
at `r round(mean(tab_post[, "r"]), 3)` per day against a truth of
`r theta_true[["r"]]`, the ceiling at
`r round(mean(tab_post[, "b_max"]))` g/m2 against `r theta_true[["b_max"]]`,
and the starting biomass at `r round(mean(tab_post[, "b0"]), 1)` g/m2
against `r theta_true[["b0"]]`. The mean misfit falls from
`r round(vec_mean_phi[1])` at the first iteration to
`r round(vec_mean_phi[s1_it])` at the eighth, with
`r round(phi_drop_pct)` per cent of that whole reduction achieved by the
fourth iteration; the rest is the tail visible in the first figure.

The second half of the question was which parameters the data pin down, and
the answer is in the ratio column of the recovery table. Conditioning cuts
the ceiling's spread to `r round(100 * vec_ratio[["b_max"]])` per cent of its
prior spread and the growth rate's to
`r round(100 * vec_ratio[["r"]])` per cent, while the starting biomass keeps
`r round(100 * vec_ratio[["b0"]])` per cent of its prior spread. That
ordering is mechanistically what a season of biomass measurements should
teach: the ceiling is read almost directly off the plateau, the growth rate
off the slope through the middle of the season, and the starting biomass
only weakly off a first measurement taken on day `r vec_days[1]`, by which
point the crop already carries `r round(vec_biomass[1])` g/m2. The
posterior fan makes the same point without reference to any parameter: it
closes onto the points and
keeps a spread of `r round(post_pred_sd)` g/m2 averaged across the twelve
measurement days. That is the spread of modelled trajectories alone, so it
should and does sit below the `r obs_sd` g/m2 of measurement noise: the
ensemble has narrowed to the band the data support without narrowing onto a
single curve.

The run is not perfectly clean, and the diagnostics say so.
`r round(100 * fit$failure_rate, 1)` per cent of forward evaluations
returned nothing: a small number of realisations were pushed to a starting
biomass below zero, where the growth model has no trajectory to return, and
the lowest posterior value of `b0` is
`r round(min(tab_post[, "b0"]), 1)` g/m2. Those realisations appear as gaps
in the posterior fan rather than as a silently shortened ensemble, which is
the behaviour to want: a failure that is counted can be reasoned about, and
`on_failure = "stop"` is available when it should abort the run instead.

`adaptive_ensemble_size()` recommends `r sizing$recommended_size`
realisations against the `r s1_real` used here, because the coefficient of
variation of the final misfit is `r round(sizing$cv_phi, 2)`: a few
realisations still fit markedly better than the rest, and more of them would
even that out. One update of the C++ kernel costs
`r round(elapsed_ms, 2)` milliseconds on this problem, with the
decomposition backend chosen automatically as `r res_svd$method_used`; both
sit far below the cost of a real forward model, which is the regime the
whole design assumes.

## Limits

This season is synthetic, and that is what makes the recovery check
possible: the truth used to generate the measurements is the truth the table
compares against, so nothing here establishes that the logistic curve is the
right model for a real crop. A real inversion is only as good as the forward
model behind it, and a structurally wrong model will produce a tight
posterior around the wrong answer without complaint. The noise model is
equally clean: Gaussian, independent, and known, whereas field measurements
carry correlated error across dates and a discrepancy between the model and
the system that this vignette does not represent. The single most
consequential number is `obs_sd`, and passing one that is too small collapses
the posterior into false confidence; that failure and its diagnosis are
worked through in *Wiring your own simulator into PESTO*. Coverage is not tested
here either, only the ratio of posterior to prior spread, and a spread that
looks reasonable is not the same as an interval that contains the truth at
its nominal rate. Nothing in this vignette exercises an external simulator or
an external binary, so it says nothing about how PESTO behaves when a single
forward evaluation takes a minute rather than a microsecond.

## What to read next

*Wiring your own simulator into PESTO* is the next step for anyone with a
real simulator: the in-process callback, the typed forward-model contract,
the parallel and fault-tolerant evaluation path, and the guard on `obs_sd`
that decides whether the posterior above is credible. *My posterior looks
too confident: is the ensemble collapsing?* takes up what to do when the
posterior spread comes back too narrow for the right reason rather than the
wrong one. *When can a surrogate stand in for the simulator?* is where to go
when the forward model is too expensive to run once per realisation per
iteration. *Handing an ensemble to somebody else* documents the record
`as_manifest()` produced above and what a downstream tool is entitled to
assume about it. *Can PESTO recover APSIM's own parameters?* runs the whole
workflow against a real simulator and real measured biomass, and *Is PESTO
the same algorithm as the tools it descends from?* places the smoother
beside the tools it came out of.

## References

- Chen, Y., & Oliver, D. S. (2013). Levenberg-Marquardt forms of the
  iterative ensemble smoother for efficient history matching and uncertainty
  quantification. *Computational Geosciences*, 17(4), 689-703.
- Doherty, J. (2015). *Calibration and Uncertainty Analysis for Complex
  Environmental Models*. Watermark Numerical Computing, Brisbane.
- Evensen, G. (2018). Analysis of iterative ensemble smoothers for solving
  inverse problems. *Computational Geosciences*, 22(3), 885-908.
- White, J. T., Hunt, R. J., Fienen, M. N., & Doherty, J. E. (2020).
  *Approaches to Highly Parameterized Inversion: PEST++ Version 5*. U.S.
  Geological Survey Techniques and Methods 7-C26.

## Reproduce

Seed `20260902` sets the synthetic season, the prior draw, and the
decomposition demonstration; the same seed is passed to
`pesto_ies_callback()`, which uses it for the per-realisation observation
perturbation, and is recorded on the manifest. Package versions follow.

```{r session-info}
sessionInfo()
```
