---
title: "Countering Ensemble Collapse: Inflation and Localisation"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Countering Ensemble Collapse: Inflation and Localisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## The problem: finite-ensemble under-dispersion

An iterative ensemble smoother estimates a posterior parameter
distribution from a finite ensemble of realisations. With a finite
ensemble two pathologies arise, both of which make the posterior spread
*too narrow* -- the ensemble becomes over-confident:

1. **Under-dispersion (ensemble collapse).** Each assimilation step
   contracts the ensemble spread; over several iterations the posterior
   variance can fall far below the true posterior variance. The smoother
   then reports far more certainty than the data support.
2. **Spurious correlations.** A finite ensemble manufactures apparent
   correlations between parameters and observations that are not real.
   Acting on them injects noise into the update and accelerates collapse.

PESTO addresses the first with **covariance inflation**
([`pesto_inflation()`](../html/pesto_inflation.html)) and the second with
**covariance localisation**
([`pesto_localisation()`](../html/pesto_localisation.html)). Both are
opt-in: the default `NULL` leaves the update identical to the bare
smoother.

This vignette demonstrates the effect on a linear-Gaussian problem, where
the analytic posterior is known and the collapse can be measured exactly.

## A linear-Gaussian problem with a known posterior

For a linear forward model $d = G\theta + \varepsilon$ with a Gaussian
prior $\theta \sim N(0, C_0)$ and observation error
$\varepsilon \sim N(0, R)$, the posterior covariance is
$C_{\mathrm{post}} = (C_0^{-1} + G^{\top} R^{-1} G)^{-1}$. We can therefore
compare the ensemble's posterior spread directly against the truth.

```{r problem}
library(PESTO)
set.seed(42L)

npar  <- 6L
nobs  <- 10L
nreal <- 24L          # a deliberately small ensemble, to provoke collapse

G          <- matrix(rnorm(nobs * npar), nobs, npar)
theta_true <- rnorm(npar)
obs_sd     <- 0.3
y          <- as.numeric(G %*% theta_true) + rnorm(nobs, sd = obs_sd)

# Analytic posterior standard deviation (standard-normal prior).
post_cov <- solve(diag(npar) + crossprod(G) / obs_sd^2)
post_sd  <- sqrt(diag(post_cov))

forward <- function(theta) theta %*% t(G)
prior   <- matrix(rnorm(nreal * npar), nreal, npar,
                  dimnames = list(NULL, paste0("p", seq_len(npar))))
```

A small helper runs the smoother and returns the realised posterior
spread alongside the spread-ESS collapse diagnostic recorded on the final
iteration.

```{r helper}
run_ies <- function(inflation = NULL, localisation = NULL) {
  fit <- pesto_ies_callback(
    forward_model  = forward,
    prior_ensemble = prior,
    obs            = setNames(y, paste0("o", seq_len(nobs))),
    obs_sd         = obs_sd,
    noptmax        = 12L,
    inflation      = inflation,
    localisation   = localisation,
    verbose        = FALSE
  )
  par_post  <- as.matrix(fit$par_ensemble[, -1])
  last_diag <- fit$iterations[[length(fit$iterations)]]
  list(
    sd_ratio  = mean(apply(par_post, 2L, sd) / post_sd),
    ess_ratio = last_diag$spread_ess_ratio
  )
}
```

## The bare smoother collapses

```{r bare}
bare <- run_ies()
round(bare$sd_ratio, 3)
```

The mean posterior standard deviation is only a fraction of the analytic
value -- the ensemble is badly over-confident. The spread-ESS ratio
quantifies the same collapse from the eigenspectrum of the parameter
anomaly covariance (1 means variance is spread isotropically across all
directions; small values mean it has collapsed onto a few):

```{r bare-ess}
round(bare$ess_ratio, 3)
```

## Inflation re-expands the spread

`pesto_inflation()` offers four methods. The workhorse is relaxation to
prior spread (`"rtps"`, Whitaker & Hamill 2012): each parameter's
posterior anomalies are rescaled toward the pre-update spread, so the
directions that collapsed hardest are re-inflated most. The `"adaptive"`
method instead targets a global spread-retention floor.

```{r inflation}
rtps     <- run_ies(inflation = pesto_inflation("rtps", alpha = 0.6))
adaptive <- run_ies(inflation = pesto_inflation("adaptive",
                                                retention_floor = 0.7))

data.frame(
  method        = c("none", "rtps", "adaptive"),
  sd_ratio      = round(c(bare$sd_ratio, rtps$sd_ratio, adaptive$sd_ratio), 3),
  spread_ess    = round(c(bare$ess_ratio, rtps$ess_ratio, adaptive$ess_ratio), 3)
)
```

RTPS roughly doubles the retained posterior spread and lifts the
spread-ESS ratio. A caveat worth stating plainly: inflation *mitigates*
collapse, it does not abolish it. A finite-ensemble GLM smoother of this
size still under-estimates the posterior spread; inflation moves it
substantially closer to the truth without claiming to reach it. Larger
ensembles narrow the residual gap.

The spread-ESS diagnostic ([`ensemble_spread_ess()`](../html/ensemble_spread_ess.html))
is recorded on **every** iteration regardless of method, so the collapse
trajectory is always available in the result:

```{r ess-trace}
fit <- pesto_ies_callback(
  forward, prior, setNames(y, paste0("o", seq_len(nobs))),
  obs_sd = obs_sd, noptmax = 12L,
  inflation = pesto_inflation("rtps", alpha = 0.6), verbose = FALSE
)
plot(
  vapply(fit$iterations, function(d) d$spread_ess_ratio, numeric(1L)),
  type = "b", pch = 19, xlab = "iteration", ylab = "spread-ESS ratio",
  main = "Dispersion held up under RTPS inflation", ylim = c(0, 1)
)
```

## Localisation suppresses spurious correlations

For parameter-estimation problems whose parameters carry no spatial
coordinate, the recommended localiser is the correlation-based automatic
method (Luo & Bhakta 2020). It needs no metric: it estimates a noise
floor from the ensemble itself and damps sample correlations that fall
below it.

```{r localisation}
loc <- run_ies(localisation = pesto_localisation("correlation",
                                                 taper = "soft"))
round(loc$sd_ratio, 3)
```

The two countermeasures compose -- inflation restores variance magnitude,
localisation removes the spurious updates that drain it:

```{r both}
both <- run_ies(
  inflation    = pesto_inflation("rtps", alpha = 0.6),
  localisation = pesto_localisation("correlation", taper = "soft")
)
round(both$sd_ratio, 3)
```

When a genuine distance metric *does* exist, the classical Gaspari-Cohn
taper is available via `pesto_localisation("distance", ...)`, supplying
either a precomputed `distances` matrix or `par_coords` / `obs_coords`
together with a localisation `radius`.

## References

Gaspari, G. & Cohn, S. E. (1999). Construction of correlation functions in
two and three dimensions. *Quarterly Journal of the Royal Meteorological
Society*, 125(554), 723--757.

Luo, X. & Bhakta, T. (2020). Automatic and adaptive localization for
ensemble-based history matching. *Journal of Petroleum Science and
Engineering*, 184, 106559.

Whitaker, J. S. & Hamill, T. M. (2012). Evaluating methods to account for
system errors in ensemble data assimilation. *Monthly Weather Review*,
140(9), 3078--3089.

## Session information

```{r sessioninfo}
sessionInfo()
```
