---
title: "My posterior looks too confident: is the ensemble collapsing?"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{My posterior looks too confident: is the ensemble collapsing?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64 (Darwin 26.5),
     R 4.5.2, one core. The ensemble-size sweep dominates, because its
     largest run carries 500 realisations. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(PESTO)
```

## Why

My calibration came back with credible intervals I do not believe: the
parameters are pinned far more tightly than the data should allow, and if I
hand those intervals to an agronomist they will price a decision on
confidence that is not there. The question is: *is my ensemble collapsing,
and if it is, will covariance inflation put the spread back?* This vignette
answers both on a linear-Gaussian problem whose posterior is known in closed
form, so the ensemble's spread can be compared against the right answer
rather than against an impression.

## What

Two pathologies make a finite ensemble over-confident, and PESTO has one
countermeasure for each.

The first is **finite-ensemble collapse**. Each assimilation step contracts
the ensemble spread, and over several iterations the posterior variance can
fall below the true posterior variance. This is the classical effect the
ensemble-Kalman literature addresses (Whitaker and Hamill 2012; Luo and
Bhakta 2020), and it shrinks as the ensemble grows.
[`pesto_inflation()`](../html/pesto_inflation.html) is the answer: it
declares a covariance-inflation scheme, of which the workhorse is relaxation
to prior spread (`"rtps"`), rescaling each parameter's posterior anomalies
toward the pre-update spread so the directions that collapsed hardest are
re-inflated most. An `"adaptive"` method instead targets a global
spread-retention floor.

The second is **spurious correlation**. A finite ensemble manufactures
apparent correlations between parameters and observations that are not
there, and acting on them injects noise into the update and accelerates
collapse. [`pesto_localisation()`](../html/pesto_localisation.html) declares
a localisation scheme that damps them. For parameters carrying no spatial
coordinate the recommended localiser is the correlation-based automatic
method of Luo and Bhakta (2020), which needs no metric: it estimates a noise
floor from the ensemble itself and damps sample correlations below it. Where
a genuine distance metric does exist, the classical Gaspari-Cohn taper
(Gaspari and Cohn 1999) is available through
[`gaspari_cohn()`](../html/gaspari_cohn.html) and
`pesto_localisation("distance", ...)`.

Both are optional arguments to
[`pesto_ies_callback()`](../html/pesto_ies_callback.html), and the default
`NULL` leaves the update identical to the bare smoother.
[`ensemble_spread_ess()`](../html/ensemble_spread_ess.html) is the
diagnostic recorded on every iteration whether or not either is switched on:
the spectral participation ratio of the parameter anomaly covariance, which
is the effective number of variance-carrying directions.

A third pathology used to sit alongside these two and was by far the
largest: assimilating the same, unperturbed observation vector into every
realisation, which gives every member an identical data pull so that the
between-member spread stops being a posterior spread at all. Unlike
finite-ensemble collapse, it does not shrink as the ensemble grows. PESTO's
smoother now perturbs the observations per realisation under an
ensemble-smoother-with-multiple-data-assimilation schedule (Emerick and
Reynolds 2013; see *Posterior spread and observation perturbation* in
[`?pesto_ies_callback`](../html/pesto_ies_callback.html)), so what remains on
this problem is the classical, ensemble-size-dependent collapse that
inflation and localisation were designed for. The ensemble-size sweep below
is the evidence for that.

## Do

### A problem whose posterior is known

For a linear forward model $d = G\theta + \varepsilon$ with a Gaussian
prior $\theta \sim N(0, C_0)$ and observation error
$\varepsilon \sim N(0, R)$, the posterior covariance is
$C_{\mathrm{post}} = (C_0^{-1} + G^{\top} R^{-1} G)^{-1}$. The ensemble's
realised spread can therefore be compared against the exact answer. The
problem below is synthetic and deliberately small in the ensemble, to
provoke the collapse this vignette is about.

```{r problem}
set.seed(42L)

npar <- 6L
nobs <- 10L
nreal <- 24L          # a deliberately small ensemble, to provoke collapse

G <- matrix(rnorm(nobs * npar), nobs, npar)
theta_true <- rnorm(npar)
obs_sd <- 0.3
y <- as.numeric(G %*% theta_true) + rnorm(nobs, sd = obs_sd)

# Analytic posterior standard deviation (standard-normal prior).
post_cov <- solve(diag(npar) + crossprod(G) / obs_sd^2)
post_sd <- sqrt(diag(post_cov))
round(post_sd, 4)

forward <- function(theta) theta %*% t(G)
prior <- matrix(rnorm(nreal * npar), nreal, npar,
                dimnames = list(NULL, paste0("p", seq_len(npar))))
```

A small helper runs the smoother and returns the realised posterior spread
alongside the collapse diagnostic recorded on the final iteration. The seed
is pinned so every configuration below is compared on identical perturbed
data.

```{r helper}
run_ies <- function(inflation = NULL, localisation = NULL) {
  fit <- pesto_ies_callback(
    forward_model  = forward,
    prior_ensemble = prior,
    obs            = setNames(y, paste0("o", seq_len(nobs))),
    obs_sd         = obs_sd,
    noptmax        = 12L,
    inflation      = inflation,
    localisation   = localisation,
    seed           = 42L,
    verbose        = FALSE
  )
  par_post <- as.matrix(fit$par_ensemble[, -1L])
  last_diag <- fit$iterations[[length(fit$iterations)]]
  list(
    sd_ratio  = mean(apply(par_post, 2L, sd) / post_sd),
    sd_by_par = apply(par_post, 2L, sd) / post_sd,
    ess_ratio = last_diag$spread_ess_ratio
  )
}
```

### The bare smoother under-disperses

```{r bare}
bare <- run_ies()
round(c(sd_ratio = bare$sd_ratio, spread_ess_ratio = bare$ess_ratio), 3)
```

The mean posterior standard deviation falls short of the analytic value:
with only `r nreal` realisations for `r npar` parameters the ensemble is
over-confident. The spread-ESS ratio quantifies the same collapse from the
eigenspectrum of the parameter anomaly covariance, where 1 means variance is
spread isotropically across every direction and a small value means it has
collapsed onto a few.

### Is this finite-ensemble collapse, or something else?

The question is worth asking of any under-dispersed ensemble, and it has a
cheap answer. If the shortfall is finite-ensemble collapse, the ratio of
realised to analytic spread climbs toward 1 as the ensemble grows, because a
larger sample gives the update a better estimate of the true covariance
structure. If it is flat in `nreal`, the cause is structural and no amount of
inflation will reach it. The sweep below runs the same problem at increasing
ensemble sizes, kept modest here for build time. Each point draws its own
prior from the same seed, so its 24-realisation entry is a second draw and
differs slightly from the headline run above.

```{r size-sweep}
sweep_nreal <- c(24L, 50L, 100L, 250L, 500L)
sd_by_nreal <- vapply(sweep_nreal, function(n) {
  set.seed(42L)
  prior_n <- matrix(rnorm(n * npar), n, npar,
                    dimnames = list(NULL, paste0("p", seq_len(npar))))
  fit_n <- pesto_ies_callback(
    forward_model = forward, prior_ensemble = prior_n,
    obs = setNames(y, paste0("o", seq_len(nobs))), obs_sd = obs_sd,
    noptmax = 12L, seed = 42L, verbose = FALSE
  )
  mean(apply(as.matrix(fit_n$par_ensemble[, -1L]), 2L, sd) / post_sd)
}, numeric(1L))
tab_sweep <- data.frame(nreal = sweep_nreal, sd_ratio = sd_by_nreal)
```

```{r sweep-table, echo = FALSE}
knitr::kable(
  tab_sweep, digits = 3,
  col.names = c("Realisations", "Realised SD / analytic SD"),
  caption = paste0(
    "Ensemble spread against the closed-form posterior, by ensemble size. ",
    "A value of 1 means the ensemble carries exactly the right spread."
  )
)
```

```{r sweep-figure, fig.height = 4.2, fig.cap = "Realised ensemble spread relative to the closed-form posterior spread, against ensemble size. The dashed line at 1 is the correct spread. The ratio climbs with the ensemble, which is the signature of finite-ensemble collapse; a curve flat in ensemble size would mean the cause lay elsewhere and inflation would only paper over it."}
ggplot2::ggplot(tab_sweep, ggplot2::aes(x = nreal, y = sd_ratio)) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dashed",
                      colour = "grey40") +
  ggplot2::geom_line(colour = "#0072B2", linewidth = 1.1) +
  ggplot2::geom_point(colour = "#0072B2", size = 2.6) +
  ggplot2::scale_x_log10() +
  ggplot2::expand_limits(y = c(0, 1.05)) +
  ggplot2::labs(
    x = "ensemble size (realisations, log scale)",
    y = "realised SD / analytic posterior SD (dimensionless)",
    title = "Does the shortfall shrink as the ensemble grows?"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

Run this sweep on your own problem before reaching for either
countermeasure. It is also the diagnostic that exposed the
unperturbed-observation defect described above: before PESTO perturbed the
observations, the same sweep sat flat at about 0.21 from 24 realisations to
8,000.

### Inflation re-expands the spread

```{r inflation}
rtps <- run_ies(inflation = pesto_inflation("rtps", alpha = 0.2))
adaptive <- run_ies(
  inflation = pesto_inflation("adaptive", retention_floor = 0.6)
)
```

### Localisation suppresses spurious correlations

```{r localisation}
loc <- run_ies(
  localisation = pesto_localisation("correlation", taper = "hard")
)
```

The two countermeasures compose: inflation restores variance magnitude,
localisation removes the spurious updates that drain it.

```{r both}
both <- run_ies(
  inflation    = pesto_inflation("rtps", alpha = 0.1),
  localisation = pesto_localisation("correlation", taper = "hard")
)
```

```{r countermeasure-table, echo = FALSE}
lst_runs <- list(none = bare, rtps = rtps, adaptive = adaptive,
                 correlation = loc, `rtps + correlation` = both)
tab_counter <- data.frame(
  Configuration = c("none (bare smoother)", "RTPS inflation, alpha = 0.2",
                    "adaptive inflation, floor = 0.6",
                    "correlation localisation",
                    "RTPS alpha = 0.1 plus localisation"),
  `Realised SD / analytic SD` = vapply(lst_runs, `[[`, numeric(1L),
                                       "sd_ratio"),
  `Spread-ESS ratio` = vapply(lst_runs, `[[`, numeric(1L), "ess_ratio"),
  check.names = FALSE, row.names = NULL
)
knitr::kable(
  tab_counter, digits = 3,
  caption = paste0(
    "Each countermeasure on the same problem and the same perturbed data, ",
    "at ", nreal, " realisations. The target for the first column is 1."
  )
)
```

```{r per-parameter-figure, fig.height = 4.5, fig.cap = "Realised spread relative to the closed-form posterior spread, parameter by parameter, under each countermeasure. The dashed line at 1 is the correct spread. The mean over parameters is what the table reports; this figure shows that the shortfall is not spread evenly, and that a countermeasure can lift one parameter past the target while leaving another short."}
tab_bypar <- data.table::rbindlist(lapply(names(lst_runs), function(nm) {
  data.table::data.table(
    configuration = nm,
    parameter = names(lst_runs[[nm]]$sd_by_par),
    ratio = as.numeric(lst_runs[[nm]]$sd_by_par)
  )
})) # ends the loop over countermeasure configurations
tab_bypar[, configuration := factor(configuration, levels = names(lst_runs))]

ggplot2::ggplot(
  tab_bypar,
  ggplot2::aes(x = parameter, y = ratio, colour = configuration,
               group = configuration)
) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dashed",
                      colour = "grey40") +
  ggplot2::geom_line(alpha = 0.7) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.9) +
  ggplot2::labs(
    x = "parameter", y = "realised SD / analytic posterior SD",
    title = "Where the spread is missing, parameter by parameter"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

### The collapse trajectory

`ensemble_spread_ess()` is recorded on every iteration regardless of method,
so the collapse trajectory is always available in the result, not only its
endpoint.

```{r ess-trace}
ess_trace <- function(inflation = NULL) {
  fit <- pesto_ies_callback(
    forward, prior, setNames(y, paste0("o", seq_len(nobs))),
    obs_sd = obs_sd, noptmax = 12L, inflation = inflation,
    seed = 42L, verbose = FALSE
  )
  vapply(fit$iterations, function(d) d$spread_ess_ratio, numeric(1L))
}
tr_bare <- ess_trace()
tr_rtps <- ess_trace(pesto_inflation("rtps", alpha = 0.2))
tab_trace <- data.table::data.table(
  iteration = rep(seq_along(tr_bare), 2L),
  ess_ratio = c(tr_bare, tr_rtps),
  run = rep(c("bare", "RTPS inflation"), each = length(tr_bare))
)
```

```{r trace-figure, fig.height = 4.2, fig.cap = "Spread-ESS ratio by smoother iteration, without and with RTPS inflation. Both decline as the update concentrates the ensemble on the directions the data inform; the inflated run sits above the bare one at every iteration. Read the gap between the lines, not the shape of either: what inflation buys here is a slower drain, not a flat line."}
ggplot2::ggplot(
  tab_trace,
  ggplot2::aes(x = iteration, y = ess_ratio, colour = run, shape = run)
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.7) +
  ggplot2::scale_shape_manual(name = NULL, values = c(16L, 17L)) +
  ggplot2::ylim(0, 1) +
  ggplot2::labs(
    x = "smoother iteration", y = "spread-ESS ratio (dimensionless)",
    title = "Dispersion trajectory, bare against RTPS inflation"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

## Read

The bare smoother under-disperses on this problem, and the numbers say by
how much: the realised ensemble spread is
`r round(100 * bare$sd_ratio)` per cent of the closed-form posterior spread
at `r nreal` realisations, with a spread-ESS ratio of
`r round(bare$ess_ratio, 3)`.

The sweep answers the first half of the question. The ratio climbs from
`r round(tab_sweep$sd_ratio[1], 3)` at `r sweep_nreal[1]` realisations to
`r round(tab_sweep$sd_ratio[nrow(tab_sweep)], 3)` at
`r sweep_nreal[length(sweep_nreal)]`, across a
`r round(max(sweep_nreal) / min(sweep_nreal))`-fold change in ensemble size.
That is finite-ensemble collapse, which is exactly what inflation and
localisation were designed for. Had the curve been flat, the conclusion to
draw would have been that the cause lay elsewhere and that inflating the
spread would have hidden it rather than fixed it.

The second half is the countermeasure table, and it answers with a
qualification. RTPS inflation at `alpha = 0.2` lifts the realised spread
from `r round(bare$sd_ratio, 3)` of the analytic value to
`r round(rtps$sd_ratio, 3)`, which is past the target rather than up to it;
the adaptive method reaches `r round(adaptive$sd_ratio, 3)`, correlation
localisation on its own `r round(loc$sd_ratio, 3)`, and RTPS at the
weaker `alpha = 0.1` composed with localisation
`r round(both$sd_ratio, 3)`. Every
configuration moves the spread in the right direction, and the strongest of
them moves it too far, which is the tuning problem in one line. The
per-parameter figure shows what the mean hides: the shortfall is uneven
across the six parameters, ranging from
`r round(min(bare$sd_by_par), 2)` to `r round(max(bare$sd_by_par), 2)` under
the bare smoother, so a single global inflation strength cannot bring every
direction to the target at once.

The trajectory figure separates two different questions. The spread-ESS
ratio falls from `r round(tr_bare[1], 3)` to
`r round(tr_bare[length(tr_bare)], 3)` over `r length(tr_bare)` iterations
in the bare run, and from `r round(tr_rtps[1], 3)` to
`r round(tr_rtps[length(tr_rtps)], 3)` in the inflated one; the inflated
line sits above the bare one at every iteration by an average of
`r round(mean(tr_rtps - tr_bare), 3)`. That gentle decline is the update
doing its job, concentrating the ensemble on the directions the data inform,
and the gap between the lines is the effect being demonstrated. The
spread-ESS ratio is a trajectory diagnostic; the ratio against the analytic
posterior is the calibration check, and the two answer different questions.

The caveat worth stating plainly is that inflation is a *tunable* remedy,
not a self-correcting one. The strengths used here, `alpha = 0.2` and
`retention_floor = 0.6`, were chosen against this problem's known posterior.
A stronger setting overshoots and returns an ensemble that is too wide,
which misleads a reader exactly as much as one that is too narrow. Measure
the spread ratio or the interval coverage on your own problem and tune to
what you measure; do not carry these numbers across.

## Limits

The whole vignette rests on a linear-Gaussian problem with a closed-form
posterior, which is what makes the calibration check exact and also what
makes it unrepresentative: a real forward model is nonlinear, and there the
ensemble's shortfall against the true posterior cannot be computed at all,
only estimated by interval coverage on held-out data. The sweep runs to 500
realisations, so it establishes the direction of the trend rather than the
size of the ensemble at which the shortfall would close. Inflation strengths
here are tuned against a known answer, which is precisely the information a
real calibration lacks, and no default value in this vignette should be
carried to another problem. Localisation is exercised only in its
correlation-based automatic form on parameters with no spatial coordinate;
the distance-based Gaspari-Cohn taper is named but not demonstrated, and a
spatial problem would need it. Neither countermeasure addresses a
mis-specified observation standard deviation, which is a different failure
with a different fix, worked through in *Wiring your own simulator into PESTO*.

## What to read next

*Wiring your own simulator into PESTO* covers the other route to a falsely confident
posterior, an observation standard deviation set from the standard error of
a mean rather than from the replicate spread, and shows why the spread-ESS
diagnostic used here cannot see it. *Can PESTO recover APSIM's own parameters?*
applies inflation to held-out predictive coverage on a simulator-backed
calibration. *Your first inversion: from a season of measurements to a
posterior* is the shortest introduction to the smoother itself.

## References

- Emerick, A. A., & Reynolds, A. C. (2013). Ensemble smoother with multiple
  data assimilation. *Computers & Geosciences*, 55, 3-15.
- Gaspari, G., & Cohn, S. E. (1999). Construction of correlation functions
  in two and three dimensions. *Quarterly Journal of the Royal
  Meteorological Society*, 125(554), 723-757.
- Luo, X., & Bhakta, T. (2020). Automatic and adaptive localization for
  ensemble-based history matching. *Journal of Petroleum Science and
  Engineering*, 184, 106559.
- Whitaker, J. S., & Hamill, T. M. (2012). Evaluating methods to account for
  system errors in ensemble data assimilation. *Monthly Weather Review*,
  140(9), 3078-3089.

## Reproduce

Seed `42` sets the design matrix, the truth, the observation noise and every
prior draw, and is passed to `pesto_ies_callback()` on every run so that all
configurations are compared on identical perturbed observations. Package
versions follow.

```{r session-info}
sessionInfo()
```
