---
title: "Countering Ensemble Collapse: Inflation and Localisation"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Countering Ensemble Collapse: Inflation and Localisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## The problem: ensemble under-dispersion

An iterative ensemble smoother estimates a posterior parameter
distribution from a finite ensemble of realisations. Two distinct
pathologies can make the posterior spread *too narrow* -- the ensemble
becomes over-confident:

1. **Finite-ensemble collapse.** Each assimilation step contracts the
   ensemble spread; over several iterations the posterior variance can
   fall below the true posterior variance. This is the classical,
   well-characterised effect the ensemble-Kalman literature addresses
   with inflation and localisation (Whitaker & Hamill 2012; Luo & Bhakta
   2020), and it shrinks as the ensemble grows.
2. **Spurious correlations.** A finite ensemble manufactures apparent
   correlations between parameters and observations that are not real.
   Acting on them injects noise into the update and accelerates
   collapse.

A third pathology used to sit alongside these two and was by far the
largest: assimilating the *same*, unperturbed observation vector into
every realisation. That gives every member an identical data pull, so the
between-member spread stops being a posterior spread at all -- and,
unlike cause 1, it does not shrink as the ensemble grows. PESTO's smoother
now perturbs the observations per realisation under an ES-MDA schedule
(see *Posterior spread and observation perturbation* in
[`?pesto_ies_callback`](../html/pesto_ies_callback.html)), so what remains
on this problem is the classical, ensemble-size-dependent collapse that
inflation and localisation were designed for. The sweep below is the
evidence for that: the sd-ratio now climbs toward 1 with `nreal` instead
of sitting flat.

PESTO addresses the first with **covariance inflation**
([`pesto_inflation()`](../html/pesto_inflation.html)) and the second with
**covariance localisation**
([`pesto_localisation()`](../html/pesto_localisation.html)). Both are
opt-in: the default `NULL` leaves the update identical to the bare
smoother.

This vignette demonstrates the effect on a linear-Gaussian problem, where
the analytic posterior is known and the collapse can be measured exactly.

## A linear-Gaussian problem with a known posterior

For a linear forward model $d = G\theta + \varepsilon$ with a Gaussian
prior $\theta \sim N(0, C_0)$ and observation error
$\varepsilon \sim N(0, R)$, the posterior covariance is
$C_{\mathrm{post}} = (C_0^{-1} + G^{\top} R^{-1} G)^{-1}$. We can therefore
compare the ensemble's posterior spread directly against the truth.

```{r problem}
library(PESTO)
set.seed(42L)

npar  <- 6L
nobs  <- 10L
nreal <- 24L          # a deliberately small ensemble, to provoke collapse

G          <- matrix(rnorm(nobs * npar), nobs, npar)
theta_true <- rnorm(npar)
obs_sd     <- 0.3
y          <- as.numeric(G %*% theta_true) + rnorm(nobs, sd = obs_sd)

# Analytic posterior standard deviation (standard-normal prior).
post_cov <- solve(diag(npar) + crossprod(G) / obs_sd^2)
post_sd  <- sqrt(diag(post_cov))

forward <- function(theta) theta %*% t(G)
prior   <- matrix(rnorm(nreal * npar), nreal, npar,
                  dimnames = list(NULL, paste0("p", seq_len(npar))))
```

A small helper runs the smoother and returns the realised posterior
spread alongside the spread-ESS collapse diagnostic recorded on the final
iteration.

```{r helper}
run_ies <- function(inflation = NULL, localisation = NULL) {
  fit <- pesto_ies_callback(
    forward_model  = forward,
    prior_ensemble = prior,
    obs            = setNames(y, paste0("o", seq_len(nobs))),
    obs_sd         = obs_sd,
    noptmax        = 12L,
    inflation      = inflation,
    localisation   = localisation,
    # Pin the ES-MDA observation-noise stream so every configuration below
    # is compared on identical perturbed data.
    seed           = 42L,
    verbose        = FALSE
  )
  par_post  <- as.matrix(fit$par_ensemble[, -1])
  last_diag <- fit$iterations[[length(fit$iterations)]]
  list(
    sd_ratio  = mean(apply(par_post, 2L, sd) / post_sd),
    ess_ratio = last_diag$spread_ess_ratio
  )
}
```

## The bare smoother under-disperses

```{r bare}
bare <- run_ies()
round(bare$sd_ratio, 3)
```

The mean posterior standard deviation falls short of the analytic value --
with only 24 realisations for 6 parameters the ensemble is over-confident.
The spread-ESS ratio
quantifies the same collapse from the eigenspectrum of the parameter
anomaly covariance (1 means variance is spread isotropically across all
directions; small values mean it has collapsed onto a few):

```{r bare-ess}
round(bare$ess_ratio, 3)
```

### Is this finite-ensemble collapse, or something else?

The question is worth asking of any under-dispersed ensemble, and it has a
cheap answer. If the shortfall is finite-ensemble collapse, the sd-ratio
climbs toward 1 as the ensemble grows -- a larger sample gives the update a
better estimate of the true covariance structure. If it is flat in `nreal`,
the cause is structural and no amount of inflation will reach it. The sweep
below runs the same problem at increasing `nreal` (kept modest here for
vignette build time; the pattern continues out to 8,000 realisations):

```{r size-sweep}
sweep_nreal <- c(24L, 100L, 500L)
sd_by_nreal <- vapply(sweep_nreal, function(n) {
  set.seed(42L)
  prior_n <- matrix(rnorm(n * npar), n, npar,
                    dimnames = list(NULL, paste0("p", seq_len(npar))))
  fit_n <- pesto_ies_callback(
    forward_model = forward, prior_ensemble = prior_n,
    obs = setNames(y, paste0("o", seq_len(nobs))), obs_sd = obs_sd,
    noptmax = 12L, verbose = FALSE
  )
  mean(apply(as.matrix(fit_n$par_ensemble[, -1]), 2L, sd) / post_sd)
}, numeric(1L))

data.frame(nreal = sweep_nreal, sd_ratio = round(sd_by_nreal, 3))
```

The sd-ratio rises toward 1 across a 20-fold change in ensemble size:
the residual shortfall at `nreal = 24` is the ensemble-size-dependent
collapse, which is what inflation and localisation address. Run this sweep
on your own problem before reaching for either -- a flat trend would say
the cause lies elsewhere and inflation would only be papering over it.
(This sweep is also the diagnostic that exposed the unperturbed-observation
defect: before PESTO perturbed the observations the same sweep sat flat at
about 0.21 from 24 realisations to 8,000.)

## Inflation re-expands the spread

`pesto_inflation()` offers four methods. The workhorse is relaxation to
prior spread (`"rtps"`, Whitaker & Hamill 2012): each parameter's
posterior anomalies are rescaled toward the pre-update spread, so the
directions that collapsed hardest are re-inflated most. The `"adaptive"`
method instead targets a global spread-retention floor.

```{r inflation}
rtps     <- run_ies(inflation = pesto_inflation("rtps", alpha = 0.2))
adaptive <- run_ies(inflation = pesto_inflation("adaptive",
                                                retention_floor = 0.6))

data.frame(
  method        = c("none", "rtps", "adaptive"),
  sd_ratio      = round(c(bare$sd_ratio, rtps$sd_ratio, adaptive$sd_ratio), 3),
  spread_ess    = round(c(bare$ess_ratio, rtps$ess_ratio, adaptive$ess_ratio), 3)
)
```

Both methods lift the retained posterior spread toward the analytic
value. The caveat worth stating plainly is the other one: inflation is a
*tunable* remedy, not a self-correcting one. The strengths used here
(`alpha = 0.2`, `retention_floor = 0.6`) were chosen against this
problem's known posterior; a stronger setting overshoots and returns an
ensemble that is too wide, which is no more honest than one that is too
narrow. Measure the sd-ratio or the interval coverage on your own problem
and tune to it -- do not carry these numbers across.

The spread-ESS diagnostic ([`ensemble_spread_ess()`](../html/ensemble_spread_ess.html))
is recorded on **every** iteration regardless of method, so the collapse
trajectory is always available in the result:

```{r ess-trace, fig.cap = "Spread-ESS ratio by IES iteration, without and with RTPS inflation. Both decline gently as the update concentrates the ensemble on the informed directions; the inflated run sits above the bare one at every iteration, which is what inflation buys here -- a slower drain, not a flat line.", fig.alt = "Two lines plotted against IES iteration number (x axis, 1 to 12) with the spread-ESS ratio on the y axis from 0 to 1. Both start near 0.5 to 0.6 and decline gradually to between 0.3 and 0.35 by iteration 12. The inflated line sits consistently above the bare line by roughly 0.04."}
ess_trace <- function(inflation = NULL) {
  fit <- pesto_ies_callback(
    forward, prior, setNames(y, paste0("o", seq_len(nobs))),
    obs_sd = obs_sd, noptmax = 12L, inflation = inflation,
    seed = 42L, verbose = FALSE
  )
  vapply(fit$iterations, function(d) d$spread_ess_ratio, numeric(1L))
}
tr_bare <- ess_trace()
tr_rtps <- ess_trace(pesto_inflation("rtps", alpha = 0.2))

plot(tr_bare, type = "b", pch = 19, col = "grey40", ylim = c(0, 1),
     xlab = "iteration", ylab = "spread-ESS ratio",
     main = "Dispersion trajectory, bare vs RTPS inflation")
lines(tr_rtps, type = "b", pch = 17, col = "#1B7837")
legend("topright", c("bare", "RTPS"), pch = c(19, 17),
       col = c("grey40", "#1B7837"), bty = "n")
```

Read the gap between the two lines, not the shape of either. The gentle
decline is the update doing its job -- concentrating the ensemble on the
directions the data inform -- and the inflated run holding above the bare
one at every iteration is the effect being demonstrated. The trace is a
trajectory diagnostic; the sd-ratio numbers above are the calibration
check, comparing the *final* ensemble against the analytic posterior.

## Localisation suppresses spurious correlations

For parameter-estimation problems whose parameters carry no spatial
coordinate, the recommended localiser is the correlation-based automatic
method (Luo & Bhakta 2020). It needs no metric: it estimates a noise
floor from the ensemble itself and damps sample correlations that fall
below it.

```{r localisation}
loc <- run_ies(localisation = pesto_localisation("correlation",
                                                 taper = "hard"))
round(loc$sd_ratio, 3)
```

The two countermeasures compose -- inflation restores variance magnitude,
localisation removes the spurious updates that drain it:

```{r both}
both <- run_ies(
  inflation    = pesto_inflation("rtps", alpha = 0.1),
  localisation = pesto_localisation("correlation", taper = "hard")
)
round(both$sd_ratio, 3)
```

When a genuine distance metric *does* exist, the classical Gaspari-Cohn
taper is available via `pesto_localisation("distance", ...)`, supplying
either a precomputed `distances` matrix or `par_coords` / `obs_coords`
together with a localisation `radius`.

## References

Gaspari, G. & Cohn, S. E. (1999). Construction of correlation functions in
two and three dimensions. *Quarterly Journal of the Royal Meteorological
Society*, 125(554), 723--757.

Luo, X. & Bhakta, T. (2020). Automatic and adaptive localization for
ensemble-based history matching. *Journal of Petroleum Science and
Engineering*, 184, 106559.

Whitaker, J. S. & Hamill, T. M. (2012). Evaluating methods to account for
system errors in ensemble data assimilation. *Monthly Weather Review*,
140(9), 3078--3089.

## Session information

```{r sessioninfo}
sessionInfo()
```
