---
title: "Is PESTO the same algorithm as the tools it descends from?"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Is PESTO the same algorithm as the tools it descends from?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64 (Darwin 26.5),
     R 4.5.2, one core. The cross-tool benchmark is read from a frozen
     artefact because it needs the external PEST and PEST++ binaries; the
     worked example at the end runs live and takes about a second. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(PESTO)

# Colourblind-safe palette throughout (Wong 2011).
PAL <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#F0E442",
         "#56B4E9", "#E69F00", "#000000")
```

## Why

I already have a PEST workflow that works, and switching an established
calibration to a new tool is a risk I need a reason to take. Before I move,
the question I want answered is: *is PESTO's smoother actually the same
algorithm as the one I am running now, does it get the same answers, and
what does going native in R cost or buy me?* This vignette answers it in
three parts: a lineage table saying which tool implements which estimator, a
frozen head-to-head benchmark of PESTO against classic PEST and PEST++ on
identical inverse problems, and a live worked example in which PESTO's
smoother is compared against an independent reference implementation of the
same update.

## What

Three tools stand in a direct line, and a fair comparison has to respect
that they do not all implement the same estimator.

**PEST** (Doherty 2015) is the original Fortran framework. Its estimator is
deterministic Gauss-Levenberg-Marquardt with Tikhonov and
singular-value-decomposition-assisted regularisation, and its uncertainty
quantification is a *linearised* Bayesian posterior, the PREDUNC path.

**PEST++** (White et al. 2020) is the open-source C++ re-implementation. It
adds `pestpp-ies`, an iterative ensemble smoother after Chen and Oliver
(2013), which classic PEST does not provide.

**PESTO** brings that smoother family natively into R through
[`pesto_ies_callback()`](../html/pesto_ies_callback.html), and adds
surrogate acceleration, multi-fidelity, inflation and localisation, and an
in-process forward-model callback.
[`pesto_ies()`](../html/pesto_ies.html) drives the external `pestpp-ies`
binary when one is wanted, and
[`pestpp_available()`](../html/pestpp_available.html) probes for it without
erroring. [`create_pest_scenario()`](../html/create_pest_scenario.html),
[`write_pst()`](../html/write_pst.html) and
[`read_pst()`](../html/read_pst.html) build and round-trip a PEST control
file from R, so a problem defined in R is readable by the wider PEST
toolchain. [`pesto_reference_ies()`](../html/pesto_reference_ies.html) is a
compact pure-R implementation of the same smoother update, shipped as the
comparison target the worked example uses.

| Algorithm class | classic PEST | PEST++ | PESTO |
|---|---|---|---|
| Deterministic GLM with linearised UQ | yes (`pest`) | yes (`pestpp-glm`) | no |
| Iterative ensemble smoother | **no** | yes (`pestpp-ies`) | yes (native) |

So PESTO's smoother is benchmarked against `pestpp-ies`, its algorithmic
twin, while classic PEST enters as the deterministic reference. Reading a
deterministic optimum against an ensemble posterior is legitimate on a
well-posed problem, where the linearisation holds, and instructive on a
non-linear one, where it does not. Both cases appear below.

Every PESTO algorithm runs with no external dependency. The optional
cross-checks against the upstream binaries are guarded.

```{r binary-probe}
pestpp_available("pestpp-ies")
pesto_version()$pestpp_version
```

```{r binary-note, echo = FALSE, results = "asis"}
if (isTRUE(pestpp_available("pestpp-ies"))) {
  cat("A `pestpp-ies` binary is installed on this machine. The benchmark",
      "below is still read from its frozen artefact rather than re-run,",
      "because it also needs classic PEST and twenty seeds per cell.\n")
} else {
  cat("No `pestpp-ies` binary is installed on this machine, so nothing",
      "here shells out to one: the benchmark is read from a frozen",
      "artefact and the worked example runs entirely in R.\n")
}
```

## Do

### The cross-tool benchmark

This is the head-to-head picture: a fixed-seed Monte Carlo study, twenty
seeds per cell, run by a standalone reproducibility harness that drives
classic PEST, PEST++ and PESTO through identical inverse problems and scores
them on the same footing. The numbers are frozen real outputs shipped with
the package and read here; they are not recomputed at build time, because
the external binaries are absent from package check farms.

```{r multitool-load}
mt_path <- system.file("extdata", "pestpp_cache",
                       "multitool_benchmark_summary.rds", package = "PESTO")
if (!nzchar(mt_path)) {
  mt_path <- file.path("..", "inst", "extdata", "pestpp_cache",
                       "multitool_benchmark_summary.rds")
}
mt <- readRDS(mt_path)
mt_sum <- data.table::as.data.table(mt$summary)
```

```{r versions-table, echo = FALSE}
knitr::kable(
  data.frame(Tool = names(mt$tool_versions),
             Version = unname(mt$tool_versions)),
  caption = sprintf("Benchmarked versions (run %s).", mt$generated_on)
)
```

```{r problems-table, echo = FALSE}
knitr::kable(
  mt$problems[, c("id", "regime", "n_params", "n_obs")],
  col.names = c("Problem", "Regime", "Parameters", "Observations"),
  caption = paste0(
    "The two benchmark problems, chosen to straddle the boundary between ",
    "the algorithm classes. Twenty seeds each."
  )
)
```

```{r multitool-table}
primary <- mt_sum[tool %in% c("pesto_callback", "pest_predunc",
                              "pestpp_ies")]
primary[, problem := ifelse(tier == "tier1",
                            "linear (well-posed)", "non-linear ODE")]
m_order <- c("PESTO (native IES)", "PEST (classic, GLM + PREDUNC)",
             "PEST++ (pestpp-ies)")
primary[, method := factor(method, levels = m_order)]
data.table::setorder(primary, tier, method)

tab_primary <- primary[, .(
  Problem          = problem,
  Method           = method,
  `RMSE (median)`  = round(rmse_med, 4),
  `CI90 coverage`  = round(ci90_cov, 3),
  `Wall-clock (s)` = round(wallclock_s, 2),
  `Forward evals`  = round(fwd_evals)
)]
```

```{r multitool-kable, echo = FALSE}
knitr::kable(
  tab_primary,
  caption = paste0(
    "Medians over 20 seeds. RMSE is to the known truth; the target for ",
    "CI90 coverage is 0.90; forward evaluations are per single inversion."
  )
)
```

```{r multitool-speedup}
wide <- data.table::dcast(primary, problem ~ tool, value.var = "wallclock_s")
tab_speed <- data.table::data.table(
  Problem                 = wide$problem,
  `PESTO vs classic PEST` = round(wide$pest_predunc / wide$pesto_callback),
  `PESTO vs PEST++`       = round(wide$pestpp_ies / wide$pesto_callback)
)
```

```{r speedup-kable, echo = FALSE}
knitr::kable(
  tab_speed,
  caption = paste0(
    "Wall-clock speed-up factor, median over 20 seeds: how many times ",
    "faster one inversion is under PESTO's in-process driver."
  )
)
```

```{r accuracy-figure, fig.height = 4.5, fig.cap = "Median posterior error to the known truth, with inter-quartile bars, on a logarithmic axis. On the linear problem the three methods sit together; on the non-linear ODE the two ensemble methods stay close while the linearised deterministic posterior degrades by an order of magnitude with a long upper whisker across seeds."}
ggplot2::ggplot(primary, ggplot2::aes(method, rmse_med, colour = method)) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = rmse_iqr_lo, ymax = rmse_iqr_hi), linewidth = 0.7
  ) +
  ggplot2::facet_wrap(~ problem) +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_manual(values = stats::setNames(
    c(PAL[1], PAL[2], PAL[3]), m_order
  )) +
  ggplot2::labs(
    x = NULL, y = "posterior RMSE to truth (log scale)",
    title = "Accuracy: median and inter-quartile range over 20 seeds"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "none",
                 axis.text.x = ggplot2::element_text(angle = 20, hjust = 1))
```

```{r speed-figure, fig.height = 4.2, fig.cap = "Median wall-clock seconds per inversion on a logarithmic axis. PESTO's in-process driver sits below one second on the linear problem and near one second on the ODE; the file-coupled binaries sit in the tens to hundreds of seconds. The gap is two to three orders of magnitude and is an exchange cost, not an algorithmic one."}
ggplot2::ggplot(primary, ggplot2::aes(method, wallclock_s, fill = method)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ problem) +
  ggplot2::scale_y_log10() +
  ggplot2::scale_fill_manual(values = stats::setNames(
    c(PAL[1], PAL[2], PAL[3]), m_order
  )) +
  ggplot2::labs(
    x = NULL, y = "wall-clock per inversion, s (log scale)",
    title = "Cost: in-process PESTO against file-coupled binaries"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "none",
                 axis.text.x = ggplot2::element_text(angle = 20, hjust = 1))
```

### Does PESTO drive the binary faithfully?

The benchmark above compares algorithms. A separate question is whether
PESTO, used as a *driver* for `pestpp-ies` through `pesto_ies()`, reproduces
what invoking the binary directly would give. The conformance arm of the
PEST, PEST++ and PESTO benchmark harness answered that on 2026-07-16, running
the same problem, the same prior ensemble and the same options twice, once
through the harness calling the binary itself and once through PESTO, and
comparing the posterior ensembles element by element: the two agreed to the
last bit, with a maximum absolute difference of zero. That record is cited
here, not recomputed: reproducing it needs the external `pestpp-ies` binary,
which no package build can assume.

### Where the cost model comes from

PESTO's performance rests on four levers, each with a cost model and a
measured outcome.

**Speed at matched accuracy.** The speed-up is
$S = t_{\text{ref}} / t_{\text{PESTO}}$, median wall-clock per inversion,
which the table above reports. It follows from a cost model rather than from
fewer solves. A smoother with $N$ realisations over $T$ iterations performs
about $N(T{+}1)$ forward solves; a file-coupled engine pays, per solve, the
model cost $c_m$ plus an exchange overhead $c_{io}$ for the template write,
the output parse and the process spawn, while the in-process callback pays
$c_m$ alone:
$$t_{\text{file}} \approx N(T{+}1)\,(c_m + c_{io}), \qquad
  t_{\text{PESTO}} \approx N(T{+}1)\,c_m + c_{\text{upd}} .$$
When the exchange dominates, which is the cheap-model regime of this
benchmark, the speed-up approaches $1 + c_{io} / c_m$: the same number of
solves at far lower per-solve overhead, with a C++ update kernel. The
accuracy column and *Can PESTO recover APSIM's own parameters?*, which
recovers a known truth and brackets an independent optimiser on real data,
are what say the accuracy is not paid away for it.

**Evaluation economy.** When the forward model itself is expensive, cost
shifts to the *number* of expensive solves. A Gaussian-process surrogate
(Rasmussen and Williams 2006) predicts the confident realisations and sends
only the uncertain ones to the full model,
$$\text{saving} = 1 - \frac{n_{\text{model}}}{n_{\text{total}}},$$
a realisation taking the full model only when its predictive standard
deviation exceeds a threshold. The saving is regime-dependent and the
current release computes it as a projection rather than realising it, which
*When can a surrogate stand in for the simulator?* sets out in full.

**Cost-accuracy trade.** With a cheap and an expensive fidelity correlated at
$\rho$, the control-variate combiner
[`mf_control_variate()`](../html/mf_control_variate.html) (Kennedy and
O'Hagan 2000; Glasserman 2003) attains
$\mathrm{Var}(\hat\mu_{\text{CV}}) =
(1-\rho^2)\,\mathrm{Var}(\hat\mu_{\text{H}})$,
so a well-correlated cheap model cuts the expensive-evaluation budget for a
target variance by about $(1-\rho^2)$.

**Iteration economy.** With
$\phi(\theta) = \sum_i w_i (d_i - g_i(\theta))^2$ the weighted objective, the
optional `phi_tol` stops once the relative reduction
$(\phi_{k-1}-\phi_k)/\phi_{k-1}$ falls below a tolerance, spending no
iterations on a plateau.

These are gains in wall-clock time and evaluation budget, not in the number
of base solves, and not in raw-interval calibration.

### A worked example that runs here

The benchmark above used frozen binary outputs. This example runs end to end
at build time with no external dependency, and compares PESTO's native
smoother against the shipped pure-R reference implementation of the same
update.

The problem is synthetic: a one-dimensional exponential-decay response with
eight unknown parameters and fifteen observations,

$$y_i \;=\; \sum_{k=1}^{8} k\,\theta_k \, e^{-i / 10}\;+\;\varepsilon_i,
   \qquad i = 1, \ldots, 15, \quad
   \varepsilon_i \sim \mathcal{N}(0, \sigma^2).$$

The response is linear in $\boldsymbol{\theta}$, so the posterior is
analytically tractable, and it is worth noticing what the design actually
identifies: every observation is the same weighted sum
$S = \sum_k k\,\theta_k$ scaled by $e^{-i/10}$, so the design matrix has
rank one. Fifteen observations therefore pin down one number, and no
individual parameter, which is what makes this a searching test of two
implementations rather than an easy one.

```{r scen-a-truth}
set.seed(20260425L)
n_par <- 8L
par_names_A <- sprintf("p%d", seq_len(n_par))
theta_true_A <- c(1.20, 0.85, 0.55, 0.40, 0.30, 0.22, 0.16, 0.10)

forward_A <- function(p) {
  vapply(seq_len(15L),
         function(ii) sum(seq_along(p) * p * exp(-ii / 10.0)), numeric(1L))
}

y_true_A <- forward_A(theta_true_A)
obs_noise_sd_A <- 0.02
y_obs_A <- y_true_A + rnorm(length(y_true_A), sd = obs_noise_sd_A)
weights_A <- rep(1 / obs_noise_sd_A, length(y_obs_A))
```

PESTO defines problems programmatically, with no pre-existing control file
needed, and round-trips them through the PEST control-file format, so a
problem built in R is readable by the wider PEST toolchain:

```{r scen-a-pst}
parameters_A <- data.table::data.table(
  parnme = par_names_A, partrans = "log", parchglim = "factor",
  parval1 = 0.5, parlbnd = 0.001, parubnd = 5.0, pargp = "pgrp"
)
observations_A <- data.table::data.table(
  obsnme = sprintf("obs_%02d", seq_len(15L)),
  obsval = y_obs_A, weight = weights_A, obgnme = "obs_g"
)
pst_A <- create_pest_scenario(
  parameters = parameters_A, observations = observations_A,
  model_command = "python3 forward.py",
  pestpp_options = list(ies_num_reals = 60L)
)

scratch_A <- file.path(tempdir(), "pesto_scenA")
dir.create(scratch_A, showWarnings = FALSE, recursive = TRUE)
write_pst(pst_A, file.path(scratch_A, "scenA.pst"))
pst_A_rt <- read_pst(file.path(scratch_A, "scenA.pst"))   # round-trip check
c(parameters_written = pst_A_rt$control_data$npar,
  observations_written = pst_A_rt$control_data$nobs)
```

The in-process callback driver is the recommended path for an R forward
model: it keeps the C++ ensemble kernel but drives the outer loop in R,
calling the forward model directly. The contract is the subject of *Wiring
your own simulator into PESTO*.

```{r scen-a-pesto}
set.seed(20260425L)
n_real_A <- 60L
prior_A <- matrix(rnorm(n_real_A * n_par, mean = 0.5, sd = 0.4),
                  nrow = n_real_A, ncol = n_par,
                  dimnames = list(NULL, par_names_A))
forward_ens_A <- function(theta) t(apply(theta, 1L, forward_A))

t0 <- proc.time()[["elapsed"]]
fit_A <- pesto_ies_callback(
  forward_model  = forward_ens_A,
  prior_ensemble = prior_A,
  obs            = setNames(y_obs_A, sprintf("obs_%02d", seq_len(15L))),
  obs_sd         = obs_noise_sd_A,
  noptmax        = 6L,
  seed           = 20260425L,
  verbose        = FALSE
)
runtime_pesto_A <- proc.time()[["elapsed"]] - t0

post_A <- as.matrix(fit_A$par_ensemble[, -1L])[, par_names_A]
post_mean_pesto <- colMeans(post_A)
post_q05_pesto <- apply(post_A, 2L, quantile, 0.05)
post_q95_pesto <- apply(post_A, 2L, quantile, 0.95)
rmse_pesto_A <- sqrt(mean((post_mean_pesto - theta_true_A)^2))
round(c(seconds = runtime_pesto_A, posterior_rmse = rmse_pesto_A), 4)
```

The package ships a compact pure-R reference produced by
`pesto_reference_ies()`, an independent implementation of the Chen and
Oliver update on the same problem. It is the comparison target every reader
sees by default.

```{r scen-a-reference}
cache_path <- system.file("extdata", "pestpp_cache",
                          "scenario_a_reference.rds", package = "PESTO")
if (!nzchar(cache_path)) {
  src_guess <- file.path("..", "inst", "extdata", "pestpp_cache",
                         "scenario_a_reference.rds")
  if (file.exists(src_guess)) cache_path <- normalizePath(src_guess)
}
stopifnot(nzchar(cache_path), file.exists(cache_path))
ies_cache <- readRDS(cache_path)
# Same problem instance: the cached reference must have seen these data.
max(abs(ies_cache$y_obs - y_obs_A)) < 1e-10

posterior_ref_A <- as.matrix(
  ies_cache$posterior_par[, ies_cache$par_names, with = FALSE]
)
post_mean_ref <- colMeans(posterior_ref_A)
post_q05_ref <- apply(posterior_ref_A, 2L, quantile, 0.05)
post_q95_ref <- apply(posterior_ref_A, 2L, quantile, 0.95)
rmse_ref_A <- sqrt(mean((post_mean_ref - theta_true_A)^2))
ref_label <- "pure-R reference (Chen & Oliver 2013)"
round(c(reference_posterior_rmse = rmse_ref_A), 4)
```

The quantity to compare is therefore $S$, the combination the data
identify, whose posterior has a closed form. Its posterior standard
deviation follows from the prior precision plus the summed squared design
weights over the observation variance.

```{r identified-combination}
s_of <- function(theta) sum(seq_along(theta) * theta)
kernel_A <- exp(-seq_len(15L) / 10.0)
prior_var_S <- 0.4^2 * sum(seq_len(n_par)^2)
post_sd_S <- sqrt(1 / (1 / prior_var_S +
                         sum(kernel_A^2) / obs_noise_sd_A^2))

s_truth <- s_of(theta_true_A)
s_pesto <- apply(post_A, 1L, s_of)
s_ref <- apply(posterior_ref_A, 1L, s_of)

round(c(truth = s_truth, pesto_mean = mean(s_pesto),
        reference_mean = mean(s_ref), analytic_post_sd = post_sd_S,
        pesto_sd = sd(s_pesto), reference_sd = sd(s_ref)), 4)
```

```{r combination-figure, fig.height = 4.2, fig.cap = "Posterior density of the one quantity this design identifies, the weighted sum S, under PESTO's smoother and under the reference implementation. The solid line is the true value and the shaded band is the closed-form posterior at plus or minus two analytic standard deviations. Both implementations locate S; PESTO's spread matches the analytic width closely, while the reference's is markedly wider."}
tab_s <- data.table::rbindlist(list(
  data.table::data.table(value = s_pesto, source = "PESTO"),
  data.table::data.table(value = s_ref, source = ref_label)
))
ggplot2::ggplot(tab_s, ggplot2::aes(x = value, fill = source)) +
  ggplot2::annotate(
    "rect", xmin = s_truth - 2 * post_sd_S, xmax = s_truth + 2 * post_sd_S,
    ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.6
  ) +
  ggplot2::geom_density(alpha = 0.55, colour = NA) +
  ggplot2::geom_vline(xintercept = s_truth, linewidth = 0.6) +
  ggplot2::scale_fill_manual(values = stats::setNames(
    c(PAL[1], PAL[2]), c("PESTO", ref_label)
  ), name = NULL) +
  ggplot2::labs(
    x = "identified combination S (model units)", y = "posterior density",
    title = "The one quantity fifteen observations identify"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

```{r scen-a-table}
tab_agreement <- data.table::data.table(
  Parameter  = par_names_A,
  Truth      = theta_true_A,
  `PESTO mean` = round(post_mean_pesto, 4),
  `PESTO 5%`   = round(post_q05_pesto, 4),
  `PESTO 95%`  = round(post_q95_pesto, 4),
  `Reference mean` = round(post_mean_ref, 4),
  `Reference 5%`   = round(post_q05_ref, 4),
  `Reference 95%`  = round(post_q95_ref, 4)
)
tab_agreement[, `Relative difference (%)` :=
                round(100 * abs(`PESTO mean` - `Reference mean`) /
                        pmax(abs(`Reference mean`), 1e-6), 2)]
```

```{r scen-a-kable, echo = FALSE}
knitr::kable(
  tab_agreement,
  caption = paste0(
    "Posterior summaries on the same problem instance: PESTO's native ",
    "smoother against the shipped pure-R reference implementation."
  )
)
```

```{r scen-a-figure, fig.height = 4.8, fig.cap = "Posterior 90 per cent credible intervals for the eight individual parameters, from PESTO and from the reference implementation, with the true generating values marked by crosses. Neither implementation recovers the individual parameters, and neither should: the design identifies only their weighted sum, so these intervals record where each smoother's prior draw and damping schedule left the seven unconstrained directions."}
tab_pp <- data.table::rbindlist(list(
  data.table::data.table(parameter = par_names_A, mean = post_mean_pesto,
                         q05 = post_q05_pesto, q95 = post_q95_pesto,
                         source = "PESTO"),
  data.table::data.table(parameter = par_names_A, mean = post_mean_ref,
                         q05 = post_q05_ref, q95 = post_q95_ref,
                         source = ref_label)
))
tab_truth <- data.table::data.table(parameter = par_names_A,
                                    truth = theta_true_A)

ggplot2::ggplot(tab_pp, ggplot2::aes(parameter, mean, colour = source)) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = q05, ymax = q95),
    position = ggplot2::position_dodge(width = 0.45),
    size = 0.5, linewidth = 0.9
  ) +
  ggplot2::geom_point(
    data = tab_truth, ggplot2::aes(parameter, truth), inherit.aes = FALSE,
    shape = 4L, size = 3.5, colour = PAL[8], stroke = 1.2
  ) +
  ggplot2::scale_colour_manual(values = stats::setNames(
    c(PAL[1], PAL[2]), c("PESTO", ref_label)
  )) +
  ggplot2::labs(
    x = "parameter", y = "posterior value (model units)", colour = NULL,
    title = "Posterior 90 per cent credible intervals",
    subtitle = "Crosses mark the true generating parameter values"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

```{r read-facts, include = FALSE}
lin <- primary[problem == "linear (well-posed)"]
ode <- primary[problem == "non-linear ODE"]
rmse_lin_lo <- min(lin$rmse_med)
rmse_lin_hi <- max(lin$rmse_med)
rmse_ode_ens <- ode[tool %in% c("pesto_callback", "pestpp_ies"), rmse_med]
rmse_ode_pest <- ode[tool == "pest_predunc", rmse_med]
ci90_pesto_lin <- lin[tool == "pesto_callback", ci90_cov]
speed_min <- min(unlist(tab_speed[, -1L]))
speed_max <- max(unlist(tab_speed[, -1L]))
evals_pesto <- lin[tool == "pesto_callback", fwd_evals]
evals_pest <- lin[tool == "pest_predunc", fwd_evals]
max_rel_diff <- max(tab_agreement$`Relative difference (%)`)
s_rel_diff <- 100 * abs(mean(s_pesto) - mean(s_ref)) / mean(s_ref)
s_err_pesto <- 100 * abs(mean(s_pesto) - s_truth) / s_truth
s_err_ref <- 100 * abs(mean(s_ref) - s_truth) / s_truth
s_width_pesto <- sd(s_pesto) / post_sd_S
s_width_ref <- sd(s_ref) / post_sd_S
resid_pesto <- sqrt(mean((forward_A(post_mean_pesto) - y_obs_A)^2))
resid_ref <- sqrt(mean((forward_A(post_mean_ref) - y_obs_A)^2))
```

## Read

The three methods separate exactly where the lineage table predicts.

**On the linear problem they agree.** All three recover the truth to a
median error between `r round(rmse_lin_lo, 4)` and `r round(rmse_lin_hi, 4)`,
a spread of `r round(100 * (rmse_lin_hi - rmse_lin_lo) / rmse_lin_lo)` per
cent between the best and the worst of them. PESTO's R-native smoother is
neither better nor worse than the reference implementations, which is what a
faithful re-implementation should show.

**On the non-linear problem the algorithm classes separate.** The two
ensemble methods stay accurate, at median errors of
`r round(min(rmse_ode_ens), 3)` and `r round(max(rmse_ode_ens), 3)`, while
classic PEST's linearised posterior degrades to `r round(rmse_ode_pest, 2)`,
a factor of `r round(rmse_ode_pest / max(rmse_ode_ens))` worse, with an
upper quartile of `r round(ode[tool == "pest_predunc", rmse_iqr_hi])` across
seeds. The linearisation is reliable where the forward map is near-linear
and fragile where it is not, and that is the whole content of the choice
between the two families.

**Speed is wall-clock, not fewer evaluations.** PESTO is between
`r speed_min` and `r speed_max` times faster in wall-clock across the four
comparisons in the speed-up table. The cause is not fewer forward solves:
on the linear problem PESTO takes `r round(evals_pesto)` forward evaluations
against classic PEST's `r round(evals_pest)`, so the gradient-based method
uses the fewest. PESTO evaluates the forward model in process and updates
the ensemble in C++, avoiding the per-evaluation file exchange and process
spawn the file-coupled binaries pay on every call. The conformance arm cited
above says the other half of the same story: when PESTO is asked to drive
the binary rather than replace it, the answer it returns is the binary's own
answer, bit for bit.

**The CI90 column is a pre-fix figure and should be read as one.** PESTO's
benchmarked raw credible intervals under-cover badly, at
`r round(ci90_pesto_lin, 3)` on the linear problem against a nominal 0.90.
The cause has since been identified and fixed: the smoother assimilated the
same unperturbed observation vector into every realisation, so the ensemble
was a cloud around the best fit rather than a posterior sample. The
development version perturbs the observations per realisation under an
ES-MDA schedule and recovers the analytic posterior covariance and nominal
interval coverage on a linear-Gaussian problem with a closed-form answer, as
*Posterior spread and observation perturbation* in
[`?pesto_ies_callback`](../html/pesto_ies_callback.html) sets out. Re-running
this benchmark needs the external PEST and PEST++ binaries, which are absent
here, so the table still carries the pre-fix calibration row; the accuracy,
cost and forward-evaluation columns are unaffected. Measure coverage on your
own problem in either case.

The worked example closes the loop with a live comparison, and it is the
most searching of the three because the design identifies only one number.
On the quantity the data can constrain, the weighted sum $S$, PESTO's
posterior mean sits `r round(s_err_pesto, 2)` per cent from the truth and
the reference's `r round(s_err_ref, 2)` per cent, and the two agree with
each other to `r round(s_rel_diff, 2)` per cent. PESTO's posterior spread on
$S$ is `r round(s_width_pesto, 2)` times the closed-form posterior standard
deviation, so it is close to correctly calibrated on the identified
direction; the reference implementation's is `r round(s_width_ref, 1)` times
that width, which is the visible difference between the two densities in the
figure.

The per-parameter numbers tell the other half of the story and should not be
read as disagreement about the answer. Root-mean-square error to the true
parameter vector is `r round(rmse_pesto_A, 3)` for PESTO and
`r round(rmse_ref_A, 3)` for the reference, both large beside parameter
values that run from `r max(theta_true_A)` down to `r min(theta_true_A)`,
and the inversion itself took `r round(runtime_pesto_A, 2)` seconds.
Individual posterior means differ by up to `r round(max_rel_diff)` per cent
between the two implementations, and neither recovers the individual truths,
because a rank-one design leaves seven of the eight directions
unconstrained by the data: the smoothers move those directions only as far
as their own
prior draws and damping schedules take them, and there is no fact of the
matter for them to agree about. Both implementations fit the observations,
at root-mean-square residuals of `r round(resid_pesto, 4)` and
`r round(resid_ref, 4)` against an observation noise of
`r obs_noise_sd_A`. What a faithful re-implementation
owes is agreement on what the data determine, and that is what the
identified-combination figure shows.

## Limits

The benchmark is two problems, both small and both synthetic, at twenty
seeds each; it establishes the direction of the differences between the
algorithm classes and not their size on a real simulator-backed calibration,
where the forward model rather than the exchange overhead dominates the
cost. The speed-up figures come from a cheap-model regime by construction:
where a single forward solve costs a minute, the exchange overhead is a
rounding error and the same table would show ratios near one. The
calibration column predates the observation-perturbation fix and cannot be
regenerated without the external binaries, so it records the earlier
behaviour rather than the current one. Inflation strength is problem
dependent, and two independent instruments, a joint-calibration coverage
test and this cross-tool benchmark, disagreed on the optimal relaxation
strength, which means there is no universal default: report coverage and
tune to it. The conformance result cited above covers one problem, one
option set and one binary version, so it establishes that PESTO drove
`pestpp-ies` faithfully in that configuration rather than in every
configuration. The worked example compares PESTO against a reference
implementation shipped in the same package, which is a weaker independence
than the external-binary comparison; the external comparison is the frozen
benchmark, and the two are meant to be read together.

## What to read next

*Wiring your own simulator into PESTO* is the contract the in-process driver
is built on, and the reason the wall-clock gap above exists at all. *Can
PESTO recover APSIM's own parameters?* is the same comparison against a real
simulator, including a cross-check against an independent optimiser on real
measured biomass. *My posterior looks too confident: is the ensemble
collapsing?* is where the calibration question raised by the CI90 column is
worked through against a closed-form posterior. *When can a surrogate stand
in for the simulator?* takes up the evaluation-economy lever named in the
cost model.

## References

- Chen, Y., & Oliver, D. S. (2013). Levenberg-Marquardt forms of the
  iterative ensemble smoother for efficient history matching and uncertainty
  quantification. *Computational Geosciences*, 17(4), 689-703.
- Doherty, J. (2015). *Calibration and Uncertainty Analysis for Complex
  Environmental Models*. Watermark Numerical Computing, Brisbane.
- Emerick, A. A., & Reynolds, A. C. (2013). Ensemble smoother with multiple
  data assimilation. *Computers & Geosciences*, 55, 3-15.
- Evensen, G. (2018). Analysis of iterative ensemble smoothers for solving
  inverse problems. *Computational Geosciences*, 22(3), 885-908.
- Glasserman, P. (2003). *Monte Carlo Methods in Financial Engineering*.
  Springer, New York.
- Kennedy, M. C., & O'Hagan, A. (2000). Predicting the output from a complex
  computer code when fast approximations are available. *Biometrika*, 87(1),
  1-13.
- Rasmussen, C. E., & Williams, C. K. I. (2006). *Gaussian Processes for
  Machine Learning*. MIT Press, Cambridge, MA.
- White, J. T., Hunt, R. J., Fienen, M. N., & Doherty, J. E. (2020).
  *Approaches to Highly Parameterized Inversion: PEST++ Version 5*. U.S.
  Geological Survey Techniques and Methods 7-C26.
- Wong, B. (2011). Points of view: color blindness. *Nature Methods*, 8(6),
  441.

## Reproduce

Seed `20260425` sets the worked example's observations and prior ensemble,
and is passed to `pesto_ies_callback()`, where it fixes the per-realisation
observation perturbation. The frozen cross-tool benchmark was run on
`r mt$generated_on` with PESTO `r mt$tool_versions[["PESTO"]]`, classic PEST
`r mt$tool_versions[["PEST (classic)"]]` and `pestpp-ies`
`r mt$tool_versions[["PEST++ (pestpp-ies)"]]`, twenty seeds per cell; the
conformance result cited in the *Do* section was run on 2026-07-16 against
`pestpp-ies` 5.2.16. Neither is recomputed here. Package versions for this
rendering follow.

```{r session-info}
sessionInfo()
```
