---
title: "When can a surrogate stand in for the simulator?"
output:
  rmarkdown::html_vignette:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{When can a surrogate stand in for the simulator?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~5 s under rmarkdown::render(); macOS arm64 (Darwin 26.5),
     R 4.5.2, one core. The nonlinearity sweep and the training-time scaling
     block dominate; no external binary or simulator is used. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
```

```{r library}
library(PESTO)
```

## Why

A single run of my crop model takes about a minute, and an iterative
ensemble smoother wants fifty of them per iteration for six or eight
iterations. That is most of a working day for one calibration. The question
I need answered before I build a surrogate into the loop is: *can a
Gaussian process trained on the ensemble stand in for some of those model
runs without damaging the posterior, and how would I know when it cannot?*
This vignette measures both halves on synthetic problems small enough to
run while you read, and shows what PESTO does when the answer is no.

## What

[`train_gp_surrogate()`](../html/train_gp_surrogate.html) fits a Gaussian
process with a radial-basis kernel from an ensemble of parameter vectors and
the model outputs already evaluated at them, returning the fitted length
scale, signal variance and log marginal likelihood.
[`predict_gp_surrogate()`](../html/predict_gp_surrogate.html) evaluates that
process at new parameter vectors and returns both a predicted mean and a
predictive standard deviation, which is the quantity the switching rule
reads.

[`surrogate_ensemble_update()`](../html/surrogate_ensemble_update.html)
performs the classification and the blended smoother update: train a
process from the current ensemble, predict outputs for every realisation,
classify each by predictive uncertainty into surrogate-safe and
needs-the-full-model, apply a mean-offset bias correction, and compute the
update on the blended ensemble.
[`pesto_surrogate_ies()`](../html/pesto_surrogate_ies.html) is the guarded
front door to the same computation: it checks the training regime first and
returns a typed [`pesto_abstention()`](../html/pesto_abstention.html)
instead of a low-value run when the regime is unfavourable.
[`check_surrogate_regime()`](../html/check_surrogate_regime.html) is that
check, callable on its own before an expensive ensemble is scheduled.

For ensembles too large for an exact process,
[`train_rff_surrogate()`](../html/train_rff_surrogate.html) and
[`predict_rff_surrogate()`](../html/predict_rff_surrogate.html) substitute
the random Fourier feature approximation of Rahimi and Recht (2007), which
trades exactness for a training cost that grows linearly rather than
cubically in the ensemble size.

### Regime of applicability

The process has a bounded operating envelope, worth stating before the
demonstration. As a soft floor, use `n_train >= 5 * n_params`. Below that
ratio the surrogate's predictions at points it has not seen degrade, while
its predictive standard deviation *at its own training points* stays at the
noise floor; the switching rule reads the second quantity, so it cannot see
the first. That is why the floor is enforced as a separate gate rather than
left to the switching statistic, and the sweep at the end of the *Do*
section measures the degradation it guards against.
The second axis is smoothness: smooth, near-linear responses are well
approximated by a radial-basis kernel and yield large savings, whereas rough
or near-discontinuous responses such as sharp wetting fronts, threshold
activations and regime switches defeat the kernel and savings collapse. The
third is ensemble size: larger ensembles improve the fit roughly as
$\mathcal{O}(n^{-1})$ in posterior variance, and fewer than about twenty
realisations rarely produce a usable surrogate. Outside this envelope, run
pure smoother iterations through
[`ensemble_solution()`](../html/ensemble_solution.html):
`surrogate_ensemble_update()` reports near-zero savings rather than
degrading the posterior, but the training cost is paid for no return.

## Do

### The test problem

A nonlinear forward model inspired by groundwater flow, with a linear
sensitivity and a nonlinear interaction term:

$$y_j = \eta_j + \alpha \sin(\eta_j)\, e^{-0.1\,\lvert \eta_j \rvert}
   + \varepsilon_j, \qquad \eta_j = \sum_k G_{jk}\theta_k .$$

Here $\eta_j$ is the linear predictor and $\alpha$ controls the
nonlinearity: $\alpha = 0$ is linear, $\alpha = 1$ is strongly nonlinear.
The parameters, design matrix and observations below are synthetic,
generated at a known truth so the recovery checks later in this vignette
have something to compare against.

```{r test-problem}
forward_model <- function(theta, G, alpha = 0.3) {
  linear <- as.numeric(G %*% theta)
  nonlinear <- alpha * sin(linear) * exp(-0.1 * abs(linear))
  linear + nonlinear
}

set.seed(42L)
n_par <- 20L
n_obs <- 50L
n_real <- 50L

theta_true <- rnorm(n_par)
G <- matrix(rnorm(n_obs * n_par, sd = 1 / sqrt(n_par)), n_obs, n_par)
y_obs <- forward_model(theta_true, G, alpha = 0.3) + rnorm(n_obs, sd = 0.1)
```

### The prior ensemble

```{r prior-ensemble}
par_ens <- matrix(rnorm(n_real * n_par, sd = 1.5), n_real, n_par)
obs_ens <- t(apply(par_ens, 1L, function(p) {
  forward_model(p, G, alpha = 0.3) + rnorm(n_obs, sd = 0.1)
}))
dim(par_ens)
dim(obs_ens)
```

### Training a process on it

```{r train-gp}
gp <- train_gp_surrogate(par_ens, obs_ens)
c(n_train = gp$n_train, n_par = gp$n_par, n_obs = gp$n_obs)
round(c(length_scale = gp$length_scale, signal_var = gp$signal_var,
        log_marginal_likelihood = gp$log_marginal_likelihood), 3)
```

```{r predict-gp}
pred_in <- predict_gp_surrogate(gp, par_ens)
rmse_in <- sqrt(mean((pred_in$mean - obs_ens)^2))
round(c(mean_uncertainty = mean(pred_in$uncertainty),
        rmse_in_sample = rmse_in), 6)
```

The in-sample error is near zero because the process is interpolating its
own training points. That is by construction and says nothing about whether
the surrogate could replace a model run, which is a question about points it
has never seen.

### The check that matters: the surrogate against the simulator

The block below draws forty fresh parameter vectors from the same prior,
runs the real forward model at each, and asks the trained process to predict
the same outputs. Nothing here is in the training set, so the comparison is
the one a modeller actually cares about.

```{r holdout}
n_hold <- 40L
par_hold <- matrix(rnorm(n_hold * n_par, sd = 1.5), n_hold, n_par)
obs_hold <- t(apply(par_hold, 1L, function(p) forward_model(p, G, 0.3)))
pred_hold <- predict_gp_surrogate(gp, par_hold)

rmse_out <- sqrt(mean((pred_hold$mean - obs_hold)^2))
sd_sim <- sd(as.numeric(obs_hold))
round(c(rmse_out_of_sample = rmse_out, sd_of_simulator_output = sd_sim,
        ratio = rmse_out / sd_sim), 3)
```

The same held-out points let the training floor be measured rather than
asserted. The block below retrains the process on a four-times larger
ensemble from the identical prior, taking the ratio of training points to
parameters from `r round(n_real / n_par, 1)` to
`r round(4 * n_real / n_par, 1)`, and scores it on the same forty held-out
vectors.

```{r holdout-ratio}
n_big <- 4L * n_real
par_big <- matrix(rnorm(n_big * n_par, sd = 1.5), n_big, n_par)
obs_big <- t(apply(par_big, 1L, function(p) {
  forward_model(p, G, alpha = 0.3) + rnorm(n_obs, sd = 0.1)
}))
gp_big <- train_gp_surrogate(par_big, obs_big)
pred_big <- predict_gp_surrogate(gp_big, par_hold)
rmse_out_big <- sqrt(mean((pred_big$mean - obs_hold)^2))

round(c(points_per_param_small = n_real / n_par,
        rmse_ratio_small = rmse_out / sd_sim,
        points_per_param_large = n_big / n_par,
        rmse_ratio_large = rmse_out_big / sd_sim), 3)
```

```{r holdout-figure, fig.height = 5, fig.cap = "Gaussian-process prediction against the simulator it emulates, at forty parameter vectors the process has never seen. Each point is one output of one held-out realisation, coloured by the process's own predictive standard deviation at that point. The diagonal is exact agreement. Points scatter around the diagonal rather than lying on it, which is the out-of-sample error the in-sample check above cannot show."}
tab_hold <- data.table::data.table(
  simulator = as.numeric(obs_hold),
  surrogate = as.numeric(pred_hold$mean),
  uncertainty = as.numeric(pred_hold$uncertainty)
)

ggplot2::ggplot(
  tab_hold, ggplot2::aes(x = simulator, y = surrogate, colour = uncertainty)
) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                       colour = "grey40") +
  ggplot2::geom_point(alpha = 0.6, size = 1.4) +
  ggplot2::scale_colour_viridis_c(name = "predictive SD") +
  ggplot2::labs(
    x = "simulator output (model units)",
    y = "surrogate prediction (model units)",
    title = "Held-out agreement between the surrogate and the simulator"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

### The blended smoother update

`surrogate_ensemble_update()` classifies the ensemble and computes the
update. What it does *not* do in this release is call a forward model for
the realisations it classifies as needing one: `obs_ensemble` is the
caller's already-evaluated ensemble, and the classification only chooses
which of those pre-computed rows to keep as they are and which to replace
with the surrogate prediction. `n_model_runs` and `savings_pct` therefore
report how many realisations *could* skip a forward-model call, not
evaluations avoided in this run, because the caller has already paid for
every one of them. Realising the saving requires wiring a `forward_model`
argument that this verb does not yet accept, so the numbers below are a
projection of the achievable saving rather than a report of compute already
saved. Two further points of precision: the mean-offset bias correction is,
as its name says, a single additive offset between the model-evaluated and
surrogate-evaluated rows, not the variance-reduction control-variate
estimator of Glasserman (2003) that
[`mf_control_variate()`](../html/mf_control_variate.html) implements
separately; and the prior-residual term in the blended update is held at
zero, so this path does not pull the surrogate-corrected realisations back
toward the prior the way the standard update does.

```{r surrogate-update}
weights <- rep(1 / 0.1, n_obs)
parcov_inv <- rep(1 / 1.5^2, n_par)

res_blend <- surrogate_ensemble_update(
  par_ensemble = par_ens,
  obs_ensemble = obs_ens,
  obs_target = y_obs,
  weights = weights,
  parcov_inv = parcov_inv,
  cur_lam = 1.0,
  uncertainty_threshold = 0.1
)
c(needs_full_model = res_blend$n_model_runs,
  surrogate_suffices = res_blend$n_surrogate_runs,
  total = res_blend$n_total)
round(c(projected_saving_pct = res_blend$savings_pct,
        mean_uncertainty = res_blend$mean_uncertainty,
        gp_length_scale = res_blend$gp_length_scale), 4)
```

Read the classification above beside the held-out check, not on its own.
Every realisation is classified surrogate-safe even though this
twenty-parameter problem with a fifty-member ensemble sits well *below* the
`n_train >= 5 * n_params` floor, because the process is being asked about
the very points it was trained on, where its predictive standard deviation
is at the noise floor. A total classification is therefore not evidence that
the surrogate could carry the run. The next sections put the same machinery
on a favourable problem, and then show what PESTO does when the regime is
unfavourable and the caller asks anyway.

### Does the blended update change the answer?

```{r compare}
par_mean <- colMeans(par_ens)
obs_mean <- colMeans(obs_ens)
par_diff <- t(par_ens) - par_mean
obs_diff <- t(obs_ens) - obs_mean
# Sign: ensemble_solution() expects sim - obs (see ?ensemble_solution).
obs_resid <- t(obs_ens) - matrix(rep(y_obs, n_real), n_obs, n_real)
par_resid <- par_diff
mat_am <- matrix(rnorm(n_par * (n_real - 1L)), n_par, n_real - 1L)

upgrade_std <- ensemble_solution(
  par_diff, obs_diff, obs_resid, par_resid,
  weights, parcov_inv, mat_am, cur_lam = 1.0
)
upgrade_surr <- res_blend$upgrade

upgrade_rmse <- sqrt(mean((upgrade_std - upgrade_surr)^2))
upgrade_corr <- cor(as.numeric(upgrade_std), as.numeric(upgrade_surr))
round(c(upgrade_rmse = upgrade_rmse, upgrade_correlation = upgrade_corr), 6)
```

### A favourable regime

The projected saving is the fraction of realisations classified as
surrogate-safe, $1 - n_{\text{model}} / n_{\text{total}}$. Because the
process is trained on the current ensemble and predicts at those same
points, it is confident enough to classify the whole ensemble as
surrogate-safe at the default threshold, so a high projected saving is in
large part a property of the construction. What decides whether wiring up
the real saving would be worth it is not how many realisations get
classified as safe but whether the surrogate's answer stays accurate, which
is what the held-out check above and the nonlinearity sweep below measure.
The problem below is smooth, linear and comfortably above the training
floor, so it is the case where a surrogate should do well.

```{r savings-number}
set.seed(7L)
np_f <- 4L
no_f <- 12L
nr_f <- 120L
G_f <- matrix(rnorm(no_f * np_f, sd = 1 / sqrt(np_f)), no_f, np_f)
fwd_f <- function(theta) as.numeric(G_f %*% theta)   # smooth, linear
th_f <- rnorm(np_f)
y_f <- fwd_f(th_f) + rnorm(no_f, sd = 0.05)
pe_f <- matrix(rnorm(nr_f * np_f, sd = 1.0), nr_f, np_f)
oe_f <- t(apply(pe_f, 1L, function(p) fwd_f(p) + rnorm(no_f, sd = 0.05)))

r_f <- surrogate_ensemble_update(
  pe_f, oe_f, y_f, rep(1 / 0.05, no_f), rep(1, np_f),
  uncertainty_threshold = 0.1
)
post_f <- pe_f + r_f$upgrade
rmse_f <- sqrt(mean((colMeans(post_f) - th_f)^2))
round(c(projected_saving_pct = r_f$savings_pct, posterior_rmse = rmse_f), 3)
```

The corollary matters as much: in the curse-of-dimensionality regime, with
parameters outnumbering the ensemble, the process still predicts the whole
ensemble but its answer degrades, giving high projected savings with a poor
posterior.

### When the regime is unfavourable

`pesto_surrogate_ies()` is the guarded entry point, and it declines rather
than proceeding into a low-value run. The call below asks for a surrogate
over twelve parameters with only thirty training points, a ratio of 2.5
against the floor of 5.

```{r abstention}
set.seed(20260902L)
par_off <- matrix(rnorm(30L * 12L), 30L, 12L)
obs_off <- matrix(rnorm(30L * 8L), 30L, 8L)

res_off <- pesto_surrogate_ies(
  par_ensemble = par_off,
  obs_ensemble = obs_off,
  obs_target   = rnorm(8L),
  weights      = rep(1, 8L),
  parcov_inv   = rep(1, 12L)
)
is_pesto_abstention(res_off)
res_off
```

The same gate is callable on its own, before an expensive ensemble is
scheduled. It returns a logical rather than a classed object, and warns on
the console naming the ratio it objected to; only the logical is shown here.

```{r regime-check}
isTRUE(suppressWarnings(
  check_surrogate_regime(n_params = 12L, n_train = 30L)
))
isTRUE(check_surrogate_regime(n_params = 12L, n_train = 120L))
```

### Effect of nonlinearity

How does surrogate accuracy vary with model nonlinearity? The sweep below
runs the standard update and the blended update at eleven values of
$\alpha$ and compares the recovered posterior mean against the truth in
each case.

```{r nonlinearity-sweep}
tab_alpha <- data.table::rbindlist(lapply(seq(0, 1, by = 0.1), function(a) {
  set.seed(42L)
  y_a <- forward_model(theta_true, G, a) + rnorm(n_obs, sd = 0.1)
  obs_a <- t(apply(par_ens, 1L, function(p) {
    forward_model(p, G, a) + rnorm(n_obs, sd = 0.1)
  }))

  pm <- colMeans(par_ens)
  om <- colMeans(obs_a)
  pd <- t(par_ens) - pm
  od <- t(obs_a) - om
  or_ <- t(obs_a) - matrix(rep(y_a, n_real), n_obs, n_real)
  am_a <- matrix(rnorm(n_par * (n_real - 1L)), n_par, n_real - 1L)

  upg_std <- ensemble_solution(pd, od, or_, pd, weights, parcov_inv,
                               am_a, 1.0)
  rmse_std <- sqrt(mean((colMeans(par_ens + upg_std) - theta_true)^2))

  surr <- surrogate_ensemble_update(par_ens, obs_a, y_a, weights,
                                    parcov_inv,
                                    uncertainty_threshold = 0.1)
  rmse_surr <- sqrt(mean((colMeans(par_ens + surr$upgrade) - theta_true)^2))

  data.table::data.table(alpha = a, rmse_std = rmse_std,
                         rmse_surr = rmse_surr,
                         ratio = rmse_surr / rmse_std,
                         savings = surr$savings_pct)
})) # ends the sweep over nonlinearity strengths
```

```{r nonlinearity-figure, fig.height = 4.2, fig.cap = "Posterior accuracy of the blended update relative to the standard update, across nonlinearity strengths. The vertical axis is the ratio of root-mean-square parameter error, so 1.0 means the two updates recover the truth equally well; the dashed line marks that parity."}
ggplot2::ggplot(tab_alpha, ggplot2::aes(x = alpha, y = ratio)) +
  ggplot2::geom_hline(yintercept = 1.0, linetype = "dashed",
                      colour = "grey40") +
  ggplot2::geom_line(colour = "#009E73", linewidth = 1.1) +
  ggplot2::geom_point(colour = "#009E73", size = 2.4) +
  ggplot2::labs(
    x = "nonlinearity strength (alpha, dimensionless)",
    y = "RMSE ratio, blended over standard (dimensionless)",
    title = "Surrogate accuracy across nonlinearity levels"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

```{r nonlinearity-table, echo = FALSE}
knitr::kable(
  tab_alpha[alpha %in% c(0, 0.3, 0.6, 1.0),
            .(alpha, rmse_std, rmse_surr, ratio, savings)],
  digits = 3,
  col.names = c("Alpha", "RMSE standard", "RMSE blended", "Ratio",
                "Projected saving (%)"),
  caption = paste0(
    "Parameter recovery under the standard and the blended update at four ",
    "nonlinearity strengths."
  )
)
```

### Random Fourier features for large ensembles

The exact process has $O(n^3)$ training cost. For large ensembles the random
Fourier feature approximation scales as $O(nD^2)$, where $D$ is the number
of random features:

```{r rff}
rff <- train_rff_surrogate(par_ens, obs_ens, n_features = 200L)
pred_rff <- predict_rff_surrogate(rff, par_ens)
rmse_rff <- sqrt(mean((pred_rff$mean - obs_ens)^2))
round(c(rff_train_mse = rff$train_mse, rff_rmse = rmse_rff,
        exact_gp_rmse = rmse_in), 6)
```

```{r scaling}
tab_scaling <- data.table::rbindlist(
  lapply(c(30L, 50L, 100L, 200L, 500L), function(nr) {
    x_n <- matrix(rnorm(nr * 20L), nr, 20L)
    y_n <- matrix(rnorm(nr * 50L), nr, 50L)
    t_gp <- system.time(train_gp_surrogate(x_n, y_n))[["elapsed"]] * 1000
    t_rff <- system.time(
      train_rff_surrogate(x_n, y_n, 200L)
    )[["elapsed"]] * 1000
    data.table::data.table(n = nr, method = c("exact GP", "RFF (D = 200)"),
                           time_ms = c(t_gp, t_rff))
  })
) # ends the sweep over ensemble sizes
```

```{r read-facts, include = FALSE}
rmse_in_txt <- format(signif(rmse_in, 2L), scientific = FALSE)
upgrade_rmse_txt <- format(signif(upgrade_rmse, 2L), scientific = FALSE)
```

```{r scaling-facts, include = FALSE}
t_gp_vec <- tab_scaling[method == "exact GP", time_ms]
t_rff_vec <- tab_scaling[method == "RFF (D = 200)", time_ms]
gp_growth <- max(t_gp_vec) / max(min(t_gp_vec), 0.01)
rff_growth <- max(t_rff_vec) / max(min(t_rff_vec), 0.01)
```

```{r scaling-figure, fig.height = 4.2, fig.cap = "Training time against ensemble size for the exact Gaussian process and the random Fourier feature approximation, on a logarithmic time axis. The exact process climbs steeply with the ensemble; the approximation stays close to flat over the same range."}
ggplot2::ggplot(
  tab_scaling,
  ggplot2::aes(x = n, y = pmax(time_ms, 0.01), colour = method)
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_viridis_d(name = NULL, end = 0.75) +
  ggplot2::labs(
    x = "ensemble size (realisations)",
    y = "training time (ms, log scale)",
    title = "Surrogate training time by ensemble size"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

## Read

The out-of-sample check is the one that answers the question. On forty
parameter vectors the process had never seen, its predictions sit
`r round(rmse_out, 2)` model units away from the simulator in
root-mean-square, against a simulator output spread of `r round(sd_sim, 2)`
units, a ratio of `r round(rmse_out / sd_sim, 2)`. The in-sample error over
the training ensemble is `r rmse_in_txt`, smaller by more than three orders
of magnitude, and the gap between those two numbers is the whole reason the
figure plots held-out points rather than training points. Read the colour
scale in that figure as the process's own warning system: it reports a
larger predictive standard deviation where its prediction is furthest from
the diagonal, which is what makes an uncertainty-driven switching rule
workable at all.

The training floor is now measured rather than asserted. At
`r round(n_real / n_par, 1)` training points per parameter the held-out
error is `r round(100 * rmse_out / sd_sim)` per cent of the simulator's own
spread; at `r round(n_big / n_par, 1)` points per parameter, on the same
prior and scored on the same forty held-out vectors, it falls to
`r round(100 * rmse_out_big / sd_sim)` per cent. That is the degradation the
`n_train >= 5 * n_params` gate exists to catch, and it is invisible to the
switching statistic, which is computed in sample.

The classification numbers make that last point concretely. On the
twenty-parameter problem, at `r round(n_real / n_par, 1)` points per
parameter and therefore well below the floor, the blended update classified
`r res_blend$n_surrogate_runs` of `r res_blend$n_total` realisations as
surrogate-safe: a projected saving of `r round(res_blend$savings_pct)` per
cent, with a mean predictive standard deviation of
`r round(res_blend$mean_uncertainty, 4)` against a switching threshold of
0.1. The resulting parameter upgrade is the standard update to within a
root-mean-square difference of `r upgrade_rmse_txt` and a correlation
of `r sprintf("%.6f", upgrade_corr)`, because a process interpolating its
own training points returns those points back, so substituting its
predictions for the evaluated rows changes almost nothing. A projected
saving of one hundred per cent and an upgrade identical to the unaccelerated
one are the same fact seen twice, and neither is evidence about a run in
which the surrogate would actually replace a model call.

On the smooth four-parameter problem, at `r round(nr_f / np_f)` training
points per parameter, the classification is equally total at
`r round(r_f$savings_pct)` per cent, and there the posterior it produces
recovers the truth to a root-mean-square error of `r round(rmse_f, 3)`. That
is the regime the machinery is for: above the floor, on a smooth response,
where the classification and the accuracy agree.

The nonlinearity sweep confirms the substitution's neutrality rather than
the surrogate's skill. Across $\alpha$ from 0 to 1 the ratio of blended to
standard parameter error stays between `r signif(min(tab_alpha$ratio), 5)`
and `r signif(max(tab_alpha$ratio), 5)`, so the blended path costs nothing
in accuracy on this problem as the model becomes nonlinear. The result to
carry away is not that a surrogate is free but that the substitution itself
is close to neutral, which is the precondition for the saving being worth
having once it is realised.

The refusal path is where PESTO stops rather than obliging. Asked for a
surrogate over twelve parameters with thirty training points, a ratio of
`r round(30 / 12, 1)` against the floor of 5, `pesto_surrogate_ies()`
returned a typed abstention with reason
`r paste0("'", res_off$reason, "'")` and no upgrade at all. A caller
branches on `is_pesto_abstention()` and falls back to the pure smoother; the
value of the typed form over a warning is that the decline is a value in the
return position, so it cannot be missed by a caller that never reads the
console. The same gate is available as `check_surrogate_regime()` before any
compute is committed.

The approximation for large ensembles trades what it says it trades. At
`r n_real` training points the random Fourier feature surrogate reaches an
in-sample root-mean-square error of `r round(rmse_rff, 4)` against the exact
process's `r rmse_in_txt`, a large relative loss on a quantity that is near
zero by construction. What it buys is in the scaling figure: over ensemble
sizes from 30 to 500 the exact process's training time grows by a factor of
`r round(gp_growth)`, while the approximation's grows by a factor of
`r round(rff_growth)`.

## Limits

Every problem in this vignette is synthetic, and two of them are linear or
nearly so, which is the regime a radial-basis kernel emulates best; nothing
here establishes how a Gaussian process behaves on a real crop or
groundwater simulator, whose response can carry thresholds, regime switches
and hard bounds that no smooth kernel represents. The projected saving is a
projection, as the section that reports it says at length: this release
classifies which realisations could skip a forward-model call and does not
skip one, so no number in this vignette is a measurement of compute avoided.
The nonlinearity sweep varies one scalar knob on one problem, which is not
the same as varying the *kind* of nonlinearity, and a threshold response
would be expected to behave quite differently from the smooth sinusoidal
term used here. The training floor of five points per parameter is an
empirical soft floor rather than a derived bound, and a smoother forward
model justifies a lower threshold, which is why `check_surrogate_regime()`
takes one as an argument. Finally, the held-out comparison uses a noiseless
simulator evaluation as its reference, so it isolates emulation error and
says nothing about how emulation error and observation noise interact inside
a full calibration.

## What to read next

*Your first inversion: from a season of measurements to a posterior* is the
shortest introduction to the smoother the surrogate accelerates.
*Wiring your own simulator into PESTO* covers the multi-fidelity alternative, in
which a cheap version of the real model rather than a statistical emulator
carries the early iterations, and the control-variate primitive that lifts
cheap outputs toward expensive ones. *Handing an ensemble to somebody else*
shows how a run that abstained still produces a typed record a downstream
tool can read. *Is PESTO the same algorithm as the tools it descends from?* sets the cost of
the whole approach against the tools PESTO descends from.

## References

- Glasserman, P. (2003). *Monte Carlo Methods in Financial Engineering*.
  Springer, New York.
- Rahimi, A., & Recht, B. (2007). Random features for large-scale kernel
  machines. *Advances in Neural Information Processing Systems*, 20.
- Rasmussen, C. E., & Williams, C. K. I. (2006). *Gaussian Processes for
  Machine Learning*. MIT Press, Cambridge, MA.
- Chen, Y., & Oliver, D. S. (2013). Levenberg-Marquardt forms of the
  iterative ensemble smoother. *Computational Geosciences*, 17(4), 689-703.

## Reproduce

Seed `42` sets the test problem, the prior ensemble and the nonlinearity
sweep; seed `7` sets the smooth four-parameter problem; seed `20260902` sets
the off-design ensemble that triggers the abstention. Package versions
follow.

```{r session-info}
sessionInfo()
```
